---
title: "Luminex: Total IgA/IgG/IgM Assay Data Analysis"
output: 
  html_document:
    toc: true            # Enable Table of Contents
    toc_float: true      # Makes the Table of Contents float (optional)
date: "22-10-2024"
---




# Function to install/load packages:
```{r load_packages function, include=FALSE}
load_packages <- function(packages) {
  sapply(packages, function(pkg) {
    if (!require(pkg, character.only = TRUE)) {
      install.packages(pkg, dependencies = TRUE)
      library(pkg, character.only = TRUE)
    }
  })
}

packages <- c("purrr", "dplyr", "tidyr", "knitr", "magrittr", "reshape2", 
              "tidyverse", "ggplot2", "RColorBrewer", "stringr",
              "rstatix", "ggpubr", "ggsignif", "FSA",  
              "readxl", "viridis", "patchwork",  "readr",
              "ggfortify", "factoextra", "pheatmap",
              "cowplot","pheatmap","psych","cowplot")

load_packages(packages)

getwd()

```

# Import total antibody data:
```{r}
mydf <- read_excel("01-Luminex_Total_Abs_Data.xlsx", 
    col_types = c("text", "text", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "text", "text", "text", 
        "text", "text", "text", "text", "text", 
        "text", "numeric", "numeric", "numeric", 
        "text", "numeric"))
```

# Handling of out-of-range (OOR) values 
```{r}
# Summarize out-of-range (OOR) Luminex values by target and sampling method
# Input: mydf with columns Analyte, Notes, Sampling_Method
# Output: gt summary table with OOR high/low counts and percentages

library(dplyr)
library(tidyr)
library(gt)

# 1) start from your existing df
df_long <- mydf %>%
  # 1a) tag OORs
  mutate(target=Analyte,
    limits = case_when(
      str_detect(Notes, "OOR <") ~ "OOR <",
      str_detect(Notes, "OOR >") ~ "OOR >",
      TRUE                       ~ NA_character_
    ),
    
  )

# Summarize OOR values by target and Sampling_Method using the long-format df
target_summary <- df_long %>%
  group_by(target, Sampling_Method) %>%
  summarise(
    OOR_High = sum(limits == "OOR >", na.rm = TRUE),
    OOR_Low  = sum(limits == "OOR <", na.rm = TRUE),
    total_samples = n(),
    .groups = "drop"
  ) %>%
  mutate(
    OOR_High_Percent = (OOR_High / total_samples) * 100,
    OOR_Low_Percent  = (OOR_Low / total_samples) * 100,
    High_Format = paste0(OOR_High, " (", round(OOR_High_Percent, 1), "%)"),
    Low_Format  = paste0(OOR_Low, " (", round(OOR_Low_Percent, 1), "%)")
  )

# Reshape to wider format with Sampling_Method as column groups
target_table <- target_summary %>%
  select(target, Sampling_Method, High_Format, Low_Format) %>%
  pivot_wider(names_from = Sampling_Method, values_from = c(High_Format, Low_Format)) %>%
  rename_with(~ gsub("High_Format_", "High_", .x), starts_with("High_Format")) %>%
  rename_with(~ gsub("Low_Format_", "Low_", .x), starts_with("Low_Format"))

# Create the gt table. 
gt_table <- target_table %>%
  gt(rowname_col = "target") %>%
  tab_header(
    title = md("Out of Range (OOR) Values (Count and Percentage)")
  ) %>%
  tab_spanner(
    label = md("**MTS**"),  
    columns = c(High_MTS_Professional, Low_MTS_Professional, High_MTS_Self, Low_MTS_Self)
  ) %>%
  tab_spanner(
    label = md("**NPS**"),  
    columns = c(High_NPS_Professional, Low_NPS_Professional)
  ) %>%
  tab_spanner(
    label = md("**MLF**"),  
    columns = c(High_MLF_Professional, Low_MLF_Professional, High_MLF_Self, Low_MLF_Self)
  ) %>%
  tab_spanner(
    label = md("**OPS**"),  
    columns = c(High_OPS_Professional, Low_OPS_Professional)
  ) %>%
  tab_spanner(
    label = md("**Saliva**"),  
    columns = c(High_Saliva_Professional, Low_Saliva_Professional, High_Saliva_Self, Low_Saliva_Self)
  ) %>%
  cols_label(
    High_MTS_Self = "Self (OOR >)",     Low_MTS_Self = "Self (OOR <)",
    High_MTS_Professional = "Professional (OOR >)", Low_MTS_Professional = "Professional (OOR <)",
    High_NPS_Professional = "Professional (OOR >)",   Low_NPS_Professional = "Professional (OOR <)",
    High_MLF_Self = "Self (OOR >)",      Low_MLF_Self = "Self (OOR <)",
    High_MLF_Professional = "Professional (OOR >)",    Low_MLF_Professional = "Professional (OOR <)",
    High_OPS_Professional = "Professional (OOR >)",    Low_OPS_Professional = "Professional (OOR <)",
    High_Saliva_Professional = "Professional (OOR >)", Low_Saliva_Professional = "Professional (OOR <)",
    High_Saliva_Self = "Self (OOR >)",   Low_Saliva_Self = "Self (OOR <)"
  ) %>%
  tab_options(
    table.font.size = 12,
    heading.title.font.size = 14
  )

# Print the table
gt_table

# Optionally, save the table:
# gtsave(gt_table, "Luminex_Total_Abs_Samples_OOR.html")
#gtsave(gt_table, "Luminex_Total_Abs_Samples_OOR.docx")
```


# Make subset data for each isotype:
```{r Total Abs, include=FALSE}
#----- Subsets--------
df_A <- mydf %>% 
  filter(Analyte == "IgA")

df_G <- mydf %>% 
  filter(Analyte == "IgG")

df_M <- mydf %>% 
  filter(Analyte == "IgM")

```

# Explore concentrations (before OOR substitutions)
```{r, Main Concentration Plot, warning=FALSE, fig.width=11, fig.height=8, echo=FALSE}

#Subset samples that were collected by professionals:
prof_df <- mydf %>% 
  filter(Sampling_Strategy == "Professional")

# Count the number of samples that were OOR (above or below limits of detection) for each isotype
count_a <- prof_df %>% filter(Analyte == "IgA") %>%  # IgA
  filter(Notes == "OOR >" | Notes == "OOR <")

count_g <- prof_df %>% filter(Analyte == "IgG") %>%  # IgG
  filter(Notes == "OOR >" | Notes == "OOR <")

count_m <- prof_df %>% filter(Analyte == "IgM") %>%  # IgM
  filter(Notes == "OOR >" | Notes == "OOR <")

#head(count_a)

##-----Professional----
###-------IgA----------
# Count for "OOR >" (above upper limits of quantitation)
count_OOR_greater_A <- count_a %>%
  filter(Notes == "OOR >") %>%
  group_by(Sample_Type) %>%
  summarise(Count_OOR_greater = n(), .groups = 'drop')

# Count for "OOR <" (below lower limits of detection)
count_OOR_less_A <- count_a %>%
  filter(Notes == "OOR <") %>%
  group_by(Sample_Type) %>%
  summarise(Count_OOR_less = n(), .groups = 'drop')

# Merge the two counts into one dataframe for total IgA
counts_A <- full_join(count_OOR_greater_A, count_OOR_less_A, by = "Sample_Type")

###-------IgG----------
# Count for "OOR >" (above upper limits of quantitation)
count_OOR_greater_G <- count_g %>%
  filter(Notes == "OOR >") %>%
  group_by(Sample_Type) %>%
  summarise(Count_OOR_greater = n(), .groups = 'drop')

# Count for "OOR <" (below lower limits of detection)
count_OOR_less_G <- count_g %>%
  filter(Notes == "OOR <") %>%
  group_by(Sample_Type) %>%
  summarise(Count_OOR_less = n(), .groups = 'drop')

# Merge the two counts into one dataframe for total IgG
counts_G <- full_join(count_OOR_greater_G, count_OOR_less_G, by = "Sample_Type")

###-------IgM----------
# Count for "OOR >" (above upper limits of quantitation)
count_OOR_greater_M <- count_m %>%
  filter(Notes == "OOR >") %>%
  group_by(Sample_Type) %>%
  summarise(Count_OOR_greater = n(), .groups = 'drop')

# Count for "OOR <" (below lower limits of detection)
count_OOR_less_M <- count_m %>%
  filter(Notes == "OOR <") %>%
  group_by(Sample_Type) %>%
  summarise(Count_OOR_less = n(), .groups = 'drop')

# Merge the two counts into one dataframe for total IgM
counts_M <- full_join(count_OOR_greater_M, count_OOR_less_M, by = "Sample_Type")


#-------Self-collected samples-----------------
#Subset samples that were self-collected by a subset of participants:
#
self_df <- mydf %>% 
  filter(Sampling_Strategy == "Self")

# Count the number of samples that were OOR (above or below limits of detection) for each isotype
s_count_a <- self_df %>% filter(Analyte == "IgA") %>%  # IgA
  filter(Notes == "OOR >" | Notes == "OOR <")

s_count_g <- self_df %>% filter(Analyte == "IgG") %>%  # IgG
  filter(Notes == "OOR >" | Notes == "OOR <")

s_count_m <- self_df %>% filter(Analyte == "IgM") %>%  # IgM
  filter(Notes == "OOR >" | Notes == "OOR <")

#head(count_a)

##-----Self-collected OORs----
###-------IgA----------
# Count for "OOR >" (above upper limits of quantitation)
s_count_OOR_greater_A <- s_count_a %>%
  filter(Notes == "OOR >") %>%
  group_by(Sample_Type) %>%
  summarise(Count_OOR_greater = n(), .groups = 'drop')

# Count for "OOR <" (below lower limits of detection)
s_count_OOR_less_A <- s_count_a %>%
  filter(Notes == "OOR <") %>%
  group_by(Sample_Type) %>%
  summarise(Count_OOR_less = n(), .groups = 'drop')

# Merge the two counts into one dataframe for total IgA
s_counts_A <- full_join(s_count_OOR_greater_A, s_count_OOR_less_A, by = "Sample_Type")

###-------IgG----------
# Count for "OOR >" (above upper limits of quantitation)
s_count_OOR_greater_G <- s_count_g %>%
  filter(Notes == "OOR >") %>%
  group_by(Sample_Type) %>%
  summarise(Count_OOR_greater = n(), .groups = 'drop')

# Count for "OOR <" (below lower limits of detection)
s_count_OOR_less_G <- s_count_g %>%
  filter(Notes == "OOR <") %>%
  group_by(Sample_Type) %>%
  summarise(Count_OOR_less = n(), .groups = 'drop')

# Merge the two counts into one dataframe for total IgG
s_counts_G <- full_join(s_count_OOR_greater_G, s_count_OOR_less_G, by = "Sample_Type")

###-------IgM----------
# Count for "OOR >" (above upper limits of quantitation)
s_count_OOR_greater_M <- s_count_m %>%
  filter(Notes == "OOR >") %>%
  group_by(Sample_Type) %>%
  summarise(Count_OOR_greater = n(), .groups = 'drop')

# Count for "OOR <" (below lower limits of detection)
s_count_OOR_less_M <- s_count_m %>%
  filter(Notes == "OOR <") %>%
  group_by(Sample_Type) %>%
  summarise(Count_OOR_less = n(), .groups = 'drop')

# Merge the two counts into one dataframe for total IgM
s_count_m <- full_join(s_count_OOR_greater_M, s_count_OOR_less_M, by = "Sample_Type")


#------------Here, we don't substitute OOR sample values so we will annotate these in the following plot----
n_distinct(prof_df$ID) #44 but some were OOR
n_distinct(self_df$ID) #16 but some were OOR

# Define custom colors for our Sample Types
custom_colors <- c("NPS" = "#A2D2DF", 
                   "MTS" = "#F6EFBD",       
                   "MLF" = "#FFBD73",
                   "OPS" = "#BC7C7C",
                   "Saliva" = "#8ABFA3")

main_plot_2 <- ggplot(mydf, aes(x = Sampling_Method, y = log10.conc,fill = Sample_Type)) +
  geom_boxplot(aes(group = Sampling_Method), alpha = 0.5) +  # Boxplot for each analyte
  geom_point(alpha = 0.4) +  # Points with slight dodge
  facet_wrap(~Analyte, ncol = 3) +
  scale_fill_manual(values = custom_colors)+
  labs(x = "Sampling Method", y = "Concentration (log10)",
       title = "Concentration of Analytes per Sample Type",
       subtitle="Professionally Collected (N = 44)",
       fill="Sample Type") +
  theme_bw() + theme(axis.text.x = element_text(angle=45, hjust = 1))

#main_plot_2

# We annotate per sample type the OOR samples accordingly for total IgA:
annotation_data <- data.frame(
  Sampling_Method = c("MLF_Professional","MLF_Self", "MTS_Professional",
                  "OPS_Professional", "Saliva_Professional","Saliva_Self"),
  Sample_Type = c("MLF","MLF","MTS",
                  "OPS","Saliva","Saliva"),
  log10.conc = c(3.7, 4.5, 3.1, 1.9, 1, 4.5),
  label = c("OOR > (N = 11)", "OOR > (N = 5)","OOR > (N = 1)\nOOR < (N = 3)", "OOR < (N = 11)",
             "OOR > (N = 7)\nOOR < (N = 1)","OOR > (N = 1)"),
  Analyte = c("IgA", "IgA", "IgA", "IgA", "IgA","IgA")
)

main_plot_n2 <- main_plot_2 +
  geom_text(data = annotation_data, aes(x = Sampling_Method, y = log10.conc, label = label),
            size = 2.2, hjust = 0.53, vjust = 0.75)
#main_plot_n2

# We annotate per sample type the OOR samples accordingly for total IgG:
annotation_data_g <- data.frame(
  Sampling_Method = c("MLF_Professional", "OPS_Professional"),
  log10.conc = c(1.2, 0.1),
  Sample_Type=c("MLF","OPS"),
  label = c("OOR < (N = 1)", "OOR < (N = 19)"),
  Analyte = c("IgG", "IgG")
)

main_plot3 <- main_plot_n2 +
  geom_text(data = annotation_data_g, aes(x = Sampling_Method, y = log10.conc, label = label),
            size = 2.2, hjust = 0.4, vjust = -1)
# main_plot3

# We annotate per sample type the OOR samples accordingly for total IgM:
annotation_data_m <- data.frame(
  Sampling_Method = c("MLF_Professional","MTS_Professional" ,"OPS_Professional","Saliva_Professional"),
  log10.conc = c(1.001, 1.02, 0.8,1.6),
  Sample_Type=c("MLF","MTS","OPS","Saliva"),
  label = c("OOR > (N = 1)\nOOR < (N = 1)", "OOR < (N = 1)", "OOR < (N = 8)","OOR < (N = 1)"),
  Analyte = c("IgM","IgM" ,"IgM","IgM")
)

main_plot4 <- main_plot3 +
  geom_text(data = annotation_data_m, aes(x = Sampling_Method, y = log10.conc, label = label),
            size = 2.2, hjust = 0.4, vjust = -1)

# This is the final boxplot, showing sample types' total antibody concentrations for each isotype and the number of corresponding OOR samples accordingly: 
main_plot4

# Optionally, save this plot:
 # ggsave(filename = "Main_Concentrations_Boxplot_29_10_2024.tiff"
 #        , plot = main_plot4, width = 12, height = 6 ,dpi = 300,
 #        device="tiff")
```


# Visualize concentrations (before OOR substitutions) only for professionally collected samples
```{r}
# Define custom colors for your Sample Types
custom_colors <- c("NPS" = "#A2D2DF", 
                   "MTS" = "#F6EFBD",       
                   "MLF" = "#FFBD73",
                   "OPS" = "#BC7C7C",
                   "Saliva" = "#8ABFA3")

my_df_prof <- mydf %>% filter(Sampling_Strategy == "Professional")

main_plot_00 <- ggplot(my_df_prof, aes(x = Sample_Type, y = log10.conc,fill = Sample_Type)) +
  geom_boxplot(aes(group = Sample_Type), alpha = 0.5) +  # Boxplot for each analyte
  geom_point(alpha = 0.4) +  # Points with slight dodge
  facet_wrap(~Analyte, ncol = 3) +
  scale_fill_manual(values = custom_colors)+
  labs(x = "Sampling Type", y = "Concentration (log10)",
       title = "Concentration of Analytes per Sample Type",
       subtitle="Professionally Collected (N = 44)",
       fill="Sample Type") +
  theme_bw() + theme(axis.text.x = element_text(angle=45, hjust = 1))
main_plot_00

# We annotate per sample type the OOR samples accordingly for total IgA:
annotation_data <- data.frame(
  Sample_Type = c("MLF","MTS",
                  "OPS","Saliva"),
  log10.conc = c(3.7, 3.1, 1.9, 1),
  label = c("OOR > (N = 11)","OOR > (N = 1)\nOOR < (N = 3)", "OOR < (N = 11)",
             "OOR > (N = 7)\nOOR < (N = 1)"),
  Analyte = c("IgA",  "IgA", "IgA", "IgA")
)

main_plot_01 <- main_plot_00 +
  geom_text(data = annotation_data, aes(x = Sample_Type, y = log10.conc, label = label),
            size = 2.2, hjust = 0.53, vjust = 0.75)

# We annotate per sample type the OOR samples accordingly for total IgG:
annotation_data_g <- data.frame(
  log10.conc = c(1.2, 0.1),
  Sample_Type=c("MLF","OPS"),
  label = c("OOR < (N = 1)", "OOR < (N = 19)"),
  Analyte = c("IgG", "IgG")
)

main_plot_02 <- main_plot_01 +
  geom_text(data = annotation_data_g, aes(x = Sample_Type, y = log10.conc, label = label),
            size = 2.2, hjust = 0.4, vjust = -1)

# We annotate per sample type the OOR samples accordingly for total IgM:
annotation_data_m <- data.frame(
  log10.conc = c(1.001, 1.02, 0.8,1.6),
  Sample_Type=c("MLF","MTS","OPS","Saliva"),
  label = c("OOR > (N = 1)\nOOR < (N = 1)", "OOR < (N = 1)", "OOR < (N = 8)","OOR < (N = 1)"),
  Analyte = c("IgM","IgM" ,"IgM","IgM")
)

main_plot_03 <- main_plot_02 +
  geom_text(data = annotation_data_m, aes(x = Sample_Type, y = log10.conc, label = label),
            size = 2.2, hjust = 0.4, vjust = -1)
main_plot_03

# Optionally, save this plot:
# ggsave(filename = "Luminex_Main_Concentrations_Boxplot_28_04_2025.tiff"
#        , plot = main_plot_03, width = 12, height = 6 ,dpi = 300,
#        device="tiff", compression="lzw")

```


# Explore data after substitution:
Here we substitute sample values that were OOR as such:
For samples that were **<OOR**: 1/2 the minimum of the concentration with respect to the measured total antibody.
For samples that were **>OOR**: ULOQ or the maximum value with respect to the measured total antibody. 
```{r Substitutions}

# Function to replace "OOR <" values with half of the minimum or half of LLOQ if all values are NA
replace_missing_with_min <- function(x, lloq) {
  min_val <- min(x, na.rm = TRUE)
  if (is.finite(min_val)) {
    return((1/2) * min_val)
  } else {
    return((1/2) * lloq)  # Use half of LLOQ if all values are NA
  }
}

# Function to replace "OOR >" values with the maximum x2 concentration in range
replace_missing_with_max <- function(x) {
  max_val <- max(x, na.rm = TRUE)
  if (is.finite(max_val)) {
    return(max_val*2)
  } else {
    return(NA)  # If all values are NA, return NA or another desired fallback
  }
}

# Apply the functions per isotype and sample type
mydf2 <- mydf %>%
  group_by(Analyte, Sample_Type) %>%
  mutate(
    concentration = case_when(
      is.na(Conc_in_Range) & Notes == "OOR <" ~ replace_missing_with_min(Conc_in_Range, LLOQ),
      is.na(Conc_in_Range) & Notes == "OOR >" ~ replace_missing_with_max(Conc_in_Range),
      TRUE ~ Conc_in_Range
    )
  ) %>%
  ungroup()

#  Calculate log10 of the new concentration column
mydf2 <- mydf2 %>%
  mutate(log10.new = log10(concentration))

na_rows <- mydf2 %>%
  filter(is.na(concentration)) # no NAs

# Check for complete IDs with all five Sample Types in professional sampling strategy
prof_df2 <- mydf2 %>% 
  filter(Sampling_Strategy == "Professional")

id_with_all_samples <- prof_df2 %>%
  group_by(ID) %>%
  summarize(num_sample_types = n_distinct(Sample_Type)) %>%
  filter(num_sample_types == 5) %>%
  pull(ID)  #033 not complete

unique(id_with_all_samples)

```


# Concentration boxplot with substitution
```{r Plot with subs, fig.width=10, fig.height=8}
ggplot(mydf2, aes(x = Sampling_Method, y = log10.new,fill = Sample_Type)) +
  geom_boxplot(aes(group = Sampling_Method), alpha = 0.5) +  # Boxplot for each analyte
  geom_point(alpha = 0.4) +  # Points with slight dodge
  facet_wrap(~Analyte, ncol = 3) +
  scale_fill_manual(values = custom_colors)+
  labs(x = "Sampling Method", y = "Concentration (log10)",
       title = "Concentration of Analytes per Sample Type",
       subtitle="Professionally Collected (N = 44)",
       fill="Sample Type") +
  theme_bw() + theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.background = element_rect(fill = "transparent", color = NA),    # Transparent panel background
    plot.background = element_rect(fill = "transparent", color = NA),     # Transparent plot background
    legend.background = element_rect(fill = "transparent"),               # Transparent legend background
    strip.background = element_rect(fill = "transparent", color = NA)     # Transparent facet strip background
  )



# Only professional
my_df_prof <- mydf2 %>%  filter(Sampling_Strategy == "Professional")
plot_01 <- ggplot(my_df_prof, aes(x = Sample_Type, y = log10.new,fill = Sample_Type)) +
  geom_boxplot(aes(group = Sample_Type), alpha = 0.5) +  # Boxplot for each analyte
  geom_point(alpha = 0.4) +  # Points with slight dodge
  facet_wrap(~Analyte, ncol = 3) +
  scale_fill_manual(values = custom_colors)+
  labs(x = "Sample Type", y = "Concentration (log10)",
       title = "Concentration of Total Antibodies per Sample Type",
       subtitle="Professionally Collected (N = 44)",
       fill="Sample Type") +
  theme_bw() + theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.background = element_rect(fill = "transparent", color = NA),    # Transparent panel background
    plot.background = element_rect(fill = "transparent", color = NA),     # Transparent plot background
    legend.background = element_rect(fill = "transparent"),               # Transparent legend background
    strip.background = element_rect(fill = "transparent", color = NA)     # Transparent facet strip background
  )

plot_01

# Save plot:
# ggsave(filename = "Luminex_Main_Concentrations_Boxplot_with_subs_28_04_2025.tiff"
#        , plot = plot_01, width = 12, height = 6 ,dpi = 300,
#        device="tiff", compression="lzw")
```

# Make the data into wide format 
```{r Wide df, warning=FALSE}

#----- Subsets--------
#For each isotype, we make subset dataframes:
df_A <- mydf2 %>% 
  filter(Analyte == "IgA") #IgA
df_G <- mydf2 %>% 
  filter(Analyte == "IgG") # IgG

df_M <- mydf2 %>% 
  filter(Analyte == "IgM") #IgM

#-------------------------
# Find `Description` values in each dataframe
desc_A <- df_A$Description
desc_G <- df_G$Description
desc_M <- df_M$Description

missing_in_A <- setdiff(desc_G, desc_A)  # Descriptions in IgG but not in IgA
missing_in_G <- setdiff(desc_A, desc_G)  # Descriptions in IgA but not in IgG
missing_in_M_A <- setdiff(desc_A, desc_M)  # Descriptions in IgA but not in IgM
missing_in_M_G <- setdiff(desc_G, desc_M)  # Descriptions in IgG but not in IgM

missing_in_M_A #because we are missing these samples, we will have to filter them out to ensure complete cases
# Combine results in a list for clarity
missing_descriptions <- list(
  "Missing in IgA" = missing_in_A,
  "Missing in IgG" = missing_in_G,
  "Missing in IgM from IgA" = missing_in_M_A,
  "Missing in IgM from IgG" = missing_in_M_G
)
#-------------------------------
small_df <- prof_df2 %>% 
  subset(select=c(Description,Analyte,Sample_Type,ID,log10.new))

small_df_filt <- small_df %>% group_by(Analyte) %>% 
  filter(!Description %in% missing_in_M_A)

# Finally, we make the wide format dataframe for professionally collected samples:
wide_df <- small_df_filt %>% 
  pivot_wider(names_from = Analyte, values_from = log10.new)

#-------------------------------
# We repeat the process for self-collected samples:
self_df <-mydf2 %>% 
  filter(Sampling_Strategy == "Self")

small_df_self <- self_df %>% 
  subset(select=c(Description,Analyte,Sample_Type,ID, log10.new))

small_df_filt_self <- small_df_self %>% group_by(Analyte) %>% 
  filter(!Description %in% missing_in_M_A)

# We make the self-collected wide dataframe
wide_df_self <- small_df_filt_self %>% 
  pivot_wider(names_from = Analyte, values_from = log10.new)
```


# Comparing the Nostrils
```{r nostrils, warning=FALSE}
# Here we want to compare left and right nostril samples, which were taken from MTS and MLF samples.
#Statistical Significance per Analyte: Nostrils comparison----
mlf_prof <- mydf2 %>% 
  filter(Sample_Type == "MLF") %>% 
  filter(Sampling_Strategy == "Professional")

mts_prof <- mydf2 %>% 
  filter(Sample_Type == "MTS") %>% 
  filter(Sampling_Strategy == "Professional")

# We make a wide format dataframe for the MLf samples:
wide_data_nostril <- mlf_prof %>% select(Nostril,ID,log10.new,Analyte) %>% 
  pivot_wider(names_from = Nostril, 
              values_from = log10.new)

# We make an empty list to store the results of the Wilcoxon paired test 
MLF_wilcoxon_res <- list()
analytes <- unique(mydf2$Analyte)
#Define significance:
significance <- function(p){
  if(p<0.001){
    return("***")
  } else if (p<0.01) {
    return("**")
  } else if(p<0.05){
    return("*")
  } else {
    return("ns")
  }
}
#For-loops
for(analyte in analytes){
  subset_data <- wide_data_nostril %>% 
    filter(Analyte == analyte)
  test_result <- wilcox.test(subset_data$Left, subset_data$Right, paired=TRUE)
  MLF_wilcoxon_res[[analyte]] <- data.frame(
    Analyte=analyte,
    W=test_result$statistic,
    p_value=test_result$p.value,
    pvals_adjusted <- p.adjust(test_result$p.value, method = "BH")
  )
  
}

# Empty list to store p-values
p_values <- c()

# First loop to calculate test results
for(analyte in analytes){
  subset_data <- wide_data_nostril %>% 
    filter(Analyte == analyte)
  test_result <- wilcox.test(subset_data$Left, subset_data$Right, paired=TRUE)
  MLF_wilcoxon_res[[analyte]] <- data.frame(
    Analyte = analyte,
    W = test_result$statistic,
    p_value = test_result$p.value
  )
  p_values <- c(p_values, test_result$p.value)
}

# Adjust p-values for multiple testing (Benjamini-Hochberg)
adjusted_p_values <- p.adjust(p_values, method = "BH")  

# Add adjusted p-values and significance back to the results
for(i in seq_along(analytes)){
  MLF_wilcoxon_res[[analytes[i]]]$adjusted_p_value <- adjusted_p_values[i]
  MLF_wilcoxon_res[[analytes[i]]]$significance <- significance(adjusted_p_values[i])
}


#Combine results into one df:
MLF_Nostril_Wilcoxon_Res <- bind_rows(MLF_wilcoxon_res)

# We make a wide format dataframe for the MTS samples:
wide_data_nostril_mts <- mts_prof %>% select(Nostril,ID,log10.new,Analyte) %>% 
  pivot_wider(names_from = Nostril, 
              values_from = log10.new)

p_values <- c()

MTS_wilcoxon_res <- list()
#For-loops
for(analyte in analytes){
  subset_data <- wide_data_nostril_mts %>% 
    filter(Analyte == analyte)
  test_result <- wilcox.test(subset_data$Left, subset_data$Right, paired=TRUE)
  MTS_wilcoxon_res[[analyte]] <- data.frame(
    Analyte=analyte,
    W=test_result$statistic,
    p_value=test_result$p.value,
    significance=significance(test_result$p.value)
  )
  p_values <- c(p_values, test_result$p.value)
}
# Adjust p-values for multiple testing (Benjamini-Hochberg)
adjusted_p_values_mts <- p.adjust(p_values, method = "BH")  

# Add adjusted p-values and significance back to the results
for(i in seq_along(analytes)){
  MTS_wilcoxon_res[[analytes[i]]]$adjusted_p_value <- adjusted_p_values_mts[i]
  MTS_wilcoxon_res[[analytes[i]]]$significance <- significance(adjusted_p_values_mts[i])
}

#Combine results into one MTS df:

MTS_Nostril_Wilcoxon_Res <- bind_rows(MTS_wilcoxon_res)

```

## Nostrils Plots (paired test version)
```{r,warning=FALSE, fig.width=10, fig.height=8}
# We manually annotate per total isotype:
annotations_sign <- data.frame(
  Analyte = c("IgA", "IgG", "IgM"),
  Nostril = c(1.5, 1.5, 1.5),  
  y_pos = c(6,5.8,4.8),  
  label=c("*", "**", "***")) 
#MLF
paired_nostril_mlf_plot <- ggplot(mlf_prof, aes(x=Nostril, y=log10.new, fill=Nostril)) +
  geom_boxplot(position = position_dodge(width = 0.75), alpha= 0.5) +
  geom_point(alpha=0.5)  +
  facet_wrap(~Analyte, scales="free") +geom_text(data = annotations_sign, aes(x=Nostril, y=y_pos, label=label),
            vjust=-0.5, size=6, inherit.aes = FALSE)+
  geom_line(aes(group=ID), alpha=0.3) +
  labs(title = "Nasosorption Strips (MLF)",
       subtitle = "Nostril Comparison",
       y="Concentration (log10)",
       x="Nostril") +
  theme_bw() 
paired_nostril_mlf_plot

#MTS
paired_nostril_mts_plot <- ggplot(mts_prof, aes(x=Nostril, y=log10.conc, fill=Nostril))+
  geom_boxplot(position = position_dodge(width = 0.75), alpha= 0.5)+
  geom_point(alpha=0.5)+
  facet_wrap(~Analyte, scales="free")+
  geom_line(aes(group=ID), alpha=0.3)+
  labs(title = "Mid-Turbinate Nasal Swab",
       subtitle = "Nostril Comparison",
       y="Concentration (log10)",
       x="Nostril")+
  theme_bw()
paired_nostril_mts_plot
```

## Nostrils: Ratios Comparison
We calculated within-participant log10 ratios between left and right nostrils as log10(Left) - log10(Right), equivalent to log10(Left / Right). These values were used to visualize directional differences between nostrils and tested against zero using a Wilcoxon signed-rank test. 
```{r,warning=FALSE}
# Calculate log ratio for each analyte for MLF samples:
wide_data_nostril <- wide_data_nostril %>%
  mutate(log_ratio = Left - Right)

# Initialize empty list to store results
MLF_wilcoxon_res_ratios <- list()

# Loop through each analyte and perform the Wilcoxon signed-rank test on log ratios
for(analyte in analytes){
  subset_data <- wide_data_nostril %>% 
    filter(Analyte == analyte)
  
  # Perform Wilcoxon test on the log ratio
  test_result <- wilcox.test(subset_data$log_ratio, mu = 0)
  
  # Save results
  MLF_wilcoxon_res_ratios[[analyte]] <- data.frame(
    Analyte = analyte,
    W = test_result$statistic,
    p_value = test_result$p.value
  )
}

# Adjust p-values for multiple testing
p_values <- sapply(MLF_wilcoxon_res_ratios, function(x) x$p_value)
adjusted_p_values <- p.adjust(p_values, method = "BH")

# Add adjusted p-values to results
for(i in seq_along(analytes)){
  MLF_wilcoxon_res_ratios[[analytes[i]]]$adjusted_p_value <- adjusted_p_values[i]
  MLF_wilcoxon_res_ratios[[analytes[i]]]$significance <- significance(adjusted_p_values[i])
}

# Combine results into one data frame
MLF_Nostril_Wilcoxon_Res_Ratios <- bind_rows(MLF_wilcoxon_res_ratios)

# Calculate log ratio for MTS (Mid-turbinate swab) between Left and Right nostrils
wide_data_nostril_mts <- wide_data_nostril_mts %>%
  mutate(log_ratio = Left - Right)

# You can inspect the first few rows to verify
head(wide_data_nostril_mts)

# Initialize empty list to store results
MTS_wilcoxon_res_ratios <- list()

# Loop through each analyte and perform the Wilcoxon signed-rank test on log ratios
for(analyte in analytes){
  subset_data <- wide_data_nostril_mts %>% 
    filter(Analyte == analyte)
  
  # Perform Wilcoxon test on the log ratio
  test_result <- wilcox.test(subset_data$log_ratio, mu = 0)
  
  # Save results
  MTS_wilcoxon_res_ratios[[analyte]] <- data.frame(
    Analyte = analyte,
    W = test_result$statistic,
    p_value = test_result$p.value
  )
}

# Adjust p-values for multiple testing
p_values <- sapply(MTS_wilcoxon_res_ratios, function(x) x$p_value)
adjusted_p_values <- p.adjust(p_values, method = "BH")

# Add adjusted p-values to results
for(i in seq_along(analytes)){
  MTS_wilcoxon_res_ratios[[analytes[i]]]$adjusted_p_value <- adjusted_p_values[i]
  MTS_wilcoxon_res_ratios[[analytes[i]]]$significance <- significance(adjusted_p_values[i])
}

# Combine results into one data frame
MTS_Nostril_Wilcoxon_Res_Ratios <- bind_rows(MTS_wilcoxon_res_ratios)
```

## Nostrils: Ratios Plot (visualize the results)
```{r, warning=FALSE, fig.width=11, fig.height=8}
# Add a column to distinguish MLF and MTS in the data
wide_data_nostril$Sample_Type <- "MLF"
wide_data_nostril_mts$Sample_Type <- "MTS"

# Combine the MLF and MTS data into one dataframe
combined_data <- bind_rows(
  wide_data_nostril %>% select(Analyte, log_ratio, Sample_Type),
  wide_data_nostril_mts %>% select(Analyte, log_ratio, Sample_Type)
)

ratio_nostril_plot <- ggplot(combined_data, aes(x = Analyte, y = log_ratio, fill = Sample_Type)) +
  geom_boxplot(position = position_dodge(width = 0.75), alpha=0.6) +  
 geom_point( 
             position = position_dodge(width = 0.75),  # Align points with respective boxplot
             size = 1.5, alpha = 0.5) +  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +  # Reference line at log ratio =  
  labs(
       x = "Total Antibodies",
       y = "Log Ratio (log10(Left) - log10(Right))",
       fill = "Sample Type") + scale_fill_manual(values = custom_colors)+
  theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text(angle = 45, hjust = 1, size = 22),
    axis.text.y      = element_text(size = 22),
    strip.text       = element_text(size = 25),  # facet labels
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
  )

# visualize plot:  
ratio_nostril_plot

# Create a small annotation data frame to indicate the significance
annotations <- data.frame(
  Analyte = c("IgA", "IgG","IgM"),
  Sample_Type = "MLF",
  y = c(2, 2,2),
  label = c("*", "***", "***")
)

# Add to the existing plot
final_ratio_nostril_plot <- ratio_nostril_plot +
  geom_text(data = annotations,
            aes(x = Analyte, y = y, label = label,group = Sample_Type),
            inherit.aes = FALSE, position = position_nudge(x = -0.2),
            size = 5)
final_ratio_nostril_plot

# Save Plot:
# ggsave(filename = "Luminex_Nostrils_Ratios_28_04_2025.tiff"
#        , plot = final_ratio_nostril_plot, width = 13, height = 11 ,dpi = 300,
#        device="tiff", compression="lzw")
```


# Professional vs Self Collected samples 
This was applied to the subset (N = 16 participants) that additionally self-collected the non-invasive samples (MTS, MLF and saliva).
To assess agreement between collection strategies, we calculated within-participant log10 ratios as log10(Self) - log10(Professional), equivalent to log10(Self / Professional). These values were used to visualize directional differences between collection strategies and tested against zero using a Wilcoxon signed-rank test.
```{r Pro v Self, fig.width=10, fig.height=8}
#Convert columns as factors:
mydf2 <- mydf2 %>%
  convert_as_factor(Analyte, Sampling_Method) %>% 
  group_by(ID, Sample_Type, Analyte) %>% 
  # Calculate the mean for nostrils (Left and Right) for professional samples only
  mutate(mean_concentration = ifelse(Sample_Type %in% c("MLF", "MTS") & !is.na(Nostril), mean(concentration, na.rm = TRUE),  # Mean for Left/Right nostrils
                                     concentration)) %>%  # Keep other samples as they are
  ungroup() %>%
  
  # Remove the duplicated rows by keeping one row for nostrils (Left/Right) per group
  distinct(ID, Sample_Type, Analyte,Sampling_Strategy, .keep_all = TRUE) %>% 
  
  # Apply the log transformation after handling duplicates
  mutate(log_10_mean = log10(mean_concentration))

# subset for self-collected samples
self_df <- mydf2 %>% filter(Sampling_Strategy == "Self")
# calculate participants
self_ids <-  unique(self_df$ID)
#select which samples will be tested
samples_sel <- c("MLF","MTS","Saliva")

# Make dataframe for the self vs professional sampling strategy comparison:
selfvprof <- mydf2 %>% filter(ID %in% self_ids) %>% 
  filter(Sample_Type %in% samples_sel)

# Convert the dataframe to wide format  
wide_data <- selfvprof %>%  dplyr::select(Sampling_Strategy,ID,log_10_mean,Analyte, Sample_Type) %>% 
  pivot_wider(names_from = Sampling_Strategy, 
              values_from = log_10_mean) %>% drop_na()

# Rough visualization of concentrations
ggplot(selfvprof, aes(x = Analyte, y = log_10_mean, fill = Sampling_Strategy)) +
  geom_boxplot(position = position_dodge(width = 0.75), alpha=0.6) +  
 geom_point( 
             position = position_dodge(width = 0.75),  # Align points with respective boxplot
             size = 1.5, alpha = 0.5) +  
  facet_wrap(~Sample_Type)+
  labs(title = "Self vs Professionally Collected Samples",
       x = "Total Antibodies",
       y = "Concentration (log10)",
       fill = "Sampling Strategy") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```


# Log Ratio (Self vs Professionally Collected Samples)

To quantify within-participant differences between collection strategies, log10 ratios were calculated as log10(Self) - log10(Professional), equivalent to log10(Self / Professional). Values above zero indicate higher concentrations in self-collected samples, whereas values below zero indicate higher concentrations in professionally collected samples.
```{r}
# Add log ratio column
combined_strategy_data <- wide_data %>%
  mutate(log_ratio = Self - Professional)
```

# Statistical Test

For each sample type and analyte, log10 ratios were tested against zero using a one-sample Wilcoxon signed-rank test. P-values were adjusted across all self-versus-professional comparisons using the Benjamini–Hochberg method.
```{r}
strategy_wilcox_logratios <- list()
p_values <- c()

for (sample_type in c("MLF", "MTS", "Saliva")) {
  data_sub <- combined_strategy_data %>% filter(Sample_Type == sample_type)
  
  for (analyte in analytes) {
    subset_analyte <- data_sub %>% filter(Analyte == analyte)
    
    test_result <- wilcox.test(subset_analyte$log_ratio, mu = 0)
    
    strategy_wilcox_logratios[[paste0(sample_type, "_", analyte)]] <- data.frame(
      Sample_Type = sample_type,
      Analyte = analyte,
      W = test_result$statistic,
      p_value = test_result$p.value
    )
    
    p_values <- c(p_values, test_result$p.value)
  }
}

results_df_selfvproflog <- combined_strategy_data %>%
  filter(Sample_Type %in% c("MLF","MTS","Saliva")) %>%
  group_by(Sample_Type, Analyte) %>%
  summarise(
    test = list(wilcox.test(log_ratio, mu=0, exact=FALSE)),
    .groups = "drop"
  ) %>%
  mutate(
    W         = map_dbl(test, "statistic"),
    p_value   = map_dbl(test, "p.value")
  ) %>%
  select(-test) %>%
  mutate(
    adjusted_p  = p.adjust(p_value, method="BH"),
    significance = map_chr(adjusted_p, significance)
  )

```

# Visualization of results (Self vs Professionally collected samples)
```{r, warning=FALSE, fig.width=10, fig.height=8}
custom_colors2 <- c("MTS"="#F6EFBD","MLF"="#FFBD73","Saliva"="#8ABFA3")
ratio_strategy_plot <- ggplot(combined_strategy_data, aes(x = Analyte, y = log_ratio, fill = Sample_Type)) +
  geom_boxplot(position = position_dodge(width = 0.75), alpha = 0.6) +
  geom_point(position = position_dodge(width = 0.75), size = 1.5, alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  scale_fill_manual(values = custom_colors2) +
  labs(title = NULL,
       x = "Total Antibodies", y = "Log Ration (log10(Self) - log10(Professional))", fill = "Sample Type") +
  theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text(angle=45, hjust=1, size = 22),
    axis.text.y      = element_text(size = 22),
    strip.text       = element_text(size = 25),  # facet labels
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
  )

ratio_strategy_plot
#

# Save plot:
# ggsave("Log_Ratio_SelfvProf_Total_Abs.tiff", plot = ratio_strategy_plot, dpi = 300, width = 13, height = 10, units = "in", device = "tiff", compression = "lzw" )

```

# Plots: Analytes Across Samples (Isotype Distribution)

To compare total antibody isotype distributions across URT niches, concentrations of IgA, IgG, and IgM were visualized per sample type using paired subject-level profiles overlaid with group-level distributions.
```{r, warning=FALSE, fig.width=10, fig.height=8}

# Create custom colors per isotype:
customs <- c("IgA" = "lightsalmon1",
             "IgG" = "darkolivegreen3",
             "IgM" = "cadetblue2")

library(ggplot2)
library(viridis)  # For viridis color options
library(scales)   # For custom log scale formatting

# rough isotype distribution plot:
analytes_plot <- ggplot(prof_df2, aes(x = Analyte, y = concentration, group = ID)) +
  geom_line(alpha = 0.3, size = 0.6, color = "gray40") + 
  geom_point( size = 2, alpha = 0.7) + 
  geom_boxplot(aes(fill=Analyte,group = Analyte), position = position_dodge(width = 0.8), alpha = 0.5)+
  scale_y_log10(labels = trans_format("log10", math_format(10^.x))) +  # 
  facet_wrap(~Sample_Type, scales = "free_x") +  
  scale_fill_manual(name   = "Total Antibodies",values = customs)+
  labs(
    y = "Log 10 Concentration (ng/mL)",
    x = NULL,
    color = "Total Antibodies"
  ) +theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text(angle = 75, vjust = 1,  hjust = 1, size = 22),
    axis.text.y      = element_text(size = 20),
    strip.text       = element_text(size = 25),  # facet labels
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
    legend.position = "top"
  )
  
analytes_plot



prof_df2_complete <- prof_df2 %>%
  # make sure there's one row per sample
  distinct(Sample_Type, ID, Analyte, .keep_all = TRUE) %>%
  # keep only IDs within each Sample_Type that have all 3 analytes
  group_by(Sample_Type, ID) %>%
  filter(all(c("IgA", "IgG", "IgM") %in% Analyte)) %>%
  ungroup()

#--------------------
library(ggpubr)
# define pairwise comparisons
comparisons <- list(c("IgA", "IgG"),
                    c("IgA", "IgM"),
                    c("IgG", "IgM"))

# final isotype distribution plot
analytes_plot <- ggplot(prof_df2, aes(x = Analyte, y = concentration, fill = Analyte)) +
  geom_boxplot(alpha = 0.6, width = 0.7, outlier.shape = NA) +
  geom_jitter(aes(color = Analyte), width = 0.15, alpha = 0.3, size = 2,color = "black") +
  scale_y_log10(labels = trans_format("log10", math_format(10^.x))) +
  facet_wrap(~Sample_Type, scales = "free_x") +
  scale_fill_manual(values = customs, name = "Total Antibodies") +
  stat_compare_means(
    comparisons = comparisons,
    method = "wilcox.test",
    label = "p.signif",
    hide.ns = FALSE,
    size=6,
    label.y = c(6.5,7,7.5)
  ) +
  labs(
    y = expression("Log"[10]*" Concentration (ng/mL)"),
    x = NULL
  ) +
  theme_bw(base_size = 22) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 22),
    legend.text = element_text(size = 20),
    axis.text.x = element_text(angle = 75, hjust = 1, size = 20),
    axis.text.y = element_text(size = 18),
    strip.text = element_text(size = 22)
  ) +
  guides(fill = guide_legend(title.position = "bottom", title.hjust = 0.5))

analytes_plot

# Save Plot:
 # ggsave("Total_Abs_Pattern_new.tiff", plot = analytes_plot, dpi = 300, width = 13, height = 11, units = "in", device = "tiff", compression = "lzw" )

```


# Log-Linear Mixed Model

To estimate relative differences in total antibody concentrations between professionally collected sample types while accounting for repeated measurements within participants, log10-transformed concentrations were analyzed using linear mixed-effects models. NPS was used as the reference sample type, and participant ID was included as a random intercept. Model estimates were back-transformed to obtain geometric mean ratios (GMRs) and 95% confidence intervals.

## A) Model & Model Checks
```{r, warning=FALSE, fig.align='left', fig.height=8, fig.width=10}

my_df_prof_me <- mydf2 %>% 
  filter(Sampling_Strategy == "Professional") 

my_df_prof_me$Sample_Type<-factor(my_df_prof_me$Sample_Type,
                      levels = c("NPS",
                                 "MTS",
                                 "MLF",
                                 "Saliva",
                                 "OPS"))
# Check residuals for normality for a sample model
# Detach conflicting packages (if applicable)
#detach("package:reticulate", unload = TRUE)

# Run the Mixed-Effects Model:
library(lmerTest)
sample_model <- lmer(log10(concentration) ~ Sample_Type + (1 | ID), data = my_df_prof_me %>% filter(Analyte == analytes[1]))

# Residual diagnostics: Q-Q plot
qqnorm(residuals(sample_model))
qqline(residuals(sample_model), col="red")

# Residuals vs Fitted values plot
plot(fitted(sample_model), residuals(sample_model),
     main = "Residuals vs Fitted",
     xlab = "Fitted values",
     ylab = "Residuals")
abline(h = 0, lty = 2, col = "red")

# Extract random effects for ID from the model
random_effects <- ranef(sample_model)$ID[, "(Intercept)"]  # Extract the random intercept for 'ID'

# Plot Q-Q plot of random effects
qqnorm(random_effects)
qqline(random_effects, col = "red")  # Add a red line to the plot
```

## B) Geometric Mean Ratios: Comparing with the Reference (NPS)
```{r, warning=FALSE, fig.width=11,fig.height=8}
library(broom.mixed) 
set.seed(2024)
# List of unique analytes
analytes <- unique(my_df_prof_me$Analyte)
my_df_prof_me$Sample_Type <- relevel(my_df_prof_me$Sample_Type, ref = "NPS")

# Initialize a list to store results
results_list <- list()

# Loop through each analyte
for (analyte in analytes) {
  # Filter data for the current analyte
  data_analyte <- my_df_prof_me %>%
    filter(Analyte == analyte)
  
  # Fit the log-linear mixed model with simplified random effects
  model_analyte <- lmer(log10(concentration) ~ Sample_Type + (1 | ID), data = data_analyte)
  
  # Tidy the model results
  tidy_model <- tidy(model_analyte, effects = "fixed")
  
  # Calculate confidence intervals for the fixed effects
  conf_intervals <- confint(model_analyte, parm = "beta_", method = "Wald")
  
  # Convert the confidence intervals to a data frame
  conf_intervals_df <- as.data.frame(conf_intervals)
  conf_intervals_df <- conf_intervals_df %>%
    rownames_to_column(var = "term") %>%
    rename(conf.low = `2.5 %`, conf.high = `97.5 %`)
  
  # Merge the tidy model results with the confidence intervals
  tidy_model <- tidy_model %>%
    left_join(conf_intervals_df, by = "term") %>%
    mutate(
      estimate = exp(estimate),
      conf.low = exp(conf.low),
      conf.high = exp(conf.high),
      Analyte = analyte
    )
  
  # Store the results in the list
  results_list[[analyte]] <- tidy_model
}

# Combine all results into a single data frame
all_results <- bind_rows(results_list)
all_results <- all_results %>%
  mutate(term = ifelse(term == "(Intercept)", "NPS", gsub("Sample_Type", "", term)))
# Print the combined results
print(all_results)
```

# Predicted GMs
```{r predicted GMs}
# Calculate absolute geometric means from all_results
absolute_geometric_means <- all_results %>%
  group_by(Analyte) %>%
  mutate(
    # Get the intercept for NPS as the reference mean
    ref_mean = estimate[term == "NPS"],
    # Calculate absolute geometric means by multiplying the reference mean by each term's estimate
    absolute_geometric_mean = ifelse(term == "NPS", ref_mean, ref_mean * estimate),
    # Calculate confidence intervals for the absolute geometric means
    absolute_conf.low = ifelse(term == "NPS", conf.low, ref_mean * conf.low),
    absolute_conf.high = ifelse(term == "NPS", conf.high, ref_mean * conf.high)
  ) %>%
  ungroup() %>%
  select(Analyte, term, absolute_geometric_mean, absolute_conf.low, absolute_conf.high)

# Print the results
print(absolute_geometric_means)


# Combine geometric mean and 95% CI into a single column
absolute_geometric_means_table <- absolute_geometric_means %>%
  mutate(
    Geom_Mean_CI = paste0(
      round(absolute_geometric_mean, 3), " (", 
      round(absolute_conf.low, 3), " - ", 
      round(absolute_conf.high, 3), ")"
    )
  ) %>%
  select(Analyte, term, Geom_Mean_CI)

# Reshape data so that each analyte has its own column
reshaped_geom_means <- absolute_geometric_means_table %>%
  pivot_wider(
    names_from = Analyte, 
    values_from = Geom_Mean_CI
  ) %>%
  rename(Sample_Type = term)

library(gt)
# Create the `gt` table
geom_mean_ci_table <- reshaped_geom_means %>%
  gt(rowname_col = "Sample_Type") %>%
  tab_header(
    title = "Absolute Geometric Mean (95% CI) by Analyte and Sample Type",
    subtitle = "Modeled Data Concentrations (ng/mL)"
  ) %>%
    tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) %>%
  cols_align("center") %>%
    tab_style(
    style = cell_text(weight = "bold"),  # Apply bold text to sample type row labels
    locations = cells_stub()
  )

# Display the table
geom_mean_ci_table
#gtsave(geom_mean_ci_table, "Geom_Mean_CI_Table.docx")
```

## Visualization of Mixed-Effect Model results
```{r, warning=FALSE, fig.width=11,fig.height=8}
library(forcats)

# Reorder Sample_Type (term) factor levels to ensure NPS comes first
all_results <- all_results %>%
  mutate(term = fct_relevel(term, "NPS", "MTS", "MLF", "OPS", "Saliva"))

# Define custom colors for our Sample Types
custom_colors <- c("NPS" = "#A2D2DF", 
                   "MTS" = "#F6EFBD",       
                   "MLF" = "#FFBD73",
                   "OPS" = "#BC7C7C",
                   "Saliva" = "#8ABFA3")


#Plot:
# Modify data: Set NPS ratio to 1
all_results_up <- all_results %>%
  mutate(
    estimate = ifelse(term == "NPS", 1, estimate),  # Set estimate to 1 for NPS
    conf.low = ifelse(term == "NPS", 1, conf.low),   # Set conf.low to 1 for NPS
    conf.high = ifelse(term == "NPS", 1, conf.high)  # Set conf.high to 1 for NPS
  )

all_results_up$term <- relevel(all_results_up$term, ref = "NPS")

# Main figure for mixed-effects model results:
GMR_Pred_Plot_3 <- ggplot(all_results_up, aes(x = term, y = estimate, fill = term)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") + # Reference line at GMR=1
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), 
                position = position_dodge(width = 0.9), 
                width = 0.2, color = "black") +  # Error bar color
  scale_fill_manual(values = custom_colors) +  # Apply custom colors
  labs(
    x = NULL,
    y = "Geometric Mean Ratio",
    fill = "Sample Type") +
  facet_wrap(~Analyte, scales = "free_y") + # Ensure consistent y-axis
  theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text(angle = 75, vjust = 1, hjust = 1, size = 24),
    axis.text.y      = element_text(size = 22),
    strip.text       = element_text(size = 25),  # facet labels
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
  )
  

# Print the plot:
GMR_Pred_Plot_3

# Save the plot: 
# ggsave("Total_Abs_GMR_v1_new.tiff", plot = GMR_Pred_Plot_3, dpi = 300, width = 11, height = 8, units = "in", device = "tiff", compression = "lzw" )
```


## GMR Results Table
Create a table of the mixed-effects model results.
```{r, warning=FALSE}
library(gt)
 # Define a significance function using ifelse
 significance <- function(p) {
   ifelse(p < 0.001, "***", 
          ifelse(p < 0.01, "**", 
                 ifelse(p < 0.05, "*", " ")))
 }
#Make the DF a bit better:
all_results_no_np <- all_results_up %>% 
  filter(term != "NPS")
 
 
GMR_For_Table <- all_results_no_np %>% 
  rename(`Sample Type` = "term") %>% 
  rename(GMR = "estimate") %>% 
  mutate(Significance = significance(p.value))  
  
GMR_For_Table <- GMR_For_Table %>% 
  mutate(GMR_CI = paste0(sprintf("%.2f", GMR), "<br>(",
                         sprintf("%.2f", conf.low), ", ", 
                         sprintf("%.2f", conf.high), ") ", 
                         Significance))
# Pivot the data to have Analytes as columns and Sample Type as rows
wide_table_GMR <- GMR_For_Table %>% 
  subset(select=c(Analyte, GMR_CI, `Sample Type`)) %>% 
  pivot_wider(names_from = Analyte, values_from = GMR_CI)

##### Make the table ##### 
# Create the gt table
gt_table <- wide_table_GMR %>%
  gt(rowname_col = "Sample Type") %>%
  tab_header(
    title = md("**GMR and 95% CI: Analytes by Sample Type**"),
    subtitle = "Professionally Collected"
  ) %>%
  fmt_markdown(columns = everything()) %>%  # Ensure <br> is rendered as HTML
  tab_style(
    style = cell_text(weight = "bold"),  # Apply bold text to analyte column headers
    locations = cells_column_labels(columns = everything())
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),  # Apply bold text to sample type row labels
    locations = cells_stub()
  ) %>%
  cols_align("center") %>%  # Center-align all columns
  tab_style(
    style = cell_text(size = "small"),  # Apply smaller text size to the body
    locations = cells_body(columns = everything())
  )


# Show the table
gt_table
 # gtsave(gt_table, "GMR_table_Total_Antibodies_311024.html")
#Save the table
#gtsave(gt_table, "GMR_table_Total_Antibodies_30062025.docx")
```


## Geometric Means Table (Observed Data)
Make a table to report the geometric mean of the observed data per sample type and isotype.
```{r, warning=FALSE}
# Function to calculate geometric mean
geom_mean <- function(x) exp(mean(log(x), na.rm = TRUE))

# Function to calculate 95% confidence interval for geometric means
geom_ci <- function(x) {
  log_x <- log(x)
  gm <- mean(log_x, na.rm = TRUE)
  se_log <- sd(log_x, na.rm = TRUE) / sqrt(length(log_x[!is.na(log_x)]))
  lower <- gm - 1.96 * se_log
  upper <- gm + 1.96 * se_log
  return(exp(c(lower, upper)))  # Back-transform to original scale
}

# Calculate summary statistics for each analyte and sample type (professionally collected samples)
df_prof <- mydf2 %>% filter(Sampling_Strategy == "Professional")
summary_stats <- my_df_prof %>%
  group_by(Analyte, Sample_Type) %>%
  summarize(
    Geometric_Mean = geom_mean(concentration),
    Geom_CI_Lower = geom_ci(concentration)[1],
    Geom_CI_Upper = geom_ci(concentration)[2],
    Range_Min = min(concentration, na.rm = TRUE),
    Range_Max = max(concentration, na.rm = TRUE),
    n = n()
  )

# Combine Geometric Mean and CI into one column
summary_stats <- summary_stats %>%
  mutate(Geom_Mean_CI = paste0(round(Geometric_Mean, 3), " (", round(Geom_CI_Lower, 3), " - ", round(Geom_CI_Upper, 3), ")"))

# Keep only relevant columns for table
summary_stats_clean <- summary_stats %>%
  select(Sample_Type, Analyte, Geom_Mean_CI, Range_Min, Range_Max, n)

# Reshape data to have Analytes as columns
reshaped_data <- summary_stats_clean %>%
  pivot_wider(
    names_from = Analyte, 
    values_from = c(Geom_Mean_CI, Range_Min, Range_Max, n)
  )

# Combine Geometric Mean with 95% CI
reshaped_data_cleaned <- reshaped_data %>%
  rename_with(.fn = ~ str_replace(., "Geom_Mean_CI_", ""), .cols = starts_with("Geom_Mean_CI")) %>%
  mutate(Sample_Size = select(., starts_with("n_")) %>% rowMeans(na.rm = TRUE)) %>%
  select(-starts_with("n_"))  # Remove individual n_* columns


reshaped_data_cleaned_2 <- reshaped_data_cleaned[,-c(5:10)]

# Create the `gt` table
summary_table <- reshaped_data_cleaned_2 %>%
  gt(rowname_col = "Sample_Type") %>%
  tab_header(
    title = "Geometric Mean (95% CI) by Analyte and Sample Type",
    subtitle = "Observed Data Concentrations (ng/mL)"
  ) %>%
  cols_label(
    Sample_Size = "Sample Size"
  ) %>%
    tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) %>%
  fmt_number(
    columns = everything(),
    decimals = 3
  ) %>% fmt_number(
    columns = vars(Sample_Size),  
    decimals = 0  # No decimal places for Sample Size
  ) %>%  tab_style(
    style = cell_text(weight = "bold"),  
    locations = cells_row_groups()
  ) %>%
  cols_align("center") %>% tab_style(
    style = cell_text(weight = "bold"),  # Apply bold text to sample type row labels
    locations = cells_stub())
# Display the table
summary_table
#Save the table
# gtsave(summary_table, "GM_table_Observed_301024.html")
#gtsave(summary_table, "GM_table_Observed_new.docx")
```

## Descriptive Statistics
```{r, warning=FALSE}
# Compute summary statistics for each analyte and sample type
summary_stats3 <- my_df_prof %>%
  group_by(Analyte, Sample_Type) %>%
  summarize(
    Range_Min = min(concentration, na.rm = TRUE),
    Range_Max = max(concentration, na.rm = TRUE),
    Median_Conc = median(concentration, na.rm = TRUE),
    IQR_Lower = quantile(concentration, 0.25, na.rm = TRUE),
    IQR_Upper = quantile(concentration, 0.75, na.rm = TRUE),
    SD_Conc = sd(concentration, na.rm = TRUE)
  ) %>%
  ungroup()


# Convert all statistics to character format
restructured_stats <- summary_stats3 %>%
  mutate(
    Range = paste0(round(Range_Min, 3), " - ", round(Range_Max, 3)),  # Combine Min and Max into one column
    IQR = paste0(round(IQR_Lower, 3), " - ", round(IQR_Upper, 3)),  # Combine IQR into one column
    Median = as.character(round(Median_Conc, 3)),  # Convert Median to character
    SD = as.character(round(SD_Conc, 3))  # Convert SD to character
  ) %>%
  select(Sample_Type, Analyte, Range, Median, IQR, SD)


restructured_stats_long <- restructured_stats %>%
  pivot_longer(
    cols = c(Range, Median, IQR, SD),
    names_to = "Statistic",
    values_to = "Value"
  ) %>%
  pivot_wider(
    names_from = Analyte,  # Make analytes into columns
    values_from = Value  # Fill values with the statistics
  )


# Create the gt table with sub-rows for each statistic
summary_table3 <-restructured_stats_long %>%
  gt(rowname_col = "Sample_Type") %>%
  tab_header(
    title = "Range, Median, IQR, and SD by Analyte and Sample Type",
    subtitle = "Concentrations (ng/mL)"
  ) %>%
  cols_label(
    Statistic = "Statistic"
    # You can customize the column names for each analyte here
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),  # Make the Sample Type bold
    locations = cells_row_groups()  # Applies bold to the row group labels (Sample Type)
  ) %>%
  cols_align("center")%>%  tab_style(
    style = cell_text(weight = "bold"),  
    locations = cells_row_groups()
  ) %>%
  cols_align("center") %>% tab_style(
    style = cell_text(weight = "bold"),  # Apply bold text to sample type row labels
    locations = cells_stub())%>%
    tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  )  # Center-align the table

# Display the table
summary_table3
#Save the table:
# gtsave(summary_table3, "Descriptive_Stats_table_301024.html")
```



