---
title: "LegendPlex Assay: Data Analysis"
output: 
  html_document:
    toc: true            # Enable Table of Contents
    toc_float: true      # Makes the Table of Contents float (optional)
date: "13-01-2025"
---





# Function to install/load packages:
```{r load_packages function, include=FALSE}
load_packages <- function(packages) {
  sapply(packages, function(pkg) {
    if (!require(pkg, character.only = TRUE)) {
      install.packages(pkg, dependencies = TRUE)
      library(pkg, character.only = TRUE)
    }
  })
}

packages <- c("purrr", "dplyr", "tidyr", "knitr", "magrittr", "reshape2", 
              "tidyverse", "ggplot2", "RColorBrewer", "stringr", 
              "rstatix", "ggpubr", "ggsignif", "FSA",  
              "readxl", "viridis", "patchwork",  "readr",
              "ggfortify", "factoextra", "pheatmap",
              "cowplot","pheatmap","psych","cowplot")

load_packages(packages)
library(gt)
getwd()

```

# Import data:
```{r Import data}
df <- read_excel("02-LegendPlex_data.xlsx")
```


# Transform data into wide format 
```{r}
wide_df <-  df %>% 
  pivot_wider(
    id_cols = c(SampleID, Sampling_Method, Plate),
    names_from = Analytes,
    values_from = concentration.est
  )
```

# Make a new column (log10 concentration transformation)
```{r log10}
df <- df %>% 
  mutate(log10.conc = log10(concentration.est))
```

# Define custom colors per Sample Type
```{r}
# Define custom colors for our Sample Types
custom_colors <- c("Nasopharyngeal swab" = "#A2D2DF", 
                   "Mid-turbinate swab" = "bisque3",       
                   "MLF" = "#FFBD73",
                   "Oropharyngeal swab" = "#BC7C7C",
                   "Saliva" = "#8ABFA3")

```


# Comparing the Nostrils
For MTS and MLF, paired samples were collected from both the left and right nostrils. To assess intra-individual nostril-specific variation, analyte concentrations were compared between nostrils using paired analyses.
```{r nostrils, warning=FALSE}
#Statistical Significance per Analyte: Nostrils comparison----
# MLF
mlf_prof <- df %>% 
  filter(Sample_Type == "MLF") %>% 
  filter(Sampling_Strategy == "Professional")

# MTS
mts_prof <- df %>% 
  filter(Sample_Type == "Mid-turbinate swab") %>% 
  filter(Sampling_Strategy == "Professional")

# MLF wide format
wide_data_nostril <- mlf_prof %>% select(Nostril,ID,log10.conc,Analytes) %>% 
  pivot_wider(names_from = Nostril, 
              values_from = log10.conc)

# Initiate empty list to store MLF results:
MLF_wilcoxon_res <- list()
# Analyte vector 
analytes <- unique(df$Analytes)
#Define significance:
significance <- function(p){
  if(p<0.001){
    return("***")
  } else if (p<0.01) {
    return("**")
  } else if(p<0.05){
    return("*")
  } else {
    return("ns")
  }
}
#For-loops
for(analyte in analytes){
  subset_data <- wide_data_nostril %>% 
    filter(Analytes == analyte)
  test_result <- wilcox.test(subset_data$Left, subset_data$Right, paired=TRUE)
  MLF_wilcoxon_res[[analyte]] <- data.frame(
    Analyte=analyte,
    W=test_result$statistic,
    p_value=test_result$p.value,
    pvals_adjusted <- p.adjust(test_result$p.value, method = "BH")
  )
  
}


# Empty list to store p-values
p_values <- c()

# First loop to calculate test results
for(analyte in analytes){
  subset_data <- wide_data_nostril %>% 
    filter(Analytes == analyte)
  test_result <- wilcox.test(subset_data$Left, subset_data$Right, paired=TRUE)
  MLF_wilcoxon_res[[analyte]] <- data.frame(
    Analyte = analyte,
    W = test_result$statistic,
    p_value = test_result$p.value
  )
  p_values <- c(p_values, test_result$p.value)
}

# Adjust p-values for multiple testing
adjusted_p_values <- p.adjust(p_values, method = "BH")  

# Add adjusted p-values and significance back to the results
for(i in seq_along(analytes)){
  MLF_wilcoxon_res[[analytes[i]]]$adjusted_p_value <- adjusted_p_values[i]
  MLF_wilcoxon_res[[analytes[i]]]$significance <- significance(adjusted_p_values[i])
}




#Combine results into one df:

MLF_Nostril_Wilcoxon_Res <- bind_rows(MLF_wilcoxon_res)

# MTS
# MTS wide format data:
wide_data_nostril_mts <- mts_prof %>% select(Nostril,ID,log10.conc,Analytes) %>% 
  pivot_wider(names_from = Nostril, 
              values_from = log10.conc)

p_values <- c()

MTS_wilcoxon_res <- list()
#For-loops
for(analyte in analytes){
  subset_data <- wide_data_nostril_mts %>% 
    filter(Analytes == analyte)
  test_result <- wilcox.test(subset_data$Left, subset_data$Right, paired=TRUE)
  MTS_wilcoxon_res[[analyte]] <- data.frame(
    Analyte=analyte,
    W=test_result$statistic,
    p_value=test_result$p.value,
    significance=significance(test_result$p.value)
  )
  p_values <- c(p_values, test_result$p.value)
}
# Adjust p-values for multiple testing
adjusted_p_values_mts <- p.adjust(p_values, method = "BH")  

# Add adjusted p-values and significance back to the results
for(i in seq_along(analytes)){
  MTS_wilcoxon_res[[analytes[i]]]$adjusted_p_value <- adjusted_p_values_mts[i]
  MTS_wilcoxon_res[[analytes[i]]]$significance <- significance(adjusted_p_values_mts[i])
}

#Combine results into one df:

MTS_Nostril_Wilcoxon_Res <- bind_rows(MTS_wilcoxon_res)
```

## Nostrils Plots
```{r,warning=FALSE, fig.width=10, fig.height=8}
#MLF
paired_nostril_mlf_plot <- ggplot(mlf_prof, aes(x=Nostril, y=log10.conc, fill=Nostril)) +
  geom_boxplot(position = position_dodge(width = 0.75), alpha= 0.5) +
  geom_point(alpha=0.5) +
  facet_wrap(~Analytes, scales="free") +
  geom_line(aes(group=ID), alpha=0.3) +
  labs(title = "Nasosorption Strips (MLF)",
       subtitle = "Nostril Comparison",
       y="Concentration (log10)",
       x="Nostril") +
  theme_bw()

# View the plot:
paired_nostril_mlf_plot

#MTS
paired_nostril_mts_plot <- ggplot(mts_prof, aes(x=Nostril, y=log10.conc, fill=Nostril))+
  geom_boxplot(position = position_dodge(width = 0.75), alpha= 0.5)+
  geom_point(alpha=0.5)+
  facet_wrap(~Analytes, scales="free")+
  geom_line(aes(group=ID), alpha=0.3)+
  labs(title = "Mid-Turbinate Nasal Swab",
       subtitle = "Nostril Comparison",
       y="Concentration (log10)",
       x="Nostril")+
  theme_bw()
#View the plot:
paired_nostril_mts_plot
```

# Nostrils: Ratios Comparison
Left–right log-ratios were calculated for each analyte and compared against zero.
```{r,warning=FALSE}
# Calculate log ratio for each analyte (MLF)
wide_data_nostril <- wide_data_nostril %>%
  mutate(log_ratio = Left - Right)

# Initialize empty list to store results
MLF_wilcoxon_res_ratios <- list()

# Loop through each analyte and perform the Wilcoxon signed-rank test on log ratios
for(analyte in analytes){
  subset_data <- wide_data_nostril %>% 
    filter(Analytes == analyte)
  
  # Perform Wilcoxon test on the log ratio
  test_result <- wilcox.test(subset_data$log_ratio, mu = 0)
  
  # Save results
  MLF_wilcoxon_res_ratios[[analyte]] <- data.frame(
    Analyte = analyte,
    W = test_result$statistic,
    p_value = test_result$p.value
  )
}

# Adjust p-values for multiple testing
p_values <- sapply(MLF_wilcoxon_res_ratios, function(x) x$p_value)
adjusted_p_values <- p.adjust(p_values, method = "BH")

# Add adjusted p-values to results
for(i in seq_along(analytes)){
  MLF_wilcoxon_res_ratios[[analytes[i]]]$adjusted_p_value <- adjusted_p_values[i]
  MLF_wilcoxon_res_ratios[[analytes[i]]]$significance <- significance(adjusted_p_values[i])
}

# Combine results into one data frame
MLF_Nostril_Wilcoxon_Res_Ratios <- bind_rows(MLF_wilcoxon_res_ratios)

# Calculate log ratio for MTS (Mid-turbinate swab) between Left and Right nostrils
wide_data_nostril_mts <- wide_data_nostril_mts %>%
  mutate(log_ratio = Left - Right)

# You can inspect the first few rows to verify
head(wide_data_nostril_mts)

# Initialize empty list to store results
MTS_wilcoxon_res_ratios <- list()

# Loop through each analyte and perform the Wilcoxon signed-rank test on log ratios
for(analyte in analytes){
  subset_data <- wide_data_nostril_mts %>% 
    filter(Analytes == analyte)
  
  # Perform Wilcoxon test on the log ratio
  test_result <- wilcox.test(subset_data$log_ratio, mu = 0)
  
  # Save results
  MTS_wilcoxon_res_ratios[[analyte]] <- data.frame(
    Analyte = analyte,
    W = test_result$statistic,
    p_value = test_result$p.value
  )
}

# Adjust p-values for multiple testing
p_values <- sapply(MTS_wilcoxon_res_ratios, function(x) x$p_value)
adjusted_p_values <- p.adjust(p_values, method = "BH")

# Add adjusted p-values to results
for(i in seq_along(analytes)){
  MTS_wilcoxon_res_ratios[[analytes[i]]]$adjusted_p_value <- adjusted_p_values[i]
  MTS_wilcoxon_res_ratios[[analytes[i]]]$significance <- significance(adjusted_p_values[i])
}

# Combine results into one data frame
MTS_Nostril_Wilcoxon_Res_Ratios <- bind_rows(MTS_wilcoxon_res_ratios)
```

## Nostrils: Ratios Plot
Visualizing the results of the analysis above.
```{r, warning=FALSE, fig.width=11, fig.height=8}
# Add a column to distinguish MLF and MTS in the data
wide_data_nostril$Sample_Type <- "MLF"
wide_data_nostril_mts$Sample_Type <- "MTS"

# Combine the MLF and MTS data into one dataframe
combined_data <- bind_rows(
  wide_data_nostril %>% select(Analytes, log_ratio, Sample_Type),
  wide_data_nostril_mts %>% select(Analytes, log_ratio, Sample_Type)
)



ratio_nostril_plot <- ggplot(combined_data, aes(x = Analytes, y = log_ratio, fill = Sample_Type)) +
  geom_boxplot(position = position_dodge(width = 0.75), alpha=0.6) +  
 geom_point( 
             position = position_dodge(width = 0.75),  # Align points with respective boxplot
             size = 1.5, alpha = 0.5) +  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +  # Reference line at log ratio = 0
  scale_fill_manual(values = c("MLF"="#FFBD73","MTS"="#F6EFBD"))+
  labs(title =NULL,
       x = "Analytes",
       y = "Log Ratio (log10(Left) - log10(Right))",
       fill = "Sample Type") +
  theme_bw() +
   theme(
    plot.title       = element_text(size = 25, face = "bold"),
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text(angle = 45, vjust = 1, hjust = 1, size = 24),
    axis.text.y      = element_text(size = 24),
    strip.text       = element_text(size = 25),   # facet labels
    legend.position="none"
  ) 

#View the plot:
ratio_nostril_plot
#Save the plot:
# ggsave(
#   filename = "nostrils-legendplex-plot-logratio.tiff",
#   plot     = ratio_nostril_plot,           
#   device   = "tiff",     
#   dpi      = 300,        
#   width    = 14,           
#   height   = 13,
#   units    = "in",
#   compression = "lzw"     
# )
```

From the analysis above, we conclude that it seems that there are **no significant differences** between samples taken from the left or the right nostril with the MLF and MTS sample types. Some cytokines (e.g. IFN-γ, IFN-α2) show outliers, which could be because the IFNs were overall very low in concetration (possibly, close to the limit of detection of the assay, leading to noise or error measurements). Therefore, for the following analysis, we take the mean of the nostrils' concentration (respectively for each sample type).

# Reshape Data
```{r fp, warning=FALSE, fig.width=10, fig.height=8}
Mean_Nostrils_Df <- df %>%
  group_by(ID, Sample_Type, Analytes) %>% 
  # Calculate the mean for nostrils (Left and Right) for professional samples only
  mutate(mean_concentration = ifelse(Sample_Type %in% c("MLF", "Mid-turbinate swab") & !is.na(Nostril), 
                                     mean(concentration.est, na.rm = TRUE),  # Mean for Left/Right nostrils
                                     concentration.est)) %>%  # Keep other samples as they are
  ungroup() %>%
  
  # Remove the duplicated rows by keeping one row for nostrils (Left/Right) per group
  distinct(ID, Sample_Type, Analytes,Sampling_Strategy, .keep_all = TRUE) %>% 
  
  # Apply the log transformation after handling duplicates
  mutate(log_10_mean = log10(mean_concentration))

  
  #Followups: 2 participants were sampled at two timepoints. These data were not included in the manuscript due to low sample size.:
  fp <- Mean_Nostrils_Df %>% 
  filter((Follow_up == "Participant 1" | Follow_up == "Participant 2"))
fp <- fp %>%
  mutate(Followup = paste(Follow_up, Timepoint, sep = "_"))

#Scatterplot for the followups 
# ggplot(fp, aes(x=Analytes, y=log10(concentration.est), color=Sample_Type))+
#   facet_wrap(~Followup)+
#   geom_point(position = position_dodge(width = 0.75))+
#   theme_bw() +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

fp_p1 <- fp %>% filter(Follow_up == "Participant 1")
fp_p2 <- fp %>% filter(Follow_up == "Participant 2")

#Line plot for the followups
line_plot <- ggplot(fp_p1, aes(x = Sample_Type, y = log10(mean_concentration), color = Timepoint, group = Timepoint)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ Analytes, scales = "free_y") +  # Facet by Participant and Analyte
  labs(x = "Sample Type", y = "Concentration (log10)", color = "Timepoint") +
  theme_minimal()+ labs (title="Participant 1")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 
line_plot

fp_p2_plot<- ggplot(fp_p2, aes(x = Sample_Type, y = log10(mean_concentration), color = Timepoint, group = Timepoint)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ Analytes, scales = "free_y") +  # Facet by Participant and Analyte
  labs(x = "Sample Type", y = "Concentration (log10)", color = "Timepoint") +
  theme_minimal()+ labs (title="Participant 2")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 
fp_p2_plot
```


# Define colors for Sample Types
```{r}
custom_colors_abbrv <- c(NPS =  "#A2D2DF",
                         MTS =  "#F6EFBD",
                         MLF = "#FFBD73",
                         OPS = "#BC7C7C",
                         Saliva =  "#8ABFA3")


```


# Boxplots: Concentrations
We want to visualize the analytes' concentrations per sample type 
```{r warning=FALSE, fig.width=10, fig.height=8}

#Convert columns as factors:
my_dataframe <- df %>%
  convert_as_factor(Analytes, Sampling_Method) %>% 
  group_by(ID, Sample_Type, Analytes) %>% 
  # Calculate the mean for nostrils (Left and Right) for professional samples only
  mutate(mean_concentration = ifelse(Sample_Type %in% c("MLF", "Mid-turbinate swab") & !is.na(Nostril), mean(concentration.est, na.rm = TRUE),  # Mean for Left/Right nostrils
                                     concentration.est)) %>%  # Keep other samples as they are
  ungroup() %>%
  
  # Remove the duplicated rows by keeping one row for nostrils (Left/Right) per group
  distinct(ID, Sample_Type, Analytes,Sampling_Strategy, .keep_all = TRUE) %>% 
  
  # Apply the log transformation after handling duplicates
  mutate(log_10_mean = log10(mean_concentration))%>%
  mutate(Degree_of_Invasiveness = as.numeric(factor(Sample_Type, 
                                                    levels = c("Saliva","Oropharyngeal swab",  "MLF", "Mid-turbinate swab", "Nasopharyngeal swab"),
                                                    ordered = TRUE))) %>% mutate(Degree_of_Expression = as.numeric(factor(Sample_Type, 
                                                levels = c("Oropharyngeal swab","Saliva",  "MLF", "Mid-turbinate swab", "Nasopharyngeal swab"),
                                                ordered = TRUE))) 


#Re-level:
my_dataframe$Sampling_Method <- factor(my_dataframe$Sampling_Method, 
                                       levels = c("NPS_Professional",
                                                  "MTS_Professional",
                                                  "MTS_Self",
                                                  "MLF_Professional",
                                                  "MLF_Self",
                                                  "OPS_Professional",
                                                  "Saliva_Professional",
                                                  "Saliva_Self"))

my_dataframe <- my_dataframe %>%
  mutate(Sample_Type = case_when(
    Sample_Type == "Nasopharyngeal swab" ~ "NPS",
    Sample_Type == "Mid-turbinate swab"   ~ "MTS",
    Sample_Type == "Oropharyngeal swab"   ~ "OPS",
    TRUE                                   ~ Sample_Type
  ))

my_dataframe$Sample_Type <- factor(my_dataframe$Sample_Type, 
                                   levels = c("NPS",
                                              "MTS",
                                              "MLF",
                                              "OPS",
                                              "Saliva"
                                              ))
library(scales)
# Only professionally collected samples:
my_dataframe_prof <- my_dataframe %>%  filter(Sampling_Strategy =="Professional")


boxplot_samples_professional <- ggplot(my_dataframe_prof,
     aes(x = Sample_Type, y = log_10_mean, fill = Sample_Type)
  ) +
  geom_point(alpha = 0.5) +
  geom_boxplot(position = position_dodge(width = 0.75), alpha = 0.5) +
  geom_line(aes(group = ID), alpha = 0.1) +
  facet_wrap(~Analytes) +

  # set a bigger base size here
  theme_bw(base_size = 25) +  

  # then fine‐tune specific text sizes
  theme(
    plot.title       = element_text(size = 25, face = "bold"),
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text(angle = 75, vjust = 1, hjust = 1, size = 24),
    axis.text.y      = element_text(size = 24),
    strip.text       = element_text(size = 25),   # facet labels
    legend.position="none"
  ) +

  scale_fill_manual(values = custom_colors_abbrv) +
  scale_y_continuous(labels = math_format(10^.x)) +
  labs(
    y     = "Log 10 Concentration (pg/mL)",
    x     = NULL,
    fill  = NULL
  )

# View plot:
boxplot_samples_professional


# Save plot:
# ggsave(
#   filename = "LegendPlex_Boxplot_only_prof-biggertext.tiff",
#   plot     = boxplot_samples_professional,           
#   device   = "tiff",     
#   dpi      = 300,        
#   width    = 14,          
#   height   = 13,
#   units    = "in",
#   compression = "lzw"     
# )
# 

```

# PCA
```{r, warning=FALSE, fig.width=10, fig.height=8}
#darker colors for PCA:
custom_colors_abbrv <- c(NPS = "#5BA4C7",
                         MTS = "#D4B800",
                         MLF = "#F09030",
                         OPS = "#A84444",
                         Saliva = "#4A9E7A")

# Make regions instead of sample types
my_df_prof <- my_dataframe_prof %>%
  mutate(Anatomical_Site = case_when(
    Sample_Type == "NPS" ~ "Nasopharynx",
    Sample_Type == "OPS" ~ "Oropharynx",
    Sample_Type == "MTS" ~ "Mid-Turbinate Nasal Mucosa",
    Sample_Type == "MLF" ~ "Nasal Mucosa",
    Sample_Type == "Saliva" ~ "Oral Cavity",
    TRUE ~ NA_character_  # For any other cases that don't match
  ))


library(factoextra)

#Prepare the data (we will be using only the Professionally collected)
wide_df_new2 <-  my_df_prof %>% 
  pivot_wider(
    id_cols = c(SampleID, Sample_Type, Plate),
    names_from = Analytes,
    values_from = mean_concentration
  )



pca_result2 <- prcomp(wide_df_new2[,-c(1,2,3)], scale. = TRUE)
p2 <- as.data.frame(pca_result2$x)
summary(pca_result2)

#Scree Plot (explained variance)
fviz_eig(pca_result2, 
         addlabels = TRUE, 
         ylim = c(0, 70),
         main="Scree Plot - Variance Explained (%)")

#Scree Plot (eigenvalues)
fviz_eig(pca_result2, 
         addlabels = TRUE, 
         choice="eigenvalue",
         main="Scree Plot - Eigenvalues") +
  geom_hline(yintercept=1, 
             linetype="dashed", 
             color = "red")

# Loadings (rotation matrix)
pca_loadings <- as.data.frame(pca_result2$rotation)

# Contributions (scaled importance of each variable to each PC)
var_contrib <- factoextra::get_pca_var(pca_result2)
head(var_contrib$contrib)   # % contribution per variable
head(var_contrib$cos2)      # squared cosines (quality of representation)

summary(pca_result2)   # shows proportion & cumulative variance explained

# PERMANOVA test on Euclidean distances
library(vegan)

# Distance matrix for PERMANOVA
dist_mat <- dist(wide_df_new2[ , -c(1:3)], method = "euclidean")

# PERMANOVA (effect of Sample_Type)
permanova_res <- adonis2(dist_mat ~ Sample_Type, data = wide_df_new2, permutations = 999)
print(permanova_res)

# Take first few PCs (that explain most variance)
pc_scores <- as.data.frame(pca_result2$x[,1:5])  
pc_scores$Sample_Type <- wide_df_new2$Sample_Type


# Extract stats
permanova_F  <- signif(permanova_res$F[1], 3)
permanova_R2 <- signif(permanova_res$R2[1], 3)
permanova_p  <- permanova_res$`Pr(>F)`[1]

# format p-value nicely
permanova_p_label <- ifelse(permanova_p < 0.001, "< 0.001", signif(permanova_p, 3))

# Add result to PCA plot


# Extract variable contributions
var_contrib <- get_pca_var(pca_result2)

# Top 10 contributors for PC1
fviz_contrib(pca_result2, choice = "var", axes = 1, top = 10)

# Top 10 contributors for PC2
fviz_contrib(pca_result2, choice = "var", axes = 2, top = 10)

# Highlight top 10 contributors across PCs
new_pca <- fviz_pca_biplot(
  pca_result2, 
  select.var = list(contrib = 5),  # show only top contributors
  repel = TRUE,
  label = "var",
  labelsize = 10, 
  col.var = "gray20",
  addEllipses = TRUE, 
  ellipse.type = "t", 
  ellipse.alpha = 0.4,
  pointsize = 3,               # add this - bigger points
  alpha.ind = 0.75,            # add this - more opaque points
  col.ind      = "black", 
  habillage    = wide_df_new2$Sample_Type,
  palette      = "lancet"
) +
  labs(color = "Sample Type", shape = "Sample Type", fill = "Sample Type") +
  scale_fill_manual(values = custom_colors_abbrv) +
  scale_color_manual(values = custom_colors_abbrv) +
  theme_bw() +
  labs(title = NULL) +
  theme(
    axis.title.x  = element_text(size = 24),  # x-axis label
    axis.title.y  = element_text(size = 24),  # y-axis label
    axis.text.x   = element_text(size = 18),  # x-axis numbers
    axis.text.y   = element_text(size = 18),  # y-axis numbers
    legend.text   = element_text(size = 22),
    legend.title  = element_text(size = 22),
    legend.position = "right"
  ) +
  annotate("text",
           x = -20, y = 8,   # adjust to your PCA scale
           label = paste0("PERMANOVA:\n F = ", permanova_F,
                          ", R² = ", permanova_R2,
                          ",\n p = ", permanova_p_label),
           size = 8, hjust = 0)

# View PCA plot:
new_pca
#Save PCA plot:
#ggsave("PCA_LegendPlex.tiff", plot = new_pca, width = 12, height = 9,  dpi = 300)
```

# Professional vs Self Collected
```{r pro v self, warning=FALSE}
#Firstly, we extract the IDs of those that participated in the self collection
self_df <- my_dataframe %>% 
  filter(Sampling_Strategy == "Self")
self_ids <- unique(self_df$ID)
self_ids

# Next, we set Nostril to NA for "Self" samples
pre_filt <- my_dataframe %>% 
  mutate(Nostril = ifelse(Sampling_Strategy == "Self", NA, Nostril))  #

# Filter for participants with "Self" samples 
filtered_df <- pre_filt %>% 
  filter(ID %in% self_ids)
  
# Filter for each sample type 
filtered_df_noninv1 <- filtered_df %>% 
  filter(Sample_Type  == "MLF") 
filtered_df_noninv2 <- filtered_df %>% 
  filter(Sample_Type  == "MTS")
filtered_df_noninv3 <- filtered_df %>% 
  filter(Sample_Type  == "Saliva")

# Transform them into wide format
wide_data_filtered_mlf <- filtered_df_noninv1 %>% select(Sampling_Strategy,ID,log_10_mean,Analytes) %>% 
  pivot_wider(id_cols = c(ID, Analytes),
              names_from = Sampling_Strategy, 
              values_from = log_10_mean)

wide_data_filtered_mts <- filtered_df_noninv2 %>% select(Sampling_Strategy,ID,log_10_mean,Analytes) %>% 
  pivot_wider(names_from = Sampling_Strategy, 
              values_from = log_10_mean)

wide_data_filtered_saliva <- filtered_df_noninv3 %>% select(Sampling_Strategy,ID,log_10_mean,Analytes) %>% 
  pivot_wider(names_from = Sampling_Strategy, 
              values_from = log_10_mean)

# Initiate empty lists to store the results
Strategy_wilcoxon_res_mlf <- list()
Strategy_wilcoxon_res_mts <- list()
Strategy_wilcoxon_res_saliva <- list()
analytes <- unique(my_dataframe$Analytes)

#Define significance:
significance <- function(p){
  if(p<0.001){
    return("***")
  } else if (p<0.01) {
    return("**")
  } else if(p<0.05){
    return("*")
  } else {
    return("ns")
  }
}

# Empty list to store p-values
p_values <- c()

# First loop to calculate test results
for(analyte in analytes){
  subset_data <- wide_data_filtered_mlf %>% 
    filter(Analytes == analyte)
  test_result <- wilcox.test(subset_data$Professional, subset_data$Self, paired=TRUE, conf.int = TRUE)
  
  conf_int_lower <- 10^test_result$conf.int[1]
  conf_int_upper <- 10^test_result$conf.int[2] 
  
  
  Strategy_wilcoxon_res_mlf[[analyte]] <- data.frame(
    Analyte = analyte,
    W = test_result$statistic,
    p_value = test_result$p.value,
    conf_int_lower = 10^test_result$conf.int[1],
    conf_int_upper =10^test_result$conf.int[2]
  )
  p_values <- c(p_values, test_result$p.value)
}

# Adjust p-values for multiple testing with BH method
adjusted_p_values <- p.adjust(p_values, method = "BH")  

# Add adjusted p-values and significance back to the results
for(i in seq_along(analytes)){
  Strategy_wilcoxon_res_mlf[[analytes[i]]]$adjusted_p_value <- adjusted_p_values[i]
  Strategy_wilcoxon_res_mlf[[analytes[i]]]$significance <- significance(adjusted_p_values[i])
}

# Combine the results into a single data frame
Self_v_Prof_MLF <- do.call(rbind, Strategy_wilcoxon_res_mlf)


## MTS
p_values <- c()  # Reset p_values before the loop

# Loop to calculate test results
for(analyte in analytes){
  subset_data2 <- wide_data_filtered_mts %>% 
    filter(Analytes == analyte)
  test_result2 <- wilcox.test(subset_data2$Professional, subset_data2$Self, paired=TRUE, conf.int = TRUE)
  Strategy_wilcoxon_res_mts[[analyte]] <- data.frame(
    Analyte = analyte,
    W = test_result2$statistic,
    p_value = test_result2$p.value,
    conf_int_lower = 10^test_result2$conf.int[1],
    conf_int_upper =10^test_result2$conf.int[2]
  )
  p_values <- c(p_values, test_result2$p.value)
}

# Adjust p-values for multiple testing with BH method
adjusted_p_values <- p.adjust(p_values, method = "BH")  

# Add adjusted p-values and significance back to the results
for(i in seq_along(analytes)){
  Strategy_wilcoxon_res_mts[[analytes[i]]]$adjusted_p_value <- adjusted_p_values[i]
  Strategy_wilcoxon_res_mts[[analytes[i]]]$significance <- significance(adjusted_p_values[i])
}

# Combine the results into a single data frame
Self_v_Prof_MTS <- do.call(rbind, Strategy_wilcoxon_res_mts)

## Saliva
p_values <- c()  # Reset p_values before the loop

for(analyte in analytes){
  subset_data <- wide_data_filtered_saliva %>% 
    filter(Analytes == analyte)
  test_result <- wilcox.test(subset_data$Professional, subset_data$Self, paired=TRUE, conf.int = TRUE)
  Strategy_wilcoxon_res_saliva[[analyte]] <- data.frame(
    Analyte = analyte,
    W = test_result$statistic,
    p_value = test_result$p.value,
    conf_int_lower = 10^test_result$conf.int[1],
    conf_int_upper = 10^test_result$conf.int[2]
  )
  p_values <- c(p_values, test_result$p.value)
}

adjusted_p_values <- p.adjust(p_values, method = "BH")  

for(i in seq_along(analytes)){
  Strategy_wilcoxon_res_saliva[[analytes[i]]]$adjusted_p_value <- adjusted_p_values[i]
  Strategy_wilcoxon_res_saliva[[analytes[i]]]$significance <- significance(adjusted_p_values[i])
}

Self_v_Prof_Saliva <- do.call(rbind, Strategy_wilcoxon_res_saliva)


```

# Log Ratio (Self v Professionally collected samples)
```{r}
# Add Sample_Type to each df
wide_data_filtered_mlf$Sample_Type <- "MLF"
wide_data_filtered_mts$Sample_Type <- "MTS"
wide_data_filtered_saliva$Sample_Type <- "Saliva"

# Combine into one df
combined_strategy_data <- bind_rows(
  wide_data_filtered_mlf,
  wide_data_filtered_mts,
  wide_data_filtered_saliva
)

# Add log ratio column
combined_strategy_data <- combined_strategy_data %>%
  mutate(log_ratio = Self - Professional)
```

# Perform test
```{r}
strategy_wilcox_logratios <- list()
p_values <- c()

for (sample_type in c("MLF", "MTS", "Saliva")) {
  data_sub <- combined_strategy_data %>% filter(Sample_Type == sample_type)
  
  for (analyte in analytes) {
    subset_analyte <- data_sub %>% filter(Analytes == analyte)
    
    test_result <- wilcox.test(subset_analyte$log_ratio, mu = 0)
    
    strategy_wilcox_logratios[[paste0(sample_type, "_", analyte)]] <- data.frame(
      Sample_Type = sample_type,
      Analyte = analyte,
      W = test_result$statistic,
      p_value = test_result$p.value
    )
    
    p_values <- c(p_values, test_result$p.value)
  }
}

adjusted_p <- p.adjust(p_values, method = "BH")
results_df_selfvproflog <- bind_rows(strategy_wilcox_logratios)

results_df_selfvproflog$adjusted_p <- adjusted_p
results_df_selfvproflog$significance <- sapply(adjusted_p, significance)
```

# Visualize the results
```{r, warning=FALSE, fig.width=10, fig.height=8}
custom_colors2 <- c("MTS"="#F6EFBD","MLF"="#FFBD73","Saliva"="#8ABFA3")
ratio_strategy_plot <- ggplot(combined_strategy_data, aes(x = Analytes, y = log_ratio, fill = Sample_Type)) +
  geom_boxplot(position = position_dodge(width = 0.75), alpha = 0.6) +
  geom_point(position = position_dodge(width = 0.75), size = 1.5, alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  scale_fill_manual(values = custom_colors2) +
  labs(title=NULL,
       x = "Analytes", y = "Log Ratio (log10(Self) - log10(Professional))", fill = "Sample Type") +
  theme_bw() +
  theme(
    plot.title       = element_text(size = 25, face = "bold"),
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text(angle = 45, vjust = 1, hjust = 1, size = 24),
    axis.text.y      = element_text(size = 24),
    strip.text       = element_text(size = 25),   # facet labels
    legend.position="none"
  ) 

 # Create a small annotation data frame
annotations <- data.frame(
  Analytes = c("CCL2", "CXCL1"),
  Sample_Type = "Saliva",
  y = c(2, 2),
  label = c("**", "*")
)

# Add to your existing plot
final_ratio_strategy_plot <- ratio_strategy_plot +
  geom_text(data = annotations,
            aes(x = Analytes, y = y, label = label,group = Sample_Type),
            inherit.aes = FALSE,
             position = position_nudge(x = 0.2),
            size = 5)
#View plot:
final_ratio_strategy_plot

# Save plot:
# ggsave("Log_Ratio_SelfvProf.tiff", plot = final_ratio_strategy_plot, dpi = 300, width = 13, height = 11, units = "in", device = "tiff", compression = "lzw" )

```



## Paired sample plots of Professional vs Self Collected
```{r, warning=FALSE, fig.width=10, fig.height=8}
####plots----
# Annotations for significance
annotations_sign1 <- data.frame(
  Analytes = c("IL-6"),
  Sampling_Strategy = c(1.5),  
  y_pos = c(2),  
  label=c("*")) 
  
mlf_selfvsprof<- ggplot(filtered_df_noninv1, aes(x=Sampling_Strategy, y=log10(mean_concentration), fill=Sampling_Strategy))+
  geom_point(alpha=0.5) +
  geom_boxplot(position = position_dodge(width = 0.75), alpha= 0.5)+
  geom_line(aes(group=ID), alpha=0.3)+
  facet_wrap(~Analytes, ncol=6)+theme_bw() +
  geom_text(data = annotations_sign1, aes(x=Sampling_Strategy, y=y_pos, label=label),
            vjust=-0.5, size=6, inherit.aes = FALSE) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, size = 8, hjust = 1))+ scale_y_continuous(breaks = c(-3,-2,-1,0,1,2,3,4,5,6))+ ylim(-3,5)+
  labs(title="MLF: Self vs Professional Collection",
       y="Concentration (log10)",
       x="Sampling Strategy")
  
mlf_selfvsprof


annotations_sign2 <- data.frame(
  Analytes = c("APRIL","CCL2","CCL4","IFN-α2","IFN-λ1"),
  Sampling_Strategy = c(1.5,1.5,1.5,1.5,1.5),  
  y_pos = c(3.7,2.7,2.7,1.5,2),  
  label=c("*","**","*","*","*")) 

mts_selfvsprof<- ggplot(filtered_df_noninv2, aes(x=Sampling_Strategy, y=log10.conc, fill=Sampling_Strategy))+
  geom_point(alpha=0.5) +
  geom_boxplot(position = position_dodge(width = 0.75), alpha= 0.5)+
  geom_line(aes(group=ID), alpha=0.3)+ scale_y_continuous(breaks = c(-3,-2,-1,0,1,2,3,4,5,6))+ ylim(-3,5)+
  geom_text(data = annotations_sign2, aes(x=Sampling_Strategy, y=y_pos, label=label),
            vjust=-0.5, size=6, inherit.aes = FALSE) +
  facet_wrap(~Analytes,ncol=6)+theme_bw() +theme(axis.text.x = element_text(angle = 45, vjust = 1, size = 8, hjust = 1))+
  labs(title="Mid-Turbinate Nasal Swab: Self vs Professional Collection",
       y="Concentration (log10)",
       x="Sampling Strategy")
mts_selfvsprof


annotations_sign3 <- data.frame(
  Analytes = c("APRIL","CCL2","CCL4","CXCL1","IFN-λ1","Perforin"),
  Sampling_Strategy = c(1.5,1.5,1.5,1.5,1.5,1.5),  
  y_pos = c(3.7,2.7,2.7,2.7,2.7,2.7),  
  label=c("*","**","*","**","*","*")) 

sal_selfvsprof<- ggplot(filtered_df_noninv3, aes(x=Sampling_Strategy, y=log_10_mean, fill=Sampling_Strategy))+
  geom_point(alpha=0.5) +
  geom_boxplot(position = position_dodge(width = 0.75), alpha= 0.5)+
  geom_line(aes(group=ID), alpha=0.3)+ scale_y_continuous(breaks = c(-3,-2,-1,0,1,2,3,4,5,6))+ ylim(-3,5)+
  facet_wrap(~Analytes, ncol=6)+theme_bw() +geom_text(data = annotations_sign3, aes(x=Sampling_Strategy, y=y_pos, label=label),
                                                      vjust=-0.5, size=6, inherit.aes = FALSE) +
  theme(axis.text.x = element_text(angle = 55, vjust = 1, size = 8, hjust = 1))+
  labs(title="Saliva: Self vs Professional Collection",
       y="Concentration (log10)",
       x="Sampling Strategy")
sal_selfvsprof
```

## Geometric Means Table (Observed Data)
```{r, warning=FALSE}
# Function to calculate geometric mean
geom_mean <- function(x) exp(mean(log(x), na.rm = TRUE))

# Function to calculate 95% confidence interval for geometric means
geom_ci <- function(x) {
  log_x <- log(x)
  gm <- mean(log_x, na.rm = TRUE)
  se_log <- sd(log_x, na.rm = TRUE) / sqrt(length(log_x[!is.na(log_x)]))
  lower <- gm - 1.96 * se_log
  upper <- gm + 1.96 * se_log
  return(exp(c(lower, upper)))  # Back-transform to original scale
}

# Calculate summary statistics for each analyte and sample type
summary_stats <- my_df_prof %>%
  group_by(Analytes, Sample_Type) %>%
  summarize(
    Geometric_Mean = geom_mean(mean_concentration),
    Geom_CI_Lower = geom_ci(mean_concentration)[1],
    Geom_CI_Upper = geom_ci(mean_concentration)[2],
    Range_Min = min(mean_concentration, na.rm = TRUE),
    Range_Max = max(mean_concentration, na.rm = TRUE),
    n = n()
  )

# Combine Geometric Mean and CI into one column
summary_stats <- summary_stats %>%
  mutate(Geom_Mean_CI = paste0(round(Geometric_Mean, 3), " (", round(Geom_CI_Lower, 3), " - ", round(Geom_CI_Upper, 3), ")"))

# Keep only relevant columns for table
summary_stats_clean <- summary_stats %>%
  select(Sample_Type, Analytes, Geom_Mean_CI, Range_Min, Range_Max, n)

# Reshape data to have Analytes as columns
reshaped_data <- summary_stats_clean %>%
  pivot_wider(
    names_from = Analytes, 
    values_from = c(Geom_Mean_CI, Range_Min, Range_Max, n)
  )

# Combine Geometric Mean with 95% CI
reshaped_data_cleaned <- reshaped_data %>%
  rename_with(.fn = ~ str_replace(., "Geom_Mean_CI_", ""), .cols = starts_with("Geom_Mean_CI")) %>%
  mutate(Sample_Size = select(., starts_with("n_")) %>% rowMeans(na.rm = TRUE)) %>%
  select(-starts_with("n_"))  # Remove individual n_* columns


reshaped_data_cleaned_2 <- reshaped_data_cleaned[,-c(15:40)]

# Step 4: Create the `gt` table

summary_table <- reshaped_data_cleaned_2 %>%
  gt(rowname_col = "Sample_Type") %>%
  tab_header(
    title = "Geometric Mean (95% CI) by Analyte and Sample Type",
    subtitle = "Observed Data Concentrations (pg/mL)"
  ) %>%
  cols_label(
    Sample_Size = "Sample Size"
  ) %>%
    tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) %>%
  fmt_number(
    columns = everything(),
    decimals = 3
  ) %>% fmt_number(
    columns = vars(Sample_Size),  
    decimals = 0  # No decimal places for Sample Size
  ) %>%  tab_style(
    style = cell_text(weight = "bold"),  
    locations = cells_row_groups()
  ) %>%
  cols_align("center") %>% tab_style(
    style = cell_text(weight = "bold"),  # Apply bold text to sample type row labels
    locations = cells_stub())
# Display the table
summary_table
# gtsave(summary_table, "GM_table_Observed_final.html")
gtsave(summary_table, "GM_table_Observed_final.docx")


```

## Descriptive Statistics
```{r, warning=FALSE}
# Compute summary statistics for each analyte and sample type
summary_stats3 <- my_df_prof %>%
  group_by(Analytes, Sample_Type) %>%
  summarize(
    Range_Min = min(mean_concentration, na.rm = TRUE),
    Range_Max = max(mean_concentration, na.rm = TRUE),
    Median_Conc = median(mean_concentration, na.rm = TRUE),
    IQR_Lower = quantile(mean_concentration, 0.25, na.rm = TRUE),
    IQR_Upper = quantile(mean_concentration, 0.75, na.rm = TRUE),
    SD_Conc = sd(mean_concentration, na.rm = TRUE)
  ) %>%
  ungroup()


# Convert all statistics to character format
restructured_stats <- summary_stats3 %>%
  mutate(
    Range = paste0(round(Range_Min, 3), " - ", round(Range_Max, 3)),  # Combine Min and Max into one column
    IQR = paste0(round(IQR_Lower, 3), " - ", round(IQR_Upper, 3)),  # Combine IQR into one column
    Median = as.character(round(Median_Conc, 3)),  # Convert Median to character
    SD = as.character(round(SD_Conc, 3))  # Convert SD to character
  ) %>%
  select(Sample_Type, Analytes, Range, Median, IQR, SD)


restructured_stats_long <- restructured_stats %>%
  pivot_longer(
    cols = c(Range, Median, IQR, SD),
    names_to = "Statistic",
    values_to = "Value"
  ) %>%
  pivot_wider(
    names_from = Analytes,  # Make analytes into columns
    values_from = Value  # Fill values with the statistics
  )


# Create the gt table with sub-rows for each statistic
summary_table3 <-restructured_stats_long %>%
  gt(rowname_col = "Sample_Type") %>%
  tab_header(
    title = "Range, Median, IQR, and SD by Analyte and Sample Type",
    subtitle = "Concentrations (pg/mL)"
  ) %>%
  cols_label(
    Statistic = "Statistic"
    # You can customize the column names for each analyte here
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),  # Make the Sample Type bold
    locations = cells_row_groups()  # Applies bold to the row group labels (Sample Type)
  ) %>%
  cols_align("center")%>%  tab_style(
    style = cell_text(weight = "bold"),  
    locations = cells_row_groups()
  ) %>%
  cols_align("center") %>% tab_style(
    style = cell_text(weight = "bold"),  # Apply bold text to sample type row labels
    locations = cells_stub())%>%
    tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  )  # Center-align the table

# Display the table
summary_table3
gtsave(summary_table3, "Descriptive_Stats_table_final.html")

```


# I. Log-Linear Mixed Effects Model

## I1. Model 
```{r , warning=FALSE, fig.align='left', fig.height=8, fig.width=10}

library(nlme)
library(effects)
library(ggeffects)
library(marginaleffects)


set.seed(2025)
my_dataframe$Sample_Type <- factor(my_dataframe$Sample_Type, 
                                   levels = c("NPS",
                                              "MTS",
                                              "MLF",
                                              "OPS",
                                              "Saliva"
                                   ))

my_df_prof_me <- my_dataframe %>% 
  filter(Sampling_Strategy == "Professional") 

my_df_prof_me$Sample_Type<-factor(my_df_prof_me$Sample_Type,
                                  levels = c("NPS",
                                             "MTS",
                                             "MLF",
                                             "Saliva",
                                             "OPS"))

my_df_prof_me$Sample_Type <- relevel(my_df_prof_me$Sample_Type, ref = "NPS")


my_df_prof_me <- my_df_prof_me %>% 
  mutate(log.conc=log(concentration.est)) #log-transform the concentrations

# calculating ratios for each individual
my_df_prof_me <- my_df_prof_me %>%
  group_by(Analytes, ID) %>% # Group by Analyte
  mutate(
    naso_log = log.conc[Sample_Type == "NPS"], # Extract Nasopharyngeal swab log value
    log_ratio = log.conc - naso_log # Calculate log ratio for each Sample_Type
  ) %>%
  ungroup() 



#Applying the model
ctrl <- lmeControl(opt='optim')

combined_model_2 <- lme(log_ratio ~  Sample_Type:Analytes-1, random = ~ Sample_Type|ID, data = my_df_prof_me, control = ctrl, method = "REML")

# different ways to extract the outcome:
summary(combined_model_2)
intervals(combined_model_2, which = "fixed"); exp(intervals(combined_model_2, which = "fixed")$fixed)



#Calculating GMRs & CIs

library(dplyr)
library(stringr)

# Extract the fixed effects table
fixed_effects <- summary(combined_model_2)$tTable
fixed_effects_df <- as.data.frame(fixed_effects)

# Add rownames as a column for parsing Sample_Type and Analytes
fixed_effects_df <- fixed_effects_df %>%
  mutate(
    Combination = rownames(fixed_effects_df),
    Sample_Type = str_extract(Combination, "(?<=Sample_Type)[^:]+"),
    Analytes = str_extract(Combination, "(?<=Analytes).*")
  )


# Join reference rows back to the full data
adjusted_df <- fixed_effects_df %>%
  mutate(
    GMR = exp(Value),                   # Convert to GMR
    GMR_Lower = exp(Value - (1.96 * Std.Error)), # CI lower bound
    GMR_Upper = exp(Value + (1.96 * Std.Error))  # CI upper bound
  )

# Filter out reference rows if needed
GMRs <- adjusted_df %>%
  filter(Sample_Type != "NPS") %>%
  select(Sample_Type, Analytes, GMR, GMR_Lower, GMR_Upper, `p-value`)

# View the resulting data frame
print(GMRs)

# Filter out reference rows if needed
GMRs_with_NP <- adjusted_df %>%
  select(Sample_Type, Analytes, GMR, GMR_Lower, GMR_Upper, `p-value`)

```



## I2. Visualization
```{r new_vizual, warning=FALSE, fig.width=11,fig.height=8}
library(forcats)

# Reorder Sample_Type (term) factor levels to ensure NP comes first
GMRs_with_NP <- GMRs_with_NP %>%
  mutate(Sample_Type = fct_relevel(Sample_Type, "NPS"))
GMRs_with_NP <- GMRs_with_NP %>%
  mutate(Sample_Type = fct_relevel(Sample_Type, "NPS", "MTS", "MLF", "OPS", "Saliva"))
# Define custom colors for your Sample Types
custom_colors <- c("NPS" = "#A2D2DF", 
                   "MTS" = "#F6EFBD",       
                   "MLF" = "#FFBD73",
                   "OPS" = "#BC7C7C",
                   "Saliva" = "#8ABFA3")


#Updated Plot:
# Modify data: Set NPS ratio to 1
GMRs_with_NP <- GMRs_with_NP %>%
  mutate(
    GMR = ifelse(Sample_Type == "NPS", 1, GMR),  # Set estimate to 1 for NPS
    GMR_Lower = ifelse(Sample_Type == "NPS", 1, GMR_Lower),   # Set conf.low to 1 for NPS
    GMR_Upper = ifelse(Sample_Type == "NPS", 1, GMR_Upper)  # Set conf.high to 1 for NPS
  )


GMR_Vis1 <- ggplot(GMRs_with_NP, aes(x = Sample_Type, y = GMR, fill = Sample_Type)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_errorbar(aes(ymin = GMR_Lower, ymax = GMR_Upper), 
                position = position_dodge(width = 0.9), 
                width = 0.2, color = "black") +
  scale_fill_manual(values = custom_colors) +
  labs(
    x        = NULL,
    y        = "Geometric Mean Ratio",
    fill     = "Sample Type"
  ) +
  facet_wrap(~Analytes, scales = "free_y") +

  # bump the overall base font size...
  theme_bw(base_size = 25) +

  # ...then fine-tune specific pieces
  theme(
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text(angle = 65, vjust = 1, hjust = 1, size = 24),
    axis.text.y      = element_text(size = 24),
    strip.text       = element_text(size = 25),  # facet labels
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
    legend.position = "bottom"
  )

print(GMR_Vis1)

#Alternatively without the NP:

GMR_Vis2 <- ggplot(GMRs, aes(x = Sample_Type, y = GMR, fill = Sample_Type)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") + # Reference line at GMR=1
  geom_errorbar(aes(ymin = GMR_Lower, ymax = GMR_Upper), 
                position = position_dodge(width = 0.9), 
                width = 0.2, color = "black") +  # Error bar color
  scale_fill_manual(values = custom_colors) +  # Apply custom colors
  labs(
    title = "Geometric Mean Ratios by Sample Type with 95% CI",
    subtitle = "Predicted Data",
    x = "Sample Type",
    y = "Geometric Mean Ratio",
    fill = "Sample Type") +
  facet_wrap(~Analytes, scales = "free_y") + # Ensure consistent y-axis
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 65, vjust = 1, size = 9, hjust = 1)
  )

# Print the updated plot
GMR_Vis2

# ggsave("GMR_LegendPlex_v2.tiff", plot = GMR_Vis2, dpi = 300, width = 10, height = 8, units = "in", device = "tiff", compression = "lzw" )
# 
ggsave("GMR_LegendPlex_v1-new.tiff", plot = GMR_Vis1, dpi = 300, width = 13, height = 11, units = "in", device = "tiff", compression = "lzw" )
```


## I3. GMR Results Table
```{r gmr results, warning=FALSE}
library(gt)
 # Define a significance function using ifelse
 significance <- function(p) {
   ifelse(p < 0.001, "***", 
          ifelse(p < 0.01, "**", 
                 ifelse(p < 0.05, "*", " ")))
 }

 
GMR_For_Table <- GMRs %>% 
  rename(`Sample Type` = "Sample_Type") %>% 
  rename(GMR = "GMR") %>% 
  rename(Analyte = "Analytes") %>% 
  mutate(Significance = significance(`p-value`))  
  
GMR_For_Table <- GMR_For_Table %>% 
  mutate(GMR_CI = paste0(sprintf("%.2f", GMR), "<br>(",
                         sprintf("%.2f", GMR_Lower), ", ", 
                         sprintf("%.2f", GMR_Upper), ") ", 
                         Significance))
# Pivot the data to have Analytes as columns and Sample Type as rows
wide_table_GMR <- GMR_For_Table %>% 
  subset(select=c(Analyte, GMR_CI, `Sample Type`)) %>% 
  pivot_wider(names_from = Analyte, values_from = GMR_CI)

##### Make the table 


# Create the gt table
gt_table <- wide_table_GMR %>%
  gt(rowname_col = "Sample Type") %>%
  tab_header(
    title = md("**GMR and 95% CI: Analytes by Sample Type**"),
    subtitle = "Professionally Collected"
  ) %>%
  fmt_markdown(columns = everything()) %>%  # Ensure <br> is rendered as HTML
  tab_style(
    style = cell_text(weight = "bold"),  # Apply bold text to analyte column headers
    locations = cells_column_labels(columns = everything())
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),  # Apply bold text to sample type row labels
    locations = cells_stub()
  ) %>%
  cols_align("center") %>%  # Center-align all columns
  tab_style(
    style = cell_text(size = "small"),  # Apply smaller text size to the body
    locations = cells_body(columns = everything())
  )


# Show the table
gt_table
#gtsave(gt_table, "GMR_table_final.html")
gtsave(gt_table, "GMR_table_final.docx")


```

