---
title: "16S Data Analysis"
output: 
  html_document:
    toc: true            # Enable Table of Contents
    toc_float: true
    code_folding: "hide"      
date: "22-05-2025"
---

```{r Load Packages, include=FALSE}
# Function to install/load packages:

load_packages <- function(packages) {
  sapply(packages, function(pkg) {
    if (!require(pkg, character.only = TRUE)) {
      install.packages(pkg, dependencies = TRUE)
      library(pkg, character.only = TRUE)
    }
  })
}

packages <- c("purrr", "dplyr", "tidyr", "knitr", "magrittr", "reshape2", 
              "tidyverse", "ggplot2", "RColorBrewer", "stringr",
              "rstatix", "ggpubr", "ggsignif", "FSA",  
              "readxl", "viridis", "patchwork",   
              "ggfortify", "factoextra", "pheatmap",
              "cowplot","coin","psych","cowplot",
              "phyloseq","microbiome","plyr","hrbrthemes",
              "RColorBrewer","readr", "decontam", "microbiomeutilities",
              "phangorn", "tibble","utils")

load_packages(packages)
library(microViz)
library(vegan)
library(gghalves)
```


# Load File

```{r}
ps.nose.og <- readRDS("06-ps_nasal_samples.rds")
ps.oral.og <- readRDS("06-ps_oral_samples.rds")

tax_table(ps.nose.og)[is.na(tax_table(ps.nose.og))] = ""  #Super important!!!
tax_table(ps.oral.og)[is.na(tax_table(ps.oral.og))] = ""  #Super important!!!

```

# Inspect nasal samples
```{r}
print_ps(ps.nose.og)

```
# Inspect oral samples
```{r}
print_ps(ps.oral.og)

```
# Count number of samples per sample type
```{r}
table(sample_data(ps.oral.og)$sample_type)
table(sample_data(ps.nose.og)$sample_type)

```

# Compositions (from split data)
```{r}
ps.nasal <- ps.nose.og
ps.oral <- ps.oral.og
#Genus level
ps.oral.cavity.0.genus <- tax_glom(ps.oral, "Genus", NArm = TRUE)

ps.oral.cavity.rel <- transform_sample_counts(ps.oral.cavity.0.genus, function(OTU) OTU/sum(OTU) * 100) # relative abundance

ps.nasal.cavity.0.genus <- tax_glom(ps.nasal, "Genus", NArm = TRUE)

ps.nasal.cavity.rel <- transform_sample_counts(ps.nasal.cavity.0.genus, function(OTU) OTU/sum(OTU) * 100) # relative abundance

# subset of OPS and saliva samples:
ops<- ps.oral.cavity.rel %>% ps_filter(Sample_Type == "OPS")
sal<- ps.oral.cavity.rel %>% ps_filter(Sample_Type == "Saliva")
```



# Core members
```{r}
# Identify core taxa: detection > 0.1%, present in >50% of samples
core_taxa_ops <- core_members(ops, detection = 0.001, prevalence = 0.5)
core_taxa_sal <- core_members(sal, detection = 0.001, prevalence = 0.5)
core_taxa_oral <- core_members(ps.oral.cavity.rel, detection = 0.001, prevalence = 0.5)
core_taxa_nasal <- core_members(ps.nasal.cavity.rel, detection = 0.001, prevalence = 0.5)

```
# Assign taxonomy
```{r}
# Extract taxonomy table as a data frame
tax <- as.data.frame(tax_table(ps.oral.cavity.rel))
core_taxa_oral_info <- tax[core_taxa_oral, ]
core_taxa_oral_info$ASV <- rownames(core_taxa_oral_info)
# Extract taxonomy table as a data frame
tax_ops <- as.data.frame(tax_table(ops))
tax_sal <- as.data.frame(tax_table(sal))
# Subset to core taxa
core_taxa_ops_info <- tax_ops[core_taxa_ops, ]
core_taxa_sal_info <- tax_sal[core_taxa_sal, ]

# Optional: Add ASV IDs as a column
core_taxa_ops_info$ASV <- rownames(core_taxa_ops_info)
core_taxa_sal_info$ASV <- rownames(core_taxa_sal_info)

tax_nasal <- as.data.frame(tax_table(ps.nasal.cavity.rel))
core_taxa_nasal_info <- tax_nasal[core_taxa_nasal, ]
core_taxa_nasal_info$ASV <- rownames(core_taxa_nasal_info)

```


## Core taxa : Oral Cavity
```{r}
print(core_taxa_oral_info$Genus)
```
## Core taxa : Nasal Cavity
```{r}
print(core_taxa_nasal_info$Genus)
```

# Create custom colors for Genera
```{r}
top_genera_cols <- c(
  Corynebacterium = "#B9DDF1",   
  Streptococcus   = "#325F8C",   
  Staphylococcus  = "#3C8200",   
  Dolosigranulum  = "#C2E393",   
  Lawsonella      = "#EEA9B8",   
  Haemophilus     = "#CD7054",   
  Anaerococcus    = "lightsteelblue1",   
  Moraxella       = "#8B4789",   
  Prevotella_7    = "burlywood3",   
  Neisseria       = "#E2F4B4",    
  Actinomyces     = "#E0CB48",
  Porphyromonas="#99d8c9",
  Fusobacterium="#543005",
  Granulicatella = "#01665e",
  Lautropia = "#a50026",
  Listeria = "#fee08b",
  Peptoniphilus = "#c51b7d",
  Gemella ="#E87687",
  Rothia="#41B7C4",
  Peptoniphilus="#D69054",
  Finegoldia="#CDC673",
  Cutibacterium="#FFE8E8",
  Veillonella ="#D5A7C4",
  Leptotrichia = "turquoise2",
  Other = "#C3C3C3"
)
```

# Top 10 taxa: Oral cavity
```{r, fig.width=14, fig.height=10}
or_tax<- as.data.frame(tax_table(ps.oral.cavity.rel))


oral_cavity_comp_plot <- ps.oral.cavity.rel %>%
  comp_barplot("Genus",  n_taxa = 10) +
  coord_flip()+
  scale_fill_manual(values=top_genera_cols)
oral_cavity_comp_plot


oral_cav_sample_type_plot <- ps.oral.cavity.rel %>%
  comp_barplot("Genus", facet_by = "Sample_Type", n_taxa = 10) +
  coord_flip() +
  scale_fill_manual(values=top_genera_cols)
oral_cav_sample_type_plot
```




# Top 10 taxa: Nose
```{r, fig.width=14, fig.height=10}

nasal_cavity_comp_plot <- ps.nasal.cavity.rel %>%
  comp_barplot("Genus", facet_by = "Sampling_Method", n_taxa = 10) +
  coord_flip()+
  scale_fill_manual(values=top_genera_cols)
nasal_cavity_comp_plot





nasal_cav_sample_type_plot <- ps.nasal.cavity.rel %>%
  comp_barplot("Genus", facet_by = "Sample_Type", n_taxa = 10) +
  coord_flip()+
  scale_fill_manual(values=top_genera_cols)
nasal_cav_sample_type_plot
```

# Merge the two datasets at the Genus level
```{r}
# 1. collapse each to genus, or leave ASV-level:
#    (here I’ll leave them ASV-level)
# 2. prefix the ASV names so they never collide
taxa_names(ps.nasal) <- paste0("Nose__", taxa_names(ps.nasal))
taxa_names(ps.oral) <- paste0("Oral__", taxa_names(ps.oral))

# 3. now merge
ps.both <- merge_phyloseq(ps.nasal, ps.oral)

# 4.  collapse to genus and make relative abundances
ps.both.genus   <- tax_glom(ps.both, "Genus", NArm=TRUE)

ps.both.rel     <- transform_sample_counts(ps.both.genus, function(x) x / sum(x) * 100)


```

# You can save this phyloseq object for use in the later multiomics analysis
```{r}
saveRDS(ps.both, "ps.for.multiomics.rds")
```


# Top 10 taxa
```{r, fig.width=14, fig.height=10}

comp_sampling_method_plot <- ps.both.rel %>%
  comp_barplot("Genus", facet_by = "Sampling_Method", n_taxa = 10) +
  coord_flip()+
  scale_fill_manual(values=top_genera_cols)
comp_sampling_method_plot


comp_sample_type_plot <- ps.both.rel %>%
  comp_barplot("Genus", facet_by = "Sample_Type", n_taxa = 10) +
  coord_flip()+
  scale_fill_manual(values=top_genera_cols)
comp_sample_type_plot

# save plot (optionally)
# ggsave(
#   filename    = "oral_compositions-f.tiff",
#   plot        = oral_cav_sample_type_plot,
#   device      = "tiff",
#   dpi         = 300,
#   width       = 10,      # in inches
#   height      = 8,      # adjust as needed
#   units       = "in",
#   compression = "lzw"
# )

```

We will use only the **professionally collected** samples for the main comparisons
```{r}
#Only Professionally Collected Samples
ps.rel.prof <- ps.both.rel %>% ps_filter(Sampling_Strategy=="Professional")


#Only Nose Professionally Collected Samples
ps.rel.nose <- ps.rel.prof %>% ps_filter(Region=="Nasal Cavity")


#Only OP & Saliva Professionally Collected 
ps.rel.oral <- ps.rel.prof %>% ps_filter(Region=="Oral Cavity")
```


# setup: colors for each sample type
```{r}

pcoa_colors<- c(NPS="#A2D2DF", MTS="bisque3", MLF="#FFBD73", OPS ="#BC7C7C", Saliva="#8ABFA3")

region_colors <- c(`Nasal Cavity` = "#A2D2DF", `Oral Cavity`="#BC7C7C")

```



# Beta-diversity: Regions of the URT
```{r}
library(pairwiseAdonis)

set.seed(202)
#Calculate relative abundances
ps.rel.prof2 <- prune_taxa(taxa_sums(ps.rel.prof)>0, ps.rel.prof)
ps.rel.prof.comp <- microbiome::transform(ps.rel.prof2, "compositional")

ps_dist_matrix <- phyloseq::distance(ps.rel.prof.comp, method ="bray")
sampledf <- data.frame(sample_data(ps.rel.prof.comp))
data <- pairwiseAdonis::pairwise.adonis(ps_dist_matrix, sampledf$Region)
print(data)
#significantly different
```


# Bray-Curtis distance matrix : PCoA visualization of beta diversity
```{r , fig.width=11, fig.height=8}
# Using Bray Curtis distance
# Extract overall p-value or build a summary string, e.g.:
annotation_text <- paste0(
  "Nasal vs Oral Cavity:\n",
  "p = ", format(data$p.adjusted), ", ",  # Format for p-value
  "R² = ", format(data$R2 * 100, digits = 3), "%"                # Format for R²
)

out.bc <- ordinate(ps.rel.prof.comp, method = "PCoA", distance ="bray")

pcoa.bc.0 <- plot_ordination(ps.rel.prof, out.bc, color="Region") +
  geom_point(size = 2)  +
  stat_ellipse(
  aes(color = Region, fill = Region, group = Region),
  type = "norm",
  geom = "polygon",
  alpha = 0.2,
  linetype = 2,
  level = 0.25) +
  theme(legend.position = "bottom")


pcoa.bc.0 <- pcoa.bc.0 + scale_color_manual(name = "Region", values = region_colors)
pcoa.bc.0 <- pcoa.bc.0 + scale_fill_manual(values = region_colors)
pcoa.bc.0 <- pcoa.bc.0  +
  annotate("text", x = Inf, y = -Inf, label = annotation_text, 
           hjust = 1.1, vjust = -0.5, size = 6, fontface = "italic")+
  theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text( size = 22),
    axis.text.y      = element_text(size = 22),
    strip.text       = element_text(size = 25),  # facet labels
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
  )


# view plot:
pcoa.bc.0
# save plot:
ggsave(
  filename    = "PCoA_Niches0.tiff",
  plot        = pcoa.bc.0,
  device      = "tiff",
  dpi         = 300,
  width       = 10,      # in inches
  height      = 8,      # adjust as needed
  units       = "in",
  compression = "lzw")

```


Significant compositional differences (beta-diversity) between the nose and the oral cavity.



# Beta-diversity: All Sample Types
```{r}
set.seed(202)

#Calculate relative abundances
ps.rel.prof2 <- prune_taxa(taxa_sums(ps.rel.prof)>0, ps.rel.prof)
ps.rel.prof.comp <- microbiome::transform(ps.rel.prof2, "compositional")

ps_dist_matrix <- phyloseq::distance(ps.rel.prof.comp, method ="bray")
sampledf <- data.frame(sample_data(ps.rel.prof.comp))
data <- pairwiseAdonis::pairwise.adonis(ps_dist_matrix, sampledf$Sample_Type)
print(data)
```



# Bray-Curtis distance matrix : PCoA visualization of beta diversity (sample type level)
```{r , fig.width=11, fig.height=8}
# Using Bray Curtis distance

annotation_text <- paste("NPS vs MLF: R² =", format(data$R2[1] * 100, digits = 2), "%"," ", data$sig[1],
                         "\nNPS vs MTS: R² =", format(data$R2[2] * 100, digits = 2), "%", " ",data$sig[2],
                         "\nNPS vs OPS: R² =", format(data$R2[3] * 100, digits = 2),  "%"," ",data$sig[3],
                        "\nNPS vs Saliva: R² =", format(data$R2[4] * 100, digits = 2),  "%"," ", data$sig[4],
                        "\nMLF vs MTS: R² =", format(data$R2[5] * 100, digits = 2),  "%", " ",data$sig[5],
                        "\nMLF vs OPS: R² =", format(data$R2[6] * 100, digits = 2),  "%", " ",data$sig[6],
                        "\nMLF vs Saliva: R² =", format(data$R2[7] * 100, digits = 2),  "%"," ", data$sig[7],
                        "\nMTS vs OPS: R² =", format(data$R2[8] * 100, digits = 2),  "%", " ",data$sig[8],
                        "\nMTS vs Saliva: R² =", format(data$R2[9] * 100, digits = 2),  "%", " ",data$sig[9],
                        "\nOPS vs Saliva: R² =",format(data$R2[10] * 100, digits = 2),  "%", "",data$sig[10])

out.bc <- ordinate(ps.rel.prof.comp, method = "PCoA", distance ="bray")

pcoa.bc.1 <- # build the PCoA plot
pcoa.bc.1 <- plot_ordination(ps.rel.prof, out.bc, color = "Sample_Type") +
  geom_point(size = 3) +
  stat_ellipse(aes(fill = Sample_Type, color = Sample_Type, group = Sample_Type),
               geom = "polygon", alpha = 0.2, linetype = 2, level = 0.25) +

  # scales
  scale_color_manual(name = "Sample Type", values = pcoa_colors) +
  scale_fill_manual(name  = "Sample Type", values = pcoa_colors) +

    theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text( size = 22),
    axis.text.y      = element_text(size = 22),
    strip.text       = element_text(size = 25),  # facet labels
    legend.position =     "none",
  ) +

  # your multi‐line annotation in the corner
  annotate("text",
           x      = Inf, y      = -Inf,
           label  = annotation_text,
           hjust  = 1.1, vjust  = -0.05,
           size   = 6,         # larger than the default 4
           fontface = "italic"
  )

# view plot
pcoa.bc.1
#save plot
ggsave(
  filename    = "PCoA_Niches.tiff",
  plot        = pcoa.bc.1,
  device      = "tiff",
  dpi         = 300,
  width       = 13,      # in inches
  height      = 9,      # adjust as needed
  units       = "in",
  compression = "lzw")
```







# Beta-diversity: Nose Sample Types
```{r}
set.seed(202)

#Calculate relative abundances
ps.rel.nose2 <- prune_taxa(taxa_sums(ps.rel.nose)>0, ps.rel.nose)
ps.rel.nose.comp <- microbiome::transform(ps.rel.nose2, "compositional")

ps_dist_matrix <- phyloseq::distance(ps.rel.nose.comp, method ="bray")
sampledf <- data.frame(sample_data(ps.rel.nose.comp))
data <- pairwiseAdonis::pairwise.adonis(ps_dist_matrix, sampledf$Sample_Type)
print(data)
```



# Bray-Curtis distance matrix: PCoA visualization on nasal samples alone
```{r , fig.width=11, fig.height=8}
# Using Bray Curtis distance
# Extract overall p-value or build a summary string, e.g.:
annotation_text <- paste("NPS vs MLF: p =", data$p.adjusted[1],
                        "\nNPS vs MTS: p =", data$p.adjusted[2],
                        "\nMLF vs MTS: p =", data$p.adjusted[3])

out.bc <- ordinate(ps.rel.nose.comp, method = "PCoA", distance ="bray")

pcoa.bc.2 <- plot_ordination(ps.rel.nose, out.bc, color="Sample_Type") +
  geom_point(size = 2)  +
  stat_ellipse(
  aes(color = Sample_Type, fill = Sample_Type, group = Sample_Type),
  type = "norm",
  geom = "polygon",
  alpha = 0.2,
  linetype = 2,
  level = 0.25) +
  theme(legend.position = "bottom")


pcoa.bc.2 <- pcoa.bc.2 + scale_color_manual(name = "Sample_Type", values = pcoa_colors)
pcoa.bc.2 <- pcoa.bc.2 + scale_fill_manual(values = pcoa_colors)
pcoa.bc.2 <- pcoa.bc.2+theme_bw()
pcoa.bc.2 <- pcoa.bc.2  +
  annotate("text", x = -Inf, y = Inf, label = annotation_text, 
           hjust = -0.1, vjust = 1.05, size = 4, fontface = "italic")


# view plot
pcoa.bc.2
# save plot
ggsave(
  filename    = "PCoA_NasalC.tiff",
  plot        = pcoa.bc.2,
  device      = "tiff",
  dpi         = 300,
  width       = 10,      # in inches
  height      = 8,      # adjust as needed
  units       = "in",
  compression = "lzw")
```


# Beta-diversity: OPS, Saliva Sample Types
```{r}
set.seed(202)

#Calculate relative abundances
ps.rel.oral2 <- prune_taxa(taxa_sums(ps.rel.oral)>0, ps.rel.oral)
ps.rel.oral.comp <- microbiome::transform(ps.rel.oral2, "compositional")

ps_dist_matrix <- phyloseq::distance(ps.rel.oral.comp, method ="bray")
sampledf <- data.frame(sample_data(ps.rel.oral.comp))
data <- pairwiseAdonis::pairwise.adonis(ps_dist_matrix, sampledf$Sample_Type)
print(data)
```



# Bray-Curtis distance matrix: PCoA on oral cavity samples
```{r , fig.width=11, fig.height=8}
# Using Bray Curtis distance
# Extract overall p-value or build a summary string, e.g.:
annotation_text <- paste0("OPS vs Saliva: R² = ", format(data$R2 * 100, digits = 3), "%", 
                          " p = ", data$p.adjusted,"," )

out.bc <- ordinate(ps.rel.oral.comp, method = "PCoA", distance ="bray")

pcoa.bc.3 <- plot_ordination(ps.rel.oral, out.bc, color="Sample_Type") +
  geom_point(size = 2)  +
  stat_ellipse(
  aes(color = Sample_Type, fill = Sample_Type, group = Sample_Type),
  type = "norm",
  geom = "polygon",
  alpha = 0.2,
  linetype = 2,
  level = 0.25) +
  theme(legend.position = "bottom")


pcoa.bc.3 <- pcoa.bc.3 + scale_color_manual(name = "Sample_Type", values = pcoa_colors)
pcoa.bc.3 <- pcoa.bc.3 + scale_fill_manual(values = pcoa_colors)
pcoa.bc.3 <- pcoa.bc.3+  theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text( size = 22),
    axis.text.y      = element_text(size = 22),
    strip.text       = element_text(size = 25),  # facet labels
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
  )
pcoa.bc.3 <- pcoa.bc.3  +
  annotate("text", x = Inf, y = -Inf, label = annotation_text, 
           hjust = 1.1, vjust = -0.5, size = 6, fontface = "italic")


# view plot
pcoa.bc.3
# save plot
ggsave(
  filename    = "PCoA_OralC.tiff",
  plot        = pcoa.bc.3,
  device      = "tiff",
  dpi         = 300,
  width       = 13,      # in inches
  height      = 11,      # adjust as needed
  units       = "in",
  compression = "lzw")
```

Significant compositional differences (beta-diversity) between OPS and Saliva samples.

# Here, we explore differences between the left and the right nostril samples (MLF, MTS)
```{r}
ps.rel.mlf <- ps.rel.nose %>% ps_filter(Sample_Type == "MLF")

#Only Nose Professionally Collected Samples
ps.rel.mts <- ps.rel.nose %>% ps_filter(Sample_Type == "MTS")


nostril_colors <- c(Left="plum3",Right="slategray3")
```


# Beta-diversity: MLF (Left vs Right Nostril) 
```{r}
set.seed(202)

#Calculate relative abundances
ps.rel.mlf2 <- prune_taxa(taxa_sums(ps.rel.mlf)>0, ps.rel.mlf)
ps.rel.mlf.comp <- microbiome::transform(ps.rel.mlf2, "compositional")

ps_dist_matrix <- phyloseq::distance(ps.rel.mlf.comp, method ="bray")
sampledf <- data.frame(sample_data(ps.rel.mlf.comp))
data <- pairwiseAdonis::pairwise.adonis(ps_dist_matrix, sampledf$Nostril)
print(data)
```



# PCoA Bray-Curtis distance matrix: Nostrils (MLF)
```{r , fig.width=11, fig.height=8}
# Using Bray Curtis distance
# Extract overall p-value or build a summary string, e.g.:
annotation_text <- paste("Left vs Right: R² =", format(data$R2 * 100, digits = 2),"%" ,", p = ",data$p.adjusted)

out.bc <- ordinate(ps.rel.mlf.comp, method = "PCoA", distance ="bray")

pcoa.bc.4 <- plot_ordination(ps.rel.mlf, out.bc, color="Nostril") +
  geom_point(size = 2)  +
  stat_ellipse(
  aes(color = Nostril, fill = Nostril, group = Nostril),
  type = "norm",
  geom = "polygon",
  alpha = 0.2,
  linetype = 2,
  level = 0.25) +
  theme(legend.position = "bottom")


pcoa.bc.4 <- pcoa.bc.4 + scale_color_manual(name = "Nostril", values = nostril_colors)
pcoa.bc.4 <- pcoa.bc.4 + scale_fill_manual(values = nostril_colors)
pcoa.bc.4 <- pcoa.bc.4+  theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text( size = 22),
    axis.text.y      = element_text(size = 22),
    strip.text       = element_text(size = 25),  # facet labels
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
  )
pcoa.bc.4 <- pcoa.bc.4  +
  annotate("text", x = -Inf, y = Inf, label = annotation_text, 
           hjust = -0.1, vjust = 1.3, size = 6, fontface = "italic")


# view plot
pcoa.bc.4
# save plot
ggsave(
  filename    = "PCoA_MLF_Nostrils.tiff",
  plot        = pcoa.bc.4,
  device      = "tiff",
  dpi         = 300,
  width       = 10,      # in inches
  height      = 8,      # adjust as needed
  units       = "in",
  compression = "lzw")
```


# Beta-diversity: MTS (Left vs Right Nostril) 
```{r}
set.seed(202)

#Calculate relative abundances
ps.rel.mts2 <- prune_taxa(taxa_sums(ps.rel.mts)>0, ps.rel.mts)
ps.rel.mts.comp <- microbiome::transform(ps.rel.mts2, "compositional")

ps_dist_matrix <- phyloseq::distance(ps.rel.mts.comp, method ="bray")
sampledf <- data.frame(sample_data(ps.rel.mts.comp))
data <- pairwiseAdonis::pairwise.adonis(ps_dist_matrix, sampledf$Nostril)
print(data)
```



# PCoA Bray-Curtis distance matrix: Nostrils (MTS)
```{r , fig.width=11, fig.height=8}
# Using Bray Curtis distance
# Extract overall p-value or build a summary string, e.g.:
annotation_text <- paste("Left vs Right: R² =", format(data$R2 * 100, digits = 2),"%" ,", p = ",data$p.adjusted)

out.bc <- ordinate(ps.rel.mts.comp, method = "PCoA", distance ="bray")

pcoa.bc.5 <- plot_ordination(ps.rel.mts, out.bc, color="Nostril") +
  geom_point(size = 2)  +
  stat_ellipse(
  aes(color = Nostril, fill = Nostril, group = Nostril),
  type = "norm",
  geom = "polygon",
  alpha = 0.2,
  linetype = 2,
  level = 0.25) +
  theme(legend.position = "bottom")


pcoa.bc.5 <- pcoa.bc.5 + scale_color_manual(name = "Nostril", values = nostril_colors)
pcoa.bc.5 <- pcoa.bc.5 + scale_fill_manual(values = nostril_colors)
pcoa.bc.5 <- pcoa.bc.5 + theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text( size = 22),
    axis.text.y      = element_text(size = 22),
    strip.text       = element_text(size = 25),  # facet labels
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
  )
pcoa.bc.5 <- pcoa.bc.5  +
  annotate("text", x = -Inf, y = Inf, label = annotation_text, 
           hjust = -0.1, vjust = 1.3, size = 6, fontface = "italic")


# view plot
pcoa.bc.5
# save plot
ggsave(
  filename    = "PCoA_MTS_Nostrils.tiff",
  plot        = pcoa.bc.5,
  device      = "tiff",
  dpi         = 300,
  width       = 10,      # in inches
  height      = 8,      # adjust as needed
  units       = "in",
  compression = "lzw")
```



No significant compositional differences (beta-diversity) between the left and the right nostrils.



# Beta-diversity: self vs professionally collected samples (per region)
This was only performed for MTS, MLF and saliva samples
```{r}
ps.mlf.str.rel <- ps.both.rel %>% ps_filter(Sample_Type == "MLF")
ps.mts.str.rel <- ps.both.rel %>% ps_filter(Sample_Type == "MTS")
ps.saliva.str.rel <- ps.both.rel %>% ps_filter(Sample_Type == "Saliva")

sampling_strategy_colors <- c(Professional="#2F4F4F", Self ="#CD9B1D")
```


# MLF 
```{r}
set.seed(202)

#Calculate relative abundances
ps.mlf.str.2 <- prune_taxa(taxa_sums(ps.mlf.str.rel)>0, ps.mlf.str.rel)
ps.mlf.str.2.comp <- microbiome::transform(ps.mlf.str.2, "compositional")

ps_dist_matrix <- phyloseq::distance(ps.mlf.str.2.comp, method ="bray")
sampledf <- data.frame(sample_data(ps.mlf.str.2.comp))
data <- pairwiseAdonis::pairwise.adonis(ps_dist_matrix, sampledf$Sampling_Strategy)
print(data)
```



# PCoA Bray-Curtis distance matrix: MLF (professional vs self collection)
```{r, fig.width=11, fig.height=8}
# Using Bray Curtis distance
# Extract overall p-value or build a summary string, e.g.:
annotation_text <- paste0("Professional vs Self:R² = ", format(data$R2 * 100, digits = 2), "%", ",",
                          " p = ", round(data$p.adjusted, 3))

out.bc <- ordinate(ps.mlf.str.2.comp, method = "PCoA", distance ="bray")

pcoa.bc.6 <- plot_ordination(ps.mlf.str.rel, out.bc, color="Sampling_Strategy") +
  geom_point(size = 2)  +
  stat_ellipse(
  aes(color = Sampling_Strategy, fill = Sampling_Strategy, group = Sampling_Strategy),
  type = "norm",
  geom = "polygon",
  alpha = 0.2,
  linetype = 2,
  level = 0.25) +
  theme(legend.position = "bottom")


pcoa.bc.6 <- pcoa.bc.6 + scale_color_manual(name = "Sampling_Strategy", values = sampling_strategy_colors)
pcoa.bc.6 <- pcoa.bc.6 + scale_fill_manual(values = sampling_strategy_colors)
pcoa.bc.6 <- pcoa.bc.6+  theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text( size = 22),
    axis.text.y      = element_text(size = 22),
    strip.text       = element_text(size = 25),  # facet labels
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
  )
pcoa.bc.6 <- pcoa.bc.6  +
  annotate("text", x = Inf, y = -Inf, label = annotation_text, 
           hjust = 1.1, vjust = -0.5, size = 5, fontface = "italic")



pcoa.bc.6

ggsave(
  filename    = "PCoA_MLF_SampStrategy.tiff",
  plot        = pcoa.bc.6,
  device      = "tiff",
  dpi         = 300,
  width       = 13,      # in inches
  height      = 11,      # adjust as needed
  units       = "in",
  compression = "lzw")
```




# MTS
```{r}
set.seed(202)

#Calculate relative abundances
ps.mts.str.2 <- prune_taxa(taxa_sums(ps.mts.str.rel)>0, ps.mts.str.rel)
ps.mts.str.2.comp <- microbiome::transform(ps.mts.str.2, "compositional")

ps_dist_matrix <- phyloseq::distance(ps.mts.str.2.comp, method ="bray")
sampledf <- data.frame(sample_data(ps.mts.str.2.comp))
data <- pairwiseAdonis::pairwise.adonis(ps_dist_matrix, sampledf$Sampling_Strategy)
print(data)
```


# PCoA Bray-Curtis distance matrix: MTS (professional vs self collection)
```{r , fig.width=11, fig.height=8}
# Using Bray Curtis distance
# Extract overall p-value or build a summary string, e.g.:
annotation_text <- paste0("Professional vs Self:R² = ", format(data$R2 * 100, digits = 2), "%", ",",
                          " p = ", round(data$p.adjusted, 3))

out.bc <- ordinate(ps.mts.str.2.comp, method = "PCoA", distance ="bray")

pcoa.bc.7 <- plot_ordination(ps.mts.str.rel, out.bc, color="Sampling_Strategy") +
  geom_point(size = 2)  +
  stat_ellipse(
  aes(color = Sampling_Strategy, fill = Sampling_Strategy, group = Sampling_Strategy),
  type = "norm",
  geom = "polygon",
  alpha = 0.2,
  linetype = 2,
  level = 0.25) +
  theme(legend.position = "bottom")


pcoa.bc.7 <- pcoa.bc.7 + scale_color_manual(name = "Sampling_Strategy", values = sampling_strategy_colors)
pcoa.bc.7 <- pcoa.bc.7 + scale_fill_manual(values = sampling_strategy_colors)
pcoa.bc.7 <- pcoa.bc.7+  theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text( size = 22),
    axis.text.y      = element_text(size = 22),
    strip.text       = element_text(size = 25),  # facet labels
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
  )
pcoa.bc.7 <- pcoa.bc.7  +
  annotate("text", x = Inf, y = -Inf, label = annotation_text, 
           hjust = 1.1, vjust = -0.5, size = 5, fontface = "italic")



pcoa.bc.7

ggsave(
  filename    = "PCoA_MTS_SampStrategy.tiff",
  plot        = pcoa.bc.7,
  device      = "tiff",
  dpi         = 300,
  width       = 13,      # in inches
  height      = 11,      # adjust as needed
  units       = "in",
  compression = "lzw")
```





# Saliva 
```{r}
set.seed(202)
#Calculate relative abundances
ps.saliva.str.2 <- prune_taxa(taxa_sums(ps.saliva.str.rel)>0, ps.saliva.str.rel)
ps.saliva.str.2.comp <- microbiome::transform(ps.saliva.str.2, "compositional")

ps_dist_matrix <- phyloseq::distance(ps.saliva.str.2.comp, method ="bray")
sampledf <- data.frame(sample_data(ps.saliva.str.2.comp))
data <- pairwiseAdonis::pairwise.adonis(ps_dist_matrix, sampledf$Sampling_Strategy)
print(data)
```

# PCoA Bray-Curtis distance matrix: Saliva (professional vs self collection)
```{r , fig.width=11, fig.height=8}
# Using Bray Curtis distance
# Extract overall p-value or build a summary string, e.g.:
annotation_text <- paste0("Professional vs Self:R² = ", format(data$R2 * 100, digits = 2), "%", ",",
                          " p = ", round(data$p.adjusted, 3))

out.bc <- ordinate(ps.saliva.str.2.comp, method = "PCoA", distance ="bray")

pcoa.bc.8 <- plot_ordination(ps.saliva.str.rel, out.bc, color="Sampling_Strategy") +
  geom_point(size = 2)  +
  stat_ellipse(
  aes(color = Sampling_Strategy, fill = Sampling_Strategy, group = Sampling_Strategy),
  type = "norm",
  geom = "polygon",
  alpha = 0.2,
  linetype = 2,
  level = 0.25) +
  theme(legend.position = "bottom")


pcoa.bc.8 <- pcoa.bc.8 + scale_color_manual(name = "Sampling_Strategy", values = sampling_strategy_colors)
pcoa.bc.8 <- pcoa.bc.8 + scale_fill_manual(values = sampling_strategy_colors)
pcoa.bc.8 <- pcoa.bc.8+  theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text( size = 22),
    axis.text.y      = element_text(size = 22),
    strip.text       = element_text(size = 25),  # facet labels
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
  )
pcoa.bc.8 <- pcoa.bc.8  +
  annotate("text", x = Inf, y = -Inf, label = annotation_text, 
           hjust = 1.1, vjust = -0.5, size = 5, fontface = "italic")



pcoa.bc.8

ggsave(
  filename    = "PCoA_Saliva_SampStrategy.tiff",
  plot        = pcoa.bc.8,
  device      = "tiff",
  dpi         = 300,
  width       = 13,      # in inches
  height      = 11,      # adjust as needed
  units       = "in",
  compression = "lzw")
```



Significant compositional differences (beta-diversity) between self and professionally collected samples only for saliva samples.



# Alpha Diversity:  MLF  - Nostrils
```{r}
set.seed(202)

# First we use plot_richness to calculate alpha diversity values
#Calculate relative abundances
ps.rel.mlf2 <- prune_taxa(taxa_sums(ps.rel.mlf)>0, ps.rel.mlf)
ps.rel.mlf.comp <- microbiome::transform(ps.rel.mlf2, "compositional")

p <- plot_richness(ps.rel.mlf2, x="Sample_Type", color="Sample_Type", measures=c("Shannon")) 

levels((p$data)$variable)

# Subset the data values ofr the Shannon indices
data <- subset((p$data), variable== "Shannon")

ggdensity(data$value, 
          main = "Density plot of Shannon diversity",
          xlab = "diversity")

# or Q-Q plot
ggqqplot(data$value)

# Visual inspection, described in the previous section, is usually unreliable. It's possible to use a significance test comparing the sample distribution to a normal one in order to ascertain whether data show or not a serious deviation from normality.

# Shapiro-Wilk's method is widely recommended for normality test and it provides better power than K-S. It is based on the correlation between the data and the corresponding normal scores.
shapiro.test(data$value) #Shannon
#W = 0.98975, p-value = 0.9154

#   follows normal distribution

```



```{r Shannon boxplot-mlf}
library(gghalves)
mlf_df <- sample_data(ps.rel.mlf2)
# Convert to data frame if needed
mlf_df <- as.data.frame(sample_data(ps.rel.mlf2))

# Identify samples from each nostril
left_samples <- rownames(mlf_df[mlf_df$Nostril == "Left", ])
right_samples <- rownames(mlf_df[mlf_df$Nostril == "Right", ])

# Randomly sample 33 Right nostril samples to match the Left
set.seed(123)  # For reproducibility
right_samples_sub <- sample(right_samples, length(left_samples))

# Combine Left samples with the downsampled Right samples
balanced_samples <- c(left_samples, right_samples_sub)

# Subset the phyloseq object to retain only balanced samples
ps.mlf.balanced <- prune_samples(balanced_samples, ps.rel.mlf2)

# Check the new distribution
table(sample_data(ps.mlf.balanced)$Nostril)

# Check the first few values of the OTU table

# comparisons
my_comparisons <- list( c("Left", "Right"))

pal <- c(Left="plum3",Right="slategray3")
theme_set(theme_bw())

# only significant comparison
#Shannon
p1 <- plot_richness(ps.mlf.balanced, x="Nostril",color="Nostril", measures="Shannon") 
p1 <- p1 + 
  geom_point(aes(color = Nostril),
             position = position_jitter(width = 0.15),
             size = 2, alpha = 0.25) +
  geom_boxplot(fill = pal, width = 0.2, outlier.shape = NA,
               alpha = 0.5) + 
   geom_half_violin(position = position_nudge(x = 0.15, y = 0),
                   alpha = 0.25, side = "r") +
  geom_hline(yintercept = mean((p1$data)$value), linetype = 2, color= "grey") +
  guides(colour = guide_legend(nrow = 1)) +
  scale_colour_manual(values = pal) +
  scale_fill_manual(values = pal) +
  stat_compare_means(
    comparisons = list(c("Left", "Right")),
    aes(label = ..p.signif..), 
    method = "t.test",
    paired = TRUE,
    p.adjust.method = "fdr",
    label.y = 3.6, 
    label.x.npc = 0.2,
    tip.length = 0.01)+
  theme_biome_utils() + xlab("") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
p1 <- p1 + scale_fill_manual("Nostril", values = (pal))+  theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text( size = 22),
    axis.text.y      = element_text(size = 22),
    strip.text       = element_text(size = 25),  # facet labels
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
  )
p1



#for the observed
ps.mlf.prof.obs <- ps.nasal %>% ps_filter(Sample_Type =="MLF")
table(as.data.frame(sample_data(ps.mlf.prof.obs))$Nostril)
#We need to filter out the extra timepoints
ps.mlf.prof.obs <- ps.mlf.prof.obs
# Convert to data frame if needed
mlf_df <- as.data.frame(sample_data(ps.mlf.prof.obs))

# Identify samples from each nostril
left_samples <- rownames(mlf_df[mlf_df$Nostril == "Left", ])
right_samples <- rownames(mlf_df[mlf_df$Nostril == "Right", ])

# Randomly sample 33 Right nostril samples to match the Left
set.seed(123)  # For reproducibility
right_samples_sub <- sample(right_samples, length(left_samples))

# Combine Left samples with the downsampled Right samples
balanced_samples <- c(left_samples, right_samples_sub)

# Subset the phyloseq object to retain only balanced samples
ps.mlf.balanced.obs <- prune_samples(balanced_samples, ps.mlf.prof.obs)

# Check the new distribution
table(sample_data(ps.mlf.balanced.obs)$Nostril)

# Check the current "Nostril" sample sizes after pruning
nostril_table <- table(sample_data(ps.mlf.balanced.obs)$Nostril)
left_count <- nostril_table["Left"]
right_count <- nostril_table["Right"]

# Trim additional "Right" samples if necessary
if (right_count > left_count) {
  # Randomly remove excess "Right" nostril samples
  excess_right_samples <- setdiff(right_samples_sub, left_samples)
  trimmed_right_samples <- sample(excess_right_samples, size = left_count)
  
  # Create a final balanced list
  final_balanced_samples <- c(left_samples, trimmed_right_samples)
  
  # Prune the phyloseq object again
  ps.mlf.balanced.obs <- prune_samples(final_balanced_samples, ps.mlf.prof.obs)
}

# Verify balance again
table(sample_data(ps.mlf.balanced.obs)$Nostril)  # Should now be perfectly balanced
#Observed
p2 <- plot_richness(ps.mlf.balanced.obs, x="Nostril", color="Nostril", measures="Chao1")
p2 <- p2 +
  geom_point(aes(color = Nostril),
             position = position_jitter(width = 0.15),
             size = 2, alpha = 0.25) +
  geom_boxplot(fill = pal , width = 0.2, outlier.shape = NA,
               alpha = 0.5) +
   geom_half_violin(position = position_nudge(x = 0.15, y = 0),
                   alpha = 0.25, side = "r") +
  geom_hline(yintercept = mean((p2$data)$value), linetype = 2, color= "grey") +
  scale_colour_manual(values = pal) +
  scale_fill_manual(values = pal) +
  stat_compare_means(comparisons = list(c("Left", "Right")),
                     aes(label = ..p.signif..),
                     method = "t.test",
                     paired = TRUE,
                     p.adjust.method = "fdr",
                     label.y = 415,
                     label.x.npc = 0.2, tip.length = 0.01)+
  theme_biome_utils() + xlab("") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

p2 <- p2 + scale_colour_manual("Nostril", values = (pal))
p2

#Simpson
p3 <- plot_richness(ps.mlf.balanced, x="Nostril", color="Nostril", measures="Simpson") 
p3 <- p3 + 
  geom_point(aes(color = Nostril),
             position = position_jitter(width = 0.15),
             size = 2, alpha = 0.25) +
  geom_boxplot(fill = pal , width = 0.2, outlier.shape = NA,
               alpha = 0.5) + 
   geom_half_violin(position = position_nudge(x = 0.15, y = 0),
                   alpha = 0.25, side = "r") +
  geom_hline(yintercept = mean((p3$data)$value), linetype = 2, color= "grey") +
  scale_colour_manual(values = pal) +
  scale_fill_manual(values = pal) +
  stat_compare_means(comparisons = list(c("Left", "Right")),
                     aes(label = ..p.signif..), 
                     method = "t.test",
                     p.adjust.method = "fdr",
                     label.y = c(1),
                     paired=TRUE,
                     label.x.npc = 0.2, tip.length = 0.01)+
  theme_biome_utils() + xlab("") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

p3 <- p3 + scale_colour_manual("Nostril", values = (pal))+  theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text( size = 22),
    axis.text.y      = element_text(size = 22),
    strip.text       = element_text(size = 25),  # facet labels
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
  )
p3


figure_legend <- get_legend(p1)

figure_mlf <- ggarrange(p1, p3,
                    labels = c(" "),
                    ncol = 2, nrow = 1,
                    common.legend = TRUE, legend = "right", legend.grob = figure_legend)

figure_mlf

ggsave(
  filename    = "Alpha_div_MLF_nostrils.tiff",
  plot        = figure_mlf,
  device      = "tiff",
  dpi         = 300,
  width       = 13,      # in inches
  height      = 11,      # adjust as needed
  units       = "in",
  compression = "lzw")


```


# Alpha Diversity:  MTS - Nostrils
```{r}
set.seed(202)

# First we use plot_richness to calculate alpha diversity values
#Calculate relative abundances
ps.rel.mts2 <- prune_taxa(taxa_sums(ps.rel.mts)>0, ps.rel.mts)
ps.rel.mts.comp <- microbiome::transform(ps.rel.mts2, "compositional")


p <- plot_richness(ps.rel.mts2, x="Sample_Type", color="Sample_Type", measures=c("Shannon")) 

levels((p$data)$variable)

# Subset the data values ofr the Shannon indices
data <- subset((p$data), variable== "Shannon")

ggdensity(data$value, 
          main = "Density plot of Shannon diversity",
          xlab = "diversity")

# or Q-Q plot
ggqqplot(data$value)

# Visual inspection, described in the previous section, is usually unreliable. It's possible to use a significance test comparing the sample distribution to a normal one in order to ascertain whether data show or not a serious deviation from normality.

# Shapiro-Wilk's method is widely recommended for normality test and it provides better power than K-S. It is based on the correlation between the data and the corresponding normal scores.
shapiro.test(data$value) #Shannon
#W = 0.96762, p-value = 0.04671

#  does not follow normal distribution

```



```{r Shannon boxplot mts}
set.seed(202)

#We need to filter out the extra timepoints
ps.mts.proff <- ps.rel.mts2

ps.mts.prof.df <- sample_data(ps.mts.proff)
# Identify samples from each nostril
left_samples <- rownames(ps.mts.prof.df[ps.mts.prof.df$Nostril == "Left", ])
right_samples <- rownames(ps.mts.prof.df[ps.mts.prof.df$Nostril == "Right", ])

# Randomly sample 33 Right nostril samples to match the Left
set.seed(123)  # For reproducibility
right_samples_sub <- sample(left_samples, length(right_samples))

# Combine Left samples with the downsampled Right samples
balanced_samples <- c(right_samples, right_samples_sub)

# Subset the phyloseq object to retain only balanced samples
ps.mts.balanced <- prune_samples(balanced_samples, ps.mts.proff)

# Check the new distribution
table(sample_data(ps.mts.balanced)$Nostril)

# comparisons
#my_comparisons <- list( c("H_Control", "H_ILI"), c("B_Control", "B_ILI"), c("V_Control", "V_ILI"))
my_comparisons <- list(c("Left", "Right"))

pal <- c(Left="plum3",Right="slategray3")
theme_set(theme_bw())

# only significant comparison
#my_comparisons <- list( c("Control", "B"), c("H", "B"))
#Shannon
p1 <- plot_richness(ps.mts.balanced, x="Nostril",color="Nostril", measures="Shannon") 
p1 <- p1 + 
  geom_point(aes(color = Nostril),
             position = position_jitter(width = 0.15),
             size = 2, alpha = 0.25) +
  geom_boxplot(fill = pal, width = 0.2, outlier.shape = NA,
               alpha = 0.5) + 
   geom_half_violin(position = position_nudge(x = 0.15, y = 0),
                   alpha = 0.25, side = "r") +
  geom_hline(yintercept = median((p1$data)$value), linetype = 2, color= "grey") +
  guides(colour = guide_legend(nrow = 1)) +
  scale_colour_manual(values = pal) +
  scale_fill_manual(values = pal) +
  stat_compare_means(
    comparisons = list(c("Left", "Right")),
    aes(label = ..p.signif..), 
    method = "wilcox.test",
    paired = TRUE,
    p.adjust.method = "fdr",
    label.y = 3.4, 
    label.x.npc = 0.2,
    tip.length = 0.01)+
  theme_biome_utils() + xlab("") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
p1 <- p1 + scale_fill_manual("Nostril", values = (pal))+  theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text( size = 22),
    axis.text.y      = element_text(size = 22),
    strip.text       = element_text(size = 25),  # facet labels
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
  )
p1

#for the observed
ps.mts.prof.obs <- ps.nasal %>% ps_filter(Sample_Type =="MTS")
table(as.data.frame(sample_data(ps.mts.prof.obs))$Nostril)


# Convert to data frame if needed
mts_df <- as.data.frame(sample_data(ps.mts.prof.obs))

# Identify samples from each nostril
left_samples <- rownames(mts_df[mts_df$Nostril == "Left", ])
right_samples <- rownames(mts_df[mts_df$Nostril == "Right", ])

# Randomly sample 33 Right nostril samples to match the Left
set.seed(123)  # For reproducibility
right_samples_sub <- sample(right_samples, length(left_samples))

# Combine Left samples with the downsampled Right samples
balanced_samples <- c(left_samples, right_samples_sub)

# Subset the phyloseq object to retain only balanced samples
ps.mts.balanced.obs <- prune_samples(balanced_samples, ps.mts.prof.obs)

# Check the new distribution
table(sample_data(ps.mts.balanced.obs)$Nostril)
#Observed
p2 <- plot_richness(ps.mts.balanced.obs, x="Nostril", color="Nostril", measures="Chao1")
p2 <- p2 +
  geom_point(aes(color = Nostril),
             position = position_jitter(width = 0.15),
             size = 2, alpha = 0.25) +
  geom_boxplot( width = 0.2, outlier.shape = NA,
               alpha = 0.5) +
   geom_half_violin(position = position_nudge(x = 0.15, y = 0),
                   alpha = 0.25, side = "r") +
  geom_hline(yintercept = median((p2$data)$value), linetype = 2, color= "grey") +
  scale_colour_manual(values = pal) +
  scale_fill_manual(values = pal) +
  stat_compare_means(comparisons = list(c("Left", "Right")),
                     aes(label = ..p.signif..),
                     method = "wilcox.test",p.adjust.method = "fdr",
                     paired = TRUE,
                     label.y = 255,
                     label.x.npc = 0.2, tip.length = 0.01)+
  theme_biome_utils() + xlab("") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

p2 <- p2 + scale_colour_manual("Nostril", values = (pal))
p2

#Simpson
p3 <- plot_richness(ps.mts.balanced, x="Nostril", color="Nostril", measures="Simpson") 
p3 <- p3 + 
  geom_point(aes(color = Nostril),
             position = position_jitter(width = 0.15),
             size = 2, alpha = 0.25) +
  geom_boxplot(fill = pal , width = 0.2, outlier.shape = NA,
               alpha = 0.5) + 
   geom_half_violin(position = position_nudge(x = 0.15, y = 0),
                   alpha = 0.25, side = "r") +
  geom_hline(yintercept = median((p3$data)$value), linetype = 2, color= "grey") +
  scale_colour_manual(values = pal) +
  scale_fill_manual(values = pal) +
  stat_compare_means(comparisons = list(c("Left", "Right")),
                     aes(label = ..p.signif..), 
                     method = "wilcox.test",p.adjust.method = "fdr",
                     label.y = c(1.1),
                     paired=TRUE,
                     label.x.npc = 0.2, tip.length = 0.01)+
  theme_biome_utils() + xlab("") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

p3 <- p3 + scale_colour_manual("Nostril", values = (pal))+  theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text( size = 22),
    axis.text.y      = element_text(size = 22),
    strip.text       = element_text(size = 25),  # facet labels
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
  )
p3


figure_legend <- get_legend(p1)

figure_mts <- ggarrange(p1, p3,
                    labels = c(" "),
                    ncol = 2, nrow = 1,
                    common.legend = TRUE, legend = "right", legend.grob = figure_legend)

figure_mts


ggsave(
  filename    = "Alpha_div_MTS_nostrils.tiff",
  plot        = figure_mts,
  device      = "tiff",
  dpi         = 300,
  width       = 13,      # in inches
  height      = 11,      # adjust as needed
  units       = "in",
  compression = "lzw")
```




# Alpha Diversity: All Nose Sample Types
```{r}
set.seed(202)

ps.rel.nose2 <- prune_taxa(taxa_sums(ps.rel.nose)>0, ps.rel.nose)
ps.rel.nose.comp <- microbiome::transform(ps.rel.nose2, "compositional")
# First we use plot_richness to calculate alpha diversity values
p <- plot_richness(ps.rel.nose2, x="Sample_Type", color="Sample_Type", measures=c("Shannon")) 

levels((p$data)$variable)

# Subset the data values ofr the Shannon indices
data <- subset((p$data), variable== "Shannon")

ggdensity(data$value, 
          main = "Density plot of Shannon diversity",
          xlab = "diversity")

# or Q-Q plot
ggqqplot(data$value)

# Visual inspection, described in the previous section, is usually unreliable. It's possible to use a significance test comparing the sample distribution to a normal one in order to ascertain whether data show or not a serious deviation from normality.

# Shapiro-Wilk's method is widely recommended for normality test and it provides better power than K-S. It is based on the correlation between the data and the corresponding normal scores.
shapiro.test(data$value) #Shannon
#W = 0.97751, p-value = 0.009633-07
# they do not follow normal distribution

```



```{r Shannon boxplott}
set.seed(202)

ps.rel.nose2@sam_data$Sample_Type <- factor(ps.rel.nose2@sam_data$Sample_Type,
                                             levels = c("NPS", "MTS", "MLF"))


# Step 1: Extract metadata
metadata <- as.data.frame(sample_data(ps.rel.nose2))
metadata <- data.frame(metadata)
# Step 2: Identify common IDs across MTS, MLF, and NP
ids_mts <- metadata %>% filter(Sample_Type == "MTS") %>% pull(ID) %>% unique()
ids_mlf <- metadata %>% filter(Sample_Type == "MLF") %>% pull(ID) %>% unique()
ids_np  <- metadata %>% filter(Sample_Type == "NPS")  %>% pull(ID) %>% unique()

# Find common IDs present in all three sample types
common_ids <- Reduce(intersect, list(ids_mts, ids_mlf, ids_np))

# Step 3: Filter metadata to only include common IDs
filtered_metadata <- metadata %>%
  filter(ID %in% common_ids)

# Step 4: Randomly select either Left or Right for MTS and MLF (one per ID)
set.seed(123)  # For reproducibility

selected_samples_mts <- filtered_metadata %>%
  filter(Sample_Type == "MTS") %>%
  group_by(ID) %>%
  sample_n(1) %>%  # Randomly select one sample per ID (Left or Right)
  ungroup()

selected_samples_mlf <- filtered_metadata %>%
  filter(Sample_Type == "MLF") %>%
  group_by(ID) %>%
  sample_n(1) %>%  # Randomly select one sample per ID (Left or Right)
  ungroup()

# Step 5: Keep NP samples unchanged (only for common IDs)
selected_samples_np <- filtered_metadata %>%
  filter(Sample_Type == "NPS")

# Step 6: Combine all selected samples
selected_samples <- bind_rows(selected_samples_mts, selected_samples_mlf, selected_samples_np)
rownames(selected_samples) <- selected_samples$sample_identifier

# Step 7: Subset the phyloseq object based on the selected samples
ps.nose.prof.filtered <- prune_samples(
  sample_names(ps.rel.nose2) %in% rownames(selected_samples),
  ps.rel.nose2
)

# Step 8: Filter to keep only Timepoint T1 (or NA)
ps.nose.prof.filtered <- ps.nose.prof.filtered

# Step 9: Reorder sample_type factor levels
ps.nose.prof.filtered@sam_data$Sample_Type <- factor(
  ps.nose.prof.filtered@sam_data$Sample_Type,
  levels = c("NPS", "MTS", "MLF")
)

# 
table(sample_data(ps.nose.prof.filtered)$Sample_Type)
length(unique(sample_data(ps.nose.prof.filtered)$ID))  # Should match the number of common IDs


# comparisons
#my_comparisons <- list( c("H_Control", "H_ILI"), c("B_Control", "B_ILI"), c("V_Control", "V_ILI"))
my_comparisons <- list(c("NPS", "MTS"), c("NPS", "MLF"), c("MLF","MTS"))

#pal <- c("#A6CEE3","#1F78B4")
theme_set(theme_bw())

# only significant comparison
#my_comparisons <- list( c("Control", "B"), c("H", "B"))
#Shannon
p1 <- plot_richness(ps.nose.prof.filtered, x="Sample_Type",color="Sample_Type", measures="Shannon") 
p1 <- p1 + 
  geom_point(aes(color = sample_type),
             position = position_jitter(width = 0.15),
             size = 2, alpha = 0.25) +
  geom_boxplot( width = 0.2, outlier.shape = NA,
               alpha = 0.5) + 
   geom_half_violin(position = position_nudge(x = 0.15, y = 0),
                   alpha = 0.25, side = "r") +
  geom_hline(yintercept = median((p1$data)$value), linetype = 2, color= "grey") +
  guides(colour = guide_legend(nrow = 1)) +
  scale_colour_manual(values = pcoa_colors) +
  scale_fill_manual(values = pcoa_colors) +
  stat_compare_means(
    comparisons = my_comparisons,
    aes(label = ..p.signif..), 
    method = "wilcox.test",p.adjust.method = "fdr",
    paired = TRUE,
    label.y = c(3.2, 3, 2.8), 
    label.x.npc = 0.2,
    tip.length = 0.01)+
  theme_biome_utils() + xlab("") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
p1 <- p1 + scale_fill_manual("Sample Type", values = (pcoa_colors))
p1


#for the observed
sel <- c("MTS", "NPS", "MLF")
ps.nose.prof.obs <- ps.nasal %>% ps_filter(Sample_Type %in% sel)
# #We need to filter out the extra timepoints
ps.nose.prof.obs <- ps.nose.prof.obs 
# Step 1: Extract metadata
metadata <- as.data.frame(sample_data(ps.nose.prof.obs))
metadata <- data.frame(metadata)
# Step 2: Identify common IDs across MTS, MLF, and NP
ids_mts <- metadata %>% filter(Sample_Type == "MTS") %>% pull(ID) %>% unique()
ids_mlf <- metadata %>% filter(Sample_Type == "MLF") %>% pull(ID) %>% unique()
ids_np  <- metadata %>% filter(Sample_Type == "NPS")  %>% pull(ID) %>% unique()

# Find common IDs present in all three sample types
common_ids <- Reduce(intersect, list(ids_mts, ids_mlf, ids_np))

# Step 3: Filter metadata to only include common IDs
filtered_metadata <- metadata %>%
  filter(ID %in% common_ids)

# Step 4: Randomly select either Left or Right for MTS and MLF (one per ID)
set.seed(123)  # For reproducibility

selected_samples_mts <- filtered_metadata %>%
  filter(Sample_Type == "MTS") %>%
  group_by(ID) %>%
  sample_n(1) %>%  # Randomly select one sample per ID (Left or Right)
  ungroup()

selected_samples_mlf <- filtered_metadata %>%
  filter(Sample_Type == "MLF") %>%
  group_by(ID) %>%
  sample_n(1) %>%  # Randomly select one sample per ID (Left or Right)
  ungroup()

# Step 5: Keep NP samples unchanged (only for common IDs)
selected_samples_np <- filtered_metadata %>%
  filter(Sample_Type == "NPS")

# Step 6: Combine all selected samples
selected_samples <- bind_rows(selected_samples_mts, selected_samples_mlf, selected_samples_np)
rownames(selected_samples) <- selected_samples$sample_identifier

# Step 7: Subset the phyloseq object based on the selected samples
ps.nose.prof.filtered.obs <- prune_samples(
  sample_names(ps.nose.prof.obs) %in% rownames(selected_samples),
  ps.nose.prof.obs
)



# Step 9: Reorder sample_type factor levels
ps.nose.prof.filtered.obs@sam_data$Sample_Type <- factor(
  ps.nose.prof.filtered.obs@sam_data$Sample_Type,
  levels = c("NPS", "MTS", "MLF")
)

# 
table(sample_data(ps.nose.prof.filtered.obs)$Sample_Type)
length(unique(sample_data(ps.nose.prof.filtered)$ID)) 

#Observed
p2 <- plot_richness(ps.nose.prof.filtered.obs, x="Sample_Type", color="Sample_Type", measures="Chao1")
p2 <- p2 +
  geom_point(aes(color = sample_type),
             position = position_jitter(width = 0.15),
             size = 2, alpha = 0.25) +
  geom_boxplot( width = 0.2, outlier.shape = NA,
               alpha = 0.5) +
   geom_half_violin(position = position_nudge(x = 0.15, y = 0),
                   alpha = 0.25, side = "r") +
  geom_hline(yintercept = median((p2$data)$value), linetype = 2, color= "grey") +
  scale_colour_manual(values = pal) +
  scale_fill_manual(values = pal) +
  stat_compare_means(comparisons = my_comparisons,
                     aes(label = ..p.signif..),
                     method = "wilcox.test",p.adjust.method = "fdr",
                     label.y = c(420, 390, 360),
                     label.x.npc = 0.2, tip.length = 0.01)+
  theme_biome_utils() + xlab("") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

p2 <- p2 + scale_colour_manual("Sample Type", values = (pcoa_colors))
p2

#Simpson
p3 <- plot_richness(ps.nose.prof.filtered, x="Sample_Type", color="Sample_Type", measures="Simpson") 
p3 <- p3 + 
  geom_point(aes(color = Sample_Type),
             position = position_jitter(width = 0.15),
             size = 2, alpha = 0.25) +
  geom_boxplot( width = 0.2, outlier.shape = NA,
               alpha = 0.5) + 
   geom_half_violin(position = position_nudge(x = 0.15, y = 0),
                   alpha = 0.25, side = "r") +
  geom_hline(yintercept = median((p3$data)$value), linetype = 2, color= "grey") +
  scale_colour_manual(values = pal) +
  scale_fill_manual(values = pal) +
  stat_compare_means(comparisons = my_comparisons,
                     aes(label = ..p.signif..), 
                     method = "wilcox.test",p.adjust.method = "fdr",
                     paired=TRUE,
                     label.y = c(1.1, 1, 0.9),
                     label.x.npc = 0.2, tip.length = 0.01)+
  theme_biome_utils() + xlab("") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

p3 <- p3 + scale_colour_manual("Sample Type", values = (pcoa_colors))
p3


figure_legend <- get_legend(p1)

figure_1 <- ggarrange(p1,  p3,
                    labels = c(" "),
                    ncol = 2, nrow = 1,
                    common.legend = TRUE, legend = "right", legend.grob = figure_legend)

figure_1
ggsave(
  filename    = "Alpha_div_Nasal_Samples.tiff",
  plot        = figure_1,
  device      = "tiff",
  dpi         = 300,
  width       = 10,      # in inches
  height      = 8,      # adjust as needed
  units       = "in",
  compression = "lzw")
```





# Alpha Diversity: OPS & Saliva
```{r}
set.seed(202)

#Calculate relative abundances
ps.rel.oral2 <- prune_taxa(taxa_sums(ps.rel.oral)>0, ps.rel.oral)
ps.rel.oral.comp <- microbiome::transform(ps.rel.oral2, "compositional")
# First we use plot_richness to calculate alpha diversity values
p <- plot_richness(ps.rel.oral2, x="Sample_Type", color="Sample_Type", measures=c("Shannon")) 

levels((p$data)$variable)

# Subset the data values ofr the Shannon indices
data <- subset((p$data), variable== "Shannon")

ggdensity(data$value, 
          main = "Density plot of Shannon diversity",
          xlab = "diversity")

# or Q-Q plot
ggqqplot(data$value)

# Visual inspection, described in the previous section, is usually unreliable. It's possible to use a significance test comparing the sample distribution to a normal one in order to ascertain whether data show or not a serious deviation from normality.

# Shapiro-Wilk's method is widely recommended for normality test and it provides better power than K-S. It is based on the correlation between the data and the corresponding normal scores.
shapiro.test(data$value) #Shannon
#W = 0.98047, p-value = 0.2603
# not normally distributed?

```


```{r, fig.width=13, fig.height=9}
set.seed(202)

# Step 1: Extract metadata
metadata <- as.data.frame(sample_data(ps.rel.oral2))
metadata <- data.frame(metadata)
# Step 2: Identify common IDs across MTS, MLF, and NP
ids_ops <- metadata %>% filter(Sample_Type == "OPS") %>% pull(ID) %>% unique()
ids_saliva <- metadata %>% filter(Sample_Type == "Saliva") %>% pull(ID) %>% unique()


# Find common IDs present in all three sample types
common_ids <- Reduce(intersect, list(ids_ops, ids_saliva))

# Step 3: Filter metadata to only include common IDs
filtered_metadata <- metadata %>%
  filter(ID %in% common_ids)

# Step 4: Randomly select either Left or Right for MTS and MLF (one per ID)
set.seed(123)  # For reproducibility

selected_samples_ops <- filtered_metadata %>%
  filter(Sample_Type == "OPS") %>%
  group_by(ID) %>%
  sample_n(1) %>%  # Randomly select one sample per ID (Left or Right)
  ungroup()

selected_samples_saliva <- filtered_metadata %>%
  filter(Sample_Type == "Saliva") %>%
  group_by(ID) %>%
  sample_n(1) %>%  # Randomly select one sample per ID (Left or Right)
  ungroup()



# Step 6: Combine all selected samples
selected_samples <- bind_rows(selected_samples_ops, selected_samples_saliva)
rownames(selected_samples) <- selected_samples$sample_identifier

# Step 7: Subset the phyloseq object based on the selected samples
ps.oral.prof.filtered <- prune_samples(
  sample_names(ps.rel.oral2) %in% rownames(selected_samples),
  ps.rel.oral2
)


# Step 9: Reorder sample_type factor levels
ps.oral.prof.filtered@sam_data$Sample_Type <- factor(
  ps.oral.prof.filtered@sam_data$Sample_Type,
  levels = c("OPS", "Saliva")
)

# 
table(sample_data(ps.oral.prof.filtered)$Sample_Type)
length(unique(sample_data(ps.oral.prof.filtered)$ID))  # Should match the number of common IDs




# comparisons
my_comparisons <- list(c("OPS", "Saliva"))

pal <- c(OPS ="#BC7C7C", Saliva="#8ABFA3")
theme_set(theme_bw())

# only significant comparison
#my_comparisons <- list( c("Control", "B"), c("H", "B"))
#Shannon
p1 <- plot_richness(ps.oral.prof.filtered, x="Sample_Type",color="Sample_Type", measures="Shannon") 
p1 <- p1 + 
  geom_point(aes(color = Sample_Type),
             position = position_jitter(width = 0.15),
             size = 2, alpha = 0.25) +
  geom_boxplot(fill = pal, width = 0.2, outlier.shape = NA,
               alpha = 0.5) + 
   geom_half_violin(position = position_nudge(x = 0.15, y = 0),
                   alpha = 0.25, side = "r") +
  geom_hline(yintercept = median((p1$data)$value), linetype = 2, color= "grey") +
  guides(colour = guide_legend(nrow = 1)) +
  scale_colour_manual(values = pal) +
  scale_fill_manual(values = pal) +
  stat_compare_means(
    comparisons = list(c("OPS", "Saliva")),
    aes(label = ..p.signif..), 
    method = "wilcox.test",p.adjust.method = "fdr",
    paired = TRUE,
    label.y = 3.7, 
    label.x.npc = 0.2,
    tip.length = 0.01)+
  theme_biome_utils() + xlab("") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
p1 <- p1 + scale_fill_manual("Sample Type", values = (pal))
p1

#for the observed
ps.oral.prof.obs <- ps.oral 
table(as.data.frame(sample_data(ps.mts.prof.obs))$Nostril)



ps.oral.prof.obs <- ps.oral.prof.obs %>% ps_filter(Sample_Type %in% c("OPS", "Saliva"))

# Convert to data frame if needed
oral_df <- as.data.frame(sample_data(ps.oral.prof.obs))

# Identify samples from each nostril
ops_samples <- rownames(oral_df[oral_df$Sample_Type == "OPS", ])
saliva_samples <- rownames(oral_df[oral_df$Sample_Type == "Saliva", ])

# Randomly sample 33 Right nostril samples to match the Left
set.seed(123)  # For reproducibility
right_samples_sub <- sample(saliva_samples, length(ops_samples))

# Combine Left samples with the downsampled Right samples
balanced_samples <- c(ops_samples, right_samples_sub)

# Subset the phyloseq object to retain only balanced samples
ps.oral.balanced.obs <- prune_samples(balanced_samples, ps.oral.prof.obs)

# Check the new distribution
table(sample_data(ps.oral.balanced.obs)$Nostril)
#Observed
p2 <- plot_richness(ps.oral.balanced.obs, x="Sample_Type", color="Sample_Type", measures="Chao1")
p2 <- p2 +
  geom_point(aes(color = Sample_Type),
             position = position_jitter(width = 0.15),
             size = 2, alpha = 0.25) +
  geom_boxplot( width = 0.2, outlier.shape = NA,
               alpha = 0.5) +
   geom_half_violin(position = position_nudge(x = 0.15, y = 0),
                   alpha = 0.25, side = "r") +
  geom_hline(yintercept = median((p2$data)$value), linetype = 2, color= "grey") +
  scale_colour_manual(values = pal) +
  scale_fill_manual(values = pal) +
  stat_compare_means(comparisons = list(c("OPS", "Saliva")),
                     aes(label = ..p.signif..),
                     method = "wilcox.test",p.adjust.method = "fdr",
                     paired = TRUE,
                     label.y = 255,
                     label.x.npc = 0.2, tip.length = 0.01)+
  theme_biome_utils() + xlab("") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

p2 <- p2 + scale_colour_manual("Sample_Type", values = (pal))
p2

#Simpson
p3 <- plot_richness(ps.oral.prof.filtered, x="Sample_Type", color="Sample_Type", measures="Simpson") 
p3 <- p3 + 
  geom_point(aes(color = Sample_Type),
             position = position_jitter(width = 0.15),
             size = 2, alpha = 0.25) +
  geom_boxplot(fill = pal , width = 0.2, outlier.shape = NA,
               alpha = 0.5) + 
   geom_half_violin(position = position_nudge(x = 0.15, y = 0),
                   alpha = 0.25, side = "r") +
  geom_hline(yintercept = median((p3$data)$value), linetype = 2, color= "grey") +
  scale_colour_manual(values = pal) +
  scale_fill_manual(values = pal) +
  stat_compare_means(comparisons = list(c("OPS", "Saliva")),
                     aes(label = ..p.signif..), 
                     method = "wilcox.test",
                     label.y = c(0.95),
                     paired=TRUE,
                     label.x.npc = 0.2, tip.length = 0.01)+
  theme_biome_utils() + xlab("") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

p3 <- p3 + scale_colour_manual("Sample_Type", values = (pal))
p3


figure_legend <- get_legend(p1)

figure_oral <- ggarrange(p1, p3,
                    labels = c(" "),
                    ncol = 2, nrow = 1,
                    common.legend = TRUE, legend = "right", legend.grob = figure_legend)

figure_oral

ggsave(
  filename    = "Alpha_div_Oral_Samples.tiff",
  plot        = figure_oral,
  device      = "tiff",
  dpi         = 300,
  width       = 10,      # in inches
  height      = 8,      # adjust as needed
  units       = "in",
  compression = "lzw")
```



# Alpha Diversity: All Sample Types 
```{r}
set.seed(202)
#Calculate relative abundances
ps.rel.prof2 <- prune_taxa(taxa_sums(ps.rel.prof)>0, ps.rel.prof)
ps.rel.prof.comp <- microbiome::transform(ps.rel.prof2, "compositional")
# First we use plot_richness to calculate alpha diversity values
p <- plot_richness(ps.rel.prof2, x="Sample_Type", color="Sample_Type", measures=c("Shannon")) 

levels((p$data)$variable)

# Subset the data values ofr the Shannon indices
data <- subset((p$data), variable== "Shannon")

ggdensity(data$value, 
          main = "Density plot of Shannon diversity",
          xlab = "diversity")

# or Q-Q plot
ggqqplot(data$value)

# Visual inspection, described in the previous section, is usually unreliable. It's possible to use a significance test comparing the sample distribution to a normal one in order to ascertain whether data show or not a serious deviation from normality.

# Shapiro-Wilk's method is widely recommended for normality test and it provides better power than K-S. It is based on the correlation between the data and the corresponding normal scores.
shapiro.test(data$value) #Shannon
#W = 0.95444, p-value = 9.108e-07
#  not normally distributed

```

```{r}
# 1. collapse each to genus if you like, or leave ASV-level:
#    (here I’ll leave them ASV-level)
# 2. prefix the ASV names so they never collide
taxa_names(ps.nose.og) <- paste0("Nose__", taxa_names(ps.nose.og))
taxa_names(ps.oral.og) <- paste0("Oral__", taxa_names(ps.oral.og))

# 3. now merge
ps.both2 <- merge_phyloseq(ps.nose.og, ps.oral.og)



```


```{r, fig.width=13, fig.height=9}
set.seed(202)

# Step 1: Extract metadata
metadata <- as.data.frame(sample_data(ps.rel.prof2))
metadata <- data.frame(metadata)
table(metadata$Sample_Type)
length(unique(metadata$ID))             # total participants
table(table(metadata$ID))  

# Step 2: Identify common IDs across MTS, MLF, and NP
ids_mts <- metadata %>% filter(Sample_Type == "MTS") %>% pull(ID) %>% unique()
ids_mlf <- metadata %>% filter(Sample_Type == "MLF") %>% pull(ID) %>% unique()
ids_nps  <- metadata %>% filter(Sample_Type == "NPS")  %>% pull(ID) %>% unique()
ids_ops <- metadata %>% filter(Sample_Type == "OPS") %>% pull(ID) %>% unique()
ids_saliva <- metadata %>% filter(Sample_Type == "Saliva") %>% pull(ID) %>% unique()
# Pairwise overlaps
pairwise_overlaps <- list(
  NPS_MTS   = length(intersect(ids_nps, ids_mts)),
  NPS_MLF   = length(intersect(ids_nps, ids_mlf)),
  NPS_OPS   = length(intersect(ids_nps, ids_ops)),
  NPS_Saliva= length(intersect(ids_nps, ids_saliva)),
  MTS_MLF   = length(intersect(ids_mts, ids_mlf)),
  MTS_OPS   = length(intersect(ids_mts, ids_ops)),
  MTS_Saliva= length(intersect(ids_mts, ids_saliva)),
  MLF_OPS   = length(intersect(ids_mlf, ids_ops)),
  MLF_Saliva= length(intersect(ids_mlf, ids_saliva)),
  OPS_Saliva= length(intersect(ids_ops, ids_saliva))
)
pairwise_overlaps

# Force Right nostril samples for MLF and MTS
selected_samples_mlf <- filtered_metadata %>%
  filter(Sample_Type == "MLF" & Nostril == "Right")

selected_samples_mts <- filtered_metadata %>%
  filter(Sample_Type == "MTS" & Nostril == "Right")

# Keep NPS, OPS, Saliva unchanged
selected_samples_nps    <- filtered_metadata %>% filter(Sample_Type == "NPS")
selected_samples_ops    <- filtered_metadata %>% filter(Sample_Type == "OPS")
selected_samples_saliva <- filtered_metadata %>% filter(Sample_Type == "Saliva")

# Combine all
selected_samples <- bind_rows(
  selected_samples_mlf,
  selected_samples_mts,
  selected_samples_nps,
  selected_samples_ops,
  selected_samples_saliva
)

# Count unique IDs
length(unique(selected_samples$ID))

# Step 2: Identify common IDs across MTS, MLF, and NP
ids_ops <- metadata %>% filter(Sample_Type == "OPS") %>% pull(ID) %>% unique()
ids_saliva <- metadata %>% filter(Sample_Type == "Saliva") %>% pull(ID) %>% unique()
ids_nps <- metadata %>% filter(Sample_Type == "NPS") %>% pull(ID) %>% unique()
ids_mts <- metadata %>% filter(Sample_Type == "MTS") %>% pull(ID) %>% unique()
ids_mlf <- metadata %>% filter(Sample_Type == "MLF") %>% pull(ID) %>% unique()

# Find common IDs present in all three sample types
common_ids <- Reduce(intersect, list(ids_ops, ids_saliva,ids_nps,
                                     ids_mts,ids_mlf))

# Step 3: Filter metadata to only include common IDs
filtered_metadata <- metadata %>%
  filter(ID %in% common_ids)

# Step 4: Randomly select either Left or Right for MTS and MLF (one per ID)
set.seed(123)  # For reproducibility

selected_samples_ops <- filtered_metadata %>%
  filter(Sample_Type == "OPS") %>%
  group_by(ID) %>%
  sample_n(1) %>%  # Randomly select one sample per ID (Left or Right)
  ungroup()

selected_samples_saliva <- filtered_metadata %>%
  filter(Sample_Type == "Saliva") %>%
  group_by(ID) %>%
  sample_n(1) %>%  # Randomly select one sample per ID (Left or Right)
  ungroup()


selected_samples_nps <- filtered_metadata %>%
  filter(Sample_Type == "NPS") %>%
  group_by(ID) %>%
  sample_n(1) %>%  # Randomly select one sample per ID (Left or Right)
  ungroup()


selected_samples_mts <- filtered_metadata %>%
  filter(Sample_Type == "MTS") %>%
  group_by(ID) %>%
  sample_n(1) %>%  # Randomly select one sample per ID (Left or Right)
  ungroup()


selected_samples_mlf <- filtered_metadata %>%
  filter(Sample_Type == "MLF") %>%
  group_by(ID) %>%
  sample_n(1) %>%  # Randomly select one sample per ID (Left or Right)
  ungroup()
# Step 6: Combine all selected samples
selected_samples <- bind_rows(selected_samples_ops, selected_samples_saliva,selected_samples_nps,
                              selected_samples_mts,selected_samples_mlf)
rownames(selected_samples) <- selected_samples$sample_identifier

# Step 7: Subset the phyloseq object based on the selected samples
ps.prof.filtered <- prune_samples(
  sample_names(ps.rel.prof2) %in% rownames(selected_samples),
  ps.rel.prof2
)


# Step 9: Reorder sample_type factor levels
ps.prof.filtered@sam_data$Sample_Type <- factor(
  ps.prof.filtered@sam_data$Sample_Type,
  levels = c("NPS", "MTS","MLF", "OPS", "Saliva")
)

# 
table(sample_data(ps.prof.filtered)$Sample_Type)
length(unique(sample_data(ps.prof.filtered)$ID))  # Should match the number of common IDs




# comparisons
my_comparisons <- list(c("NPS", "MTS","MLF", "OPS", "Saliva"))

pal <- pcoa_colors
theme_set(theme_bw())

# only significant comparison
#my_comparisons <- list( c("Control", "B"), c("H", "B"))
#Shannon
p1 <- plot_richness(ps.prof.filtered, x="Sample_Type",color="Sample_Type", measures="Shannon") 
p1 <- p1 + 
  geom_point(aes(color = Sample_Type),
             position = position_jitter(width = 0.15),
             size = 2, alpha = 0.25) +
  geom_boxplot(fill = pal, width = 0.2, outlier.shape = NA,
               alpha = 0.5) + 
   geom_half_violin(position = position_nudge(x = 0.15, y = 0),
                   alpha = 0.25, side = "r") +
  geom_hline(yintercept = median((p1$data)$value), linetype = 2, color= "grey") +
  guides(colour = guide_legend(nrow = 1)) +
  scale_colour_manual(name= "Sample Type",values = pal) +
  scale_fill_manual(name= "Sample Type",values = pal) +
  stat_compare_means(
    comparisons = list(c("NPS", "MTS"),
  c("NPS", "MLF"),
  c("NPS", "OPS"),
  c("NPS", "Saliva"),
  c("MTS", "MLF"),
  c("MTS", "OPS"),
  c("MTS", "Saliva"),
  c("MLF", "OPS"),
  c("MLF", "Saliva"),
  c("OPS", "Saliva")),
    aes(label = ..p.signif..), 
    method = "wilcox.test",p.adjust.method = "fdr",
    paired = TRUE,
    label.y = c(5.1, 4.8, 4.5, 4.1, 3.7, 3.4, 3.2, 3.0, 4.3, 3.8), 
    label.x.npc = 0.2,
    tip.length = 0.01)+
  theme_biome_utils() + xlab("") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
p1 <- p1 + scale_fill_manual("Sample Type",name= "Sample Type", values = (pal))+  theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text(angle=45, hjust = 1, size = 22),
    axis.text.y      = element_text(size = 22),
    strip.text       = element_text(size = 25),  # facet labels
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
  )
p1

#for the observed
ps.prof.obs <- ps.both2 
#view(sample_data(ps.prof.obs))
table(as.data.frame(sample_data(ps.prof.obs))$sample_type2)



ps.prof.obs <- ps.prof.obs %>% ps_filter(sample_type2 %in% c("NPS", "MTS","MLF", "OPS", "Saliva"))

# Convert to data frame if needed
prof_df <- as.data.frame(sample_data(ps.prof.obs))

# Identify samples from each nostril
ops_samples <- rownames(prof_df[prof_df$sample_type2 == "OPS", ])
saliva_samples <- rownames(prof_df[prof_df$sample_type2 == "Saliva", ])
nps_samples <- rownames(prof_df[prof_df$sample_type2 == "NPS", ])
mts_samples <- rownames(prof_df[prof_df$sample_type2 == "MTS", ])
mlf_samples <- rownames(prof_df[prof_df$sample_type2 == "MLF", ])

# Randomly sample 33 Right nostril samples to match the Left
# Determine the minimum sample count across groups
min_count <- min(length(ops_samples), length(saliva_samples),
                 length(nps_samples), length(mts_samples),
                 length(mlf_samples))

# Set seed for reproducibility
set.seed(123)

# Randomly subsample each group to have the same number of samples
ops_sub <- sample(ops_samples, min_count)
saliva_sub <- sample(saliva_samples, min_count)
nps_sub <- sample(nps_samples, min_count)
mts_sub <- sample(mts_samples, min_count)
mlf_sub <- sample(mlf_samples, min_count)

# Combine the balanced sample IDs
balanced_samples <- c(ops_sub, saliva_sub, nps_sub, mts_sub, mlf_sub)

# Prune your phyloseq object to retain only balanced samples
ps.prof.obs.balanced <- prune_samples(balanced_samples, ps.prof.obs)

# Check the new distribution
table(sample_data(ps.prof.obs.balanced)$sample_type2)
#Observed
p2 <- plot_richness(ps.prof.obs.balanced, x="sample_type2", color="sample_type2", measures="Chao1")
p2 <- p2 +
  geom_point(aes(color = sample_type2),
             position = position_jitter(width = 0.15),
             size = 2, alpha = 0.25) +
  geom_boxplot( width = 0.2, outlier.shape = NA,
               alpha = 0.5) +
   geom_half_violin(position = position_nudge(x = 0.15, y = 0),
                   alpha = 0.25, side = "r") +
  geom_hline(yintercept = median((p2$data)$value), linetype = 2, color= "grey") +
  scale_colour_manual(values = pal) +
  scale_fill_manual(values = pal) +
  stat_compare_means(comparisons = list(c("NPS", "MTS"),
  c("NPS", "MLF"),
  c("NPS", "OPS"),
  c("NPS", "Saliva"),
  c("MTS", "MLF"),
  c("MTS", "OPS"),
  c("MTS", "Saliva"),
  c("MLF", "OPS"),
  c("MLF", "Saliva"),
  c("OPS", "Saliva")),
                     aes(label = ..p.signif..),
                     method = "wilcox.test",p.adjust.method = "fdr",
                     paired = TRUE,
                     label.y = c(1500, 1400, 1300, 1200, 1150),
                     label.x.npc = 0.2, tip.length = 0.01)+
  theme_biome_utils() + xlab("") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

p2 <- p2 + scale_colour_manual("Sample_Type",name= "Sample Type", values = (pal))
p2

#Simpson
p3 <- plot_richness(ps.prof.filtered, x="Sample_Type", color="Sample_Type", measures="Simpson") 
p3 <- p3 + 
  geom_point(aes(color = Sample_Type),
             position = position_jitter(width = 0.15),
             size = 2, alpha = 0.25) +
  geom_boxplot(fill = pal , width = 0.2, outlier.shape = NA,
               alpha = 0.5) + 
   geom_half_violin(position = position_nudge(x = 0.15, y = 0),
                   alpha = 0.25, side = "r") +
  geom_hline(yintercept = median((p3$data)$value), linetype = 2, color= "grey") +
  scale_colour_manual(name= "Sample Type", values = pal) +
  scale_fill_manual(name= "Sample Type",values = pal) +
  stat_compare_means(comparisons = list(
  c("NPS", "MTS"),
  c("NPS", "MLF"),
  c("NPS", "OPS"),
  c("NPS", "Saliva"),
  c("MTS", "MLF"),
  c("MTS", "OPS"),
  c("MTS", "Saliva"),
  c("MLF", "OPS"),
  c("MLF", "Saliva"),
  c("OPS", "Saliva")
),
                     aes(label = ..p.signif..), 
                     method = "wilcox.test",p.adjust.method = "fdr",
                     label.y = c(1.45, 1.35, 1.25, 1.15, 1.05, 1, 0.95, 1.4, 1.1, 1.2),
                     paired=TRUE,
                     label.x.npc = 0.2, tip.length = 0.01)+
  theme_biome_utils() + xlab("") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

p3 <- p3 + scale_colour_manual("sample_type2", values = (pal))+  theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text(angle=45,hjust=1, size = 22),
    axis.text.y      = element_text(size = 22),
    strip.text       = element_text(size = 25),  # facet labels
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
  )
p3







figure_legend <- get_legend(p1)

figure_all <- ggarrange(p1, p3,
                    labels = c(" "),
                    ncol = 2, nrow = 1,
                    common.legend = TRUE, legend = "bottom", legend.grob = figure_legend)

figure_all


ggsave(
  filename    = "Alpha_div_All_Samples-new.tiff",
  plot        = figure_all,
  device      = "tiff",
  dpi         = 300,
  width       = 13,      # in inches
  height      = 11,      # adjust as needed
  units       = "in",
  compression = "lzw")
```



# MLF: Self vs Professionally Collected Samples (alpha-diversity)
```{r, fig.width=13, fig.height=9}
library(gghalves)
mlf_df <- sample_data(ps.mlf.str.2)
# Convert to data frame if needed
mlf_df <- as.data.frame(sample_data(ps.mlf.str.2))

# Identify samples from each nostril
left_samples <- rownames(mlf_df[mlf_df$Sampling_Strategy == "Self", ])
right_samples <- rownames(mlf_df[mlf_df$Sampling_Strategy == "Professional", ])

# Randomly sample 33 Right nostril samples to match the Left
set.seed(123)  # For reproducibility
right_samples_sub <- sample(right_samples, length(left_samples))

# Combine Left samples with the downsampled Right samples
balanced_samples <- c(left_samples, right_samples_sub)

# Subset the phyloseq object to retain only balanced samples
ps.mlf.balanced <- prune_samples(balanced_samples, ps.mlf.str.2)

# Check the new distribution
table(sample_data(ps.mlf.balanced)$Sampling_Strategy)

# Check the first few values of the OTU table

# comparisons
my_comparisons <- list( c("Self", "Professional"))

pal <- sampling_strategy_colors
theme_set(theme_bw())

# only significant comparison
#Shannon
p1 <- plot_richness(ps.mlf.balanced, x="Sampling_Strategy",color="Sampling_Strategy", measures="Shannon") 
p1 <- p1 + 
  geom_point(aes(color = Sampling_Strategy),
             position = position_jitter(width = 0.15),
             size = 2, alpha = 0.25) +
  geom_boxplot(fill = pal, width = 0.2, outlier.shape = NA,
               alpha = 0.5) + 
   geom_half_violin(position = position_nudge(x = 0.15, y = 0),
                   alpha = 0.25, side = "r") +
  geom_hline(yintercept = mean((p1$data)$value), linetype = 2, color= "grey") +
  guides(colour = guide_legend(nrow = 1)) +
  scale_colour_manual(values = pal) +
  scale_fill_manual(values = pal) +
  stat_compare_means(
    comparisons = list(c("Self", "Professional")),
    aes(label = ..p.signif..), 
    method = "t.test",p.adjust.method = "fdr",
    paired = TRUE,
    label.y = 3.6, 
    label.x.npc = 0.2,
    tip.length = 0.01)+
  theme_biome_utils() + xlab("") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
p1 <- p1 + scale_fill_manual("Sampling Strategy", values = (pal))+  theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text(angle=45, hjust=1, size = 22),
    axis.text.y      = element_text(size = 22),
    strip.text       = element_text(size = 25),  # facet labels
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
  )
p1



#for the observed
ps.mlf.prof.obs <- ps.nasal %>% ps_filter(Sample_Type =="MLF")
table(as.data.frame(sample_data(ps.mlf.prof.obs))$Sampling_Strategy)

# Convert to data frame if needed
mlf_df <- as.data.frame(sample_data(ps.mlf.prof.obs))

# Identify samples from each nostril
left_samples <- rownames(mlf_df[mlf_df$Sampling_Strategy == "Self", ])
right_samples <- rownames(mlf_df[mlf_df$Sampling_Strategy == "Professional", ])

# Randomly sample 33 Right nostril samples to match the Left
set.seed(123)  # For reproducibility
right_samples_sub <- sample(right_samples, length(left_samples))

# Combine Left samples with the downsampled Right samples
balanced_samples <- c(left_samples, right_samples_sub)

# Subset the phyloseq object to retain only balanced samples
ps.mlf.balanced.obs <- prune_samples(balanced_samples, ps.mlf.prof.obs)

# Check the new distribution
table(sample_data(ps.mlf.balanced.obs)$Nostril)

#Observed
p2 <- plot_richness(ps.mlf.balanced.obs, x="Sampling_Strategy", color="Sampling_Strategy", measures="Chao1")
p2 <- p2 +
  geom_point(aes(color = Sampling_Strategy),
             position = position_jitter(width = 0.15),
             size = 2, alpha = 0.25) +
  geom_boxplot(fill = pal , width = 0.2, outlier.shape = NA,
               alpha = 0.5) +
   geom_half_violin(position = position_nudge(x = 0.15, y = 0),
                   alpha = 0.25, side = "r") +
  geom_hline(yintercept = mean((p2$data)$value), linetype = 2, color= "grey") +
  scale_colour_manual(values = pal) +
  scale_fill_manual(values = pal) +
  stat_compare_means(comparisons = list(c("Self", "Professional")),
                     aes(label = ..p.signif..),
                     method = "t.test",p.adjust.method = "fdr",
                     paired = TRUE,
                     label.y = 1015,
                     label.x.npc = 0.2, tip.length = 0.01)+
  theme_biome_utils() + xlab("") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

p2 <- p2 + scale_colour_manual("Sampling Strategy", values = (pal))
p2

#Simpson
p3 <- plot_richness(ps.mlf.balanced, x="Sampling_Strategy", color="Sampling_Strategy", measures="Simpson") 
p3 <- p3 + 
  geom_point(aes(color = Sampling_Strategy),
             position = position_jitter(width = 0.15),
             size = 2, alpha = 0.25) +
  geom_boxplot(fill = pal , width = 0.2, outlier.shape = NA,
               alpha = 0.5) + 
   geom_half_violin(position = position_nudge(x = 0.15, y = 0),
                   alpha = 0.25, side = "r") +
  geom_hline(yintercept = mean((p3$data)$value), linetype = 2, color= "grey") +
  scale_colour_manual(values = pal) +
  scale_fill_manual(values = pal) +
  stat_compare_means(comparisons = list(c("Self", "Professional")),
                     aes(label = ..p.signif..), 
                     method = "t.test",p.adjust.method = "fdr",
                     label.y = c(1),
                     paired=TRUE,
                     label.x.npc = 0.2, tip.length = 0.01)+
  theme_biome_utils() + xlab("") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

p3 <- p3 + scale_colour_manual("Sampling Strategy", values = (pal))+  theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text(angle=45, hjust=1, size = 22),
    axis.text.y      = element_text(size = 22),
    strip.text       = element_text(size = 25),  # facet labels
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
  )
p3


figure_legend <- get_legend(p1)

figure_mlf_str <- ggarrange(p1, p3,
                    labels = c(" "),
                    ncol = 2, nrow = 1,
                    common.legend = TRUE, legend = "right", legend.grob = figure_legend)

figure_mlf_str



ggsave(
  filename    = "Alpha_div_MLF_SampStrat.tiff",
  plot        = figure_mlf_str,
  device      = "tiff",
  dpi         = 300,
  width       = 10,      # in inches
  height      = 8,      # adjust as needed
  units       = "in",
  compression = "lzw")
```




# MTS: Self vs Professionally Collected Samples (alpha-diversity)

```{r, fig.width=13, fig.height=9}
library(gghalves)
mts_df <- sample_data(ps.mts.str.2)
# Convert to data frame if needed
mts_df <- as.data.frame(sample_data(ps.mts.str.2))

# Identify samples from each nostril
left_samples <- rownames(mts_df[mts_df$Sampling_Strategy == "Self", ])
right_samples <- rownames(mts_df[mts_df$Sampling_Strategy == "Professional", ])

# Randomly sample 33 Right nostril samples to match the Left
set.seed(123)  # For reproducibility
right_samples_sub <- sample(right_samples, length(left_samples))

# Combine Left samples with the downsampled Right samples
balanced_samples <- c(left_samples, right_samples_sub)

# Subset the phyloseq object to retain only balanced samples
ps.mts.balanced <- prune_samples(balanced_samples, ps.mts.str.2)

# Check the new distribution
table(sample_data(ps.mts.balanced)$Sampling_Strategy)

# Check the first few values of the OTU table

# comparisons
my_comparisons <- list( c("Self", "Professional"))

pal <- sampling_strategy_colors
theme_set(theme_bw())

# only significant comparison
#Shannon
p1 <- plot_richness(ps.mts.balanced, x="Sampling_Strategy",color="Sampling_Strategy", measures="Shannon") 
p1 <- p1 + 
  geom_point(aes(color = Sampling_Strategy),
             position = position_jitter(width = 0.15),
             size = 2, alpha = 0.25) +
  geom_boxplot(fill = pal, width = 0.2, outlier.shape = NA,
               alpha = 0.5) + 
   geom_half_violin(position = position_nudge(x = 0.15, y = 0),
                   alpha = 0.25, side = "r") +
  geom_hline(yintercept = mean((p1$data)$value), linetype = 2, color= "grey") +
  guides(colour = guide_legend(nrow = 1)) +
  scale_colour_manual(values = pal) +
  scale_fill_manual(values = pal) +
  stat_compare_means(
    comparisons = list(c("Self", "Professional")),
    aes(label = ..p.signif..), 
    method = "wilcox.test",p.adjust.method = "fdr",
    paired = TRUE,
    label.y = 3.6, 
    label.x.npc = 0.2,
    tip.length = 0.01)+
  theme_biome_utils() + xlab("") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
p1 <- p1 + scale_fill_manual("Sampling Strategy", values = (pal))+  theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text(angle=45, hjust=1, size = 22),
    axis.text.y      = element_text(size = 22),
    strip.text       = element_text(size = 25),  # facet labels
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
  )
p1



#for the observed
ps.mts.prof.obs <- ps.nasal %>% ps_filter(Sample_Type =="MTS")
table(as.data.frame(sample_data(ps.mts.prof.obs))$Sampling_Strategy)

# Convert to data frame if needed
mts_df <- as.data.frame(sample_data(ps.mts.prof.obs))

# Identify samples from each nostril
left_samples <- rownames(mts_df[mts_df$Sampling_Strategy == "Self", ])
right_samples <- rownames(mts_df[mts_df$Sampling_Strategy == "Professional", ])

# Randomly sample 33 Right nostril samples to match the Left
set.seed(123)  # For reproducibility
right_samples_sub <- sample(right_samples, length(left_samples))

# Combine Left samples with the downsampled Right samples
balanced_samples <- c(left_samples, right_samples_sub)

# Subset the phyloseq object to retain only balanced samples
ps.mts.balanced.obs <- prune_samples(balanced_samples, ps.mts.prof.obs)

# Check the new distribution
table(sample_data(ps.mts.balanced.obs)$Sampling_Strategy)

#Observed
p2 <- plot_richness(ps.mts.balanced.obs, x="Sampling_Strategy", color="Sampling_Strategy", measures="Chao1")
p2 <- p2 +
  geom_point(aes(color = Sampling_Strategy),
             position = position_jitter(width = 0.15),
             size = 2, alpha = 0.25) +
  geom_boxplot(fill = pal , width = 0.2, outlier.shape = NA,
               alpha = 0.5) +
   geom_half_violin(position = position_nudge(x = 0.15, y = 0),
                   alpha = 0.25, side = "r") +
  geom_hline(yintercept = mean((p2$data)$value), linetype = 2, color= "grey") +
  scale_colour_manual(values = pal) +
  scale_fill_manual(values = pal) +
  stat_compare_means(comparisons = list(c("Self", "Professional")),
                     aes(label = ..p.signif..),
                     method = "wilcox.test",p.adjust.method = "fdr",
                     paired = TRUE,
                     label.y = 915,
                     label.x.npc = 0.2, tip.length = 0.01)+
  theme_biome_utils() + xlab("") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

p2 <- p2 + scale_colour_manual("Sampling Strategy", values = (pal))
p2

#Simpson
p3 <- plot_richness(ps.mts.balanced, x="Sampling_Strategy", color="Sampling_Strategy", measures="Simpson") 
p3 <- p3 + 
  geom_point(aes(color = Sampling_Strategy),
             position = position_jitter(width = 0.15),
             size = 2, alpha = 0.25) +
  geom_boxplot(fill = pal , width = 0.2, outlier.shape = NA,
               alpha = 0.5) + 
   geom_half_violin(position = position_nudge(x = 0.15, y = 0),
                   alpha = 0.25, side = "r") +
  geom_hline(yintercept = mean((p3$data)$value), linetype = 2, color= "grey") +
  scale_colour_manual(values = pal) +
  scale_fill_manual(values = pal) +
  stat_compare_means(comparisons = list(c("Self", "Professional")),
                     aes(label = ..p.signif..), 
                     method = "wilcox.test",p.adjust.method = "fdr",
                     label.y = c(1),
                     paired=TRUE,
                     label.x.npc = 0.2, tip.length = 0.01)+
  theme_biome_utils() + xlab("") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

p3 <- p3 + scale_colour_manual("Sampling Strategy", values = (pal))+  theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text(angle=45, hjust=1,size = 22),
    axis.text.y      = element_text(size = 22),
    strip.text       = element_text(size = 25),  # facet labels
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
  )
p3


figure_legend <- get_legend(p1)

figure_mts_str <- ggarrange(p1, p3,
                    labels = c(" "),
                    ncol = 2, nrow = 1,
                    common.legend = TRUE, legend = "right", legend.grob = figure_legend)

figure_mts_str


ggsave(
  filename    = "Alpha_div_MTS_SampStrat.tiff",
  plot        = figure_mts_str,
  device      = "tiff",
  dpi         = 300,
  width       = 10,      # in inches
  height      = 8,      # adjust as needed
  units       = "in",
  compression = "lzw")

```





# Saliva: Self vs Professionally Collected Samples (alpha-diversity)
```{r, fig.width=13, fig.height=9}
library(gghalves)
sal_df <- sample_data(ps.saliva.str.2)
# Convert to data frame if needed
sal_df <- as.data.frame(sample_data(ps.saliva.str.2))

# Identify samples from each nostril
left_samples <- rownames(sal_df[sal_df$Sampling_Strategy == "Self", ])
right_samples <- rownames(sal_df[sal_df$Sampling_Strategy == "Professional", ])

# Randomly sample 33 Right nostril samples to match the Left
set.seed(123)  # For reproducibility
right_samples_sub <- sample(right_samples, length(left_samples))

# Combine Left samples with the downsampled Right samples
balanced_samples <- c(left_samples, right_samples_sub)

# Subset the phyloseq object to retain only balanced samples
ps.sal.balanced <- prune_samples(balanced_samples, ps.saliva.str.2)

# Check the new distribution
table(sample_data(ps.sal.balanced)$Sampling_Strategy)

# Check the first few values of the OTU table

# comparisons
#my_comparisons <- list( c("H_Control", "H_ILI"), c("B_Control", "B_ILI"), c("V_Control", "V_ILI"))
my_comparisons <- list( c("Self", "Professional"))

pal <- sampling_strategy_colors
theme_set(theme_bw())

# only significant comparison
#my_comparisons <- list( c("Control", "B"), c("H", "B"))
#Shannon
p1 <- plot_richness(ps.sal.balanced, x="Sampling_Strategy",color="Sampling_Strategy", measures="Shannon") 
p1 <- p1 + 
  geom_point(aes(color = Sampling_Strategy),
             position = position_jitter(width = 0.15),
             size = 2, alpha = 0.25) +
  geom_boxplot(fill = pal, width = 0.2, outlier.shape = NA,
               alpha = 0.5) + 
   geom_half_violin(position = position_nudge(x = 0.15, y = 0),
                   alpha = 0.25, side = "r") +
  geom_hline(yintercept = mean((p1$data)$value), linetype = 2, color= "grey") +
  guides(colour = guide_legend(nrow = 1)) +
  scale_colour_manual(values = pal) +
  scale_fill_manual(values = pal) +
  stat_compare_means(
    comparisons = list(c("Self", "Professional")),
    aes(label = ..p.signif..), 
    method = "wilcox.test",p.adjust.method = "fdr",
    paired = TRUE,
    label.y = 3.6, 
    label.x.npc = 0.2,
    tip.length = 0.01)+
  theme_biome_utils() + xlab("") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
p1 <- p1 + scale_fill_manual("Sampling Strategy", values = (pal))+  theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text(angle=45, hjust=1, size = 22),
    axis.text.y      = element_text(size = 22),
    strip.text       = element_text(size = 25),  # facet labels
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
  )
p1



#for the observed
ps.sal.prof.obs <- ps.oral %>% ps_filter(Sample_Type =="Saliva")
table(as.data.frame(sample_data(ps.sal.prof.obs))$Sampling_Strategy)

# Convert to data frame if needed
sal_df <- as.data.frame(sample_data(ps.sal.prof.obs))

# Identify samples from each nostril
left_samples <- rownames(sal_df[sal_df$Sampling_Strategy == "Self", ])
right_samples <- rownames(sal_df[sal_df$Sampling_Strategy == "Professional", ])

# Randomly sample 33 Right nostril samples to match the Left
set.seed(123)  # For reproducibility
right_samples_sub <- sample(right_samples, length(left_samples))

# Combine Left samples with the downsampled Right samples
balanced_samples <- c(left_samples, right_samples_sub)

# Subset the phyloseq object to retain only balanced samples
ps.sal.balanced.obs <- prune_samples(balanced_samples, ps.sal.prof.obs)

# Check the new distribution
table(sample_data(ps.sal.balanced.obs)$Sampling_Strategy)

#Observed
p2 <- plot_richness(ps.sal.balanced.obs, x="Sampling_Strategy", color="Sampling_Strategy", measures="Chao1")
p2 <- p2 +
  geom_point(aes(color = Sampling_Strategy),
             position = position_jitter(width = 0.15),
             size = 2, alpha = 0.25) +
  geom_boxplot(fill = pal , width = 0.2, outlier.shape = NA,
               alpha = 0.5) +
   geom_half_violin(position = position_nudge(x = 0.15, y = 0),
                   alpha = 0.25, side = "r") +
  geom_hline(yintercept = mean((p2$data)$value), linetype = 2, color= "grey") +
  scale_colour_manual(values = pal) +
  scale_fill_manual(values = pal) +
  stat_compare_means(comparisons = list(c("Self", "Professional")),
                     aes(label = ..p.signif..),
                     method = "wilcox.test",p.adjust.method = "fdr",
                     paired = TRUE,
                     label.y = 915,
                     label.x.npc = 0.2, tip.length = 0.01)+
  theme_biome_utils() + xlab("") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

p2 <- p2 + scale_colour_manual("Sampling Strategy", values = (pal))
p2

#Simpson
p3 <- plot_richness(ps.sal.balanced, x="Sampling_Strategy", color="Sampling_Strategy", measures="Simpson") 
p3 <- p3 + 
  geom_point(aes(color = Sampling_Strategy),
             position = position_jitter(width = 0.15),
             size = 2, alpha = 0.25) +
  geom_boxplot(fill = pal , width = 0.2, outlier.shape = NA,
               alpha = 0.5) + 
   geom_half_violin(position = position_nudge(x = 0.15, y = 0),
                   alpha = 0.25, side = "r") +
  geom_hline(yintercept = mean((p3$data)$value), linetype = 2, color= "grey") +
  scale_colour_manual(values = pal) +
  scale_fill_manual(values = pal) +
  stat_compare_means(comparisons = list(c("Self", "Professional")),
                     aes(label = ..p.signif..), 
                     method = "wilcox.test",p.adjust.method = "fdr",
                     label.y = c(1),
                     paired=TRUE,
                     label.x.npc = 0.2, tip.length = 0.01)+
  theme_biome_utils() + xlab("") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

p3 <- p3 + scale_colour_manual("Sampling Strategy", values = (pal))+  theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 25),
    axis.text.x      = element_text(angle=45, hjust=1, size = 22),
    axis.text.y      = element_text(size = 22),
    strip.text       = element_text(size = 25),  # facet labels
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
  )
p3


figure_legend <- get_legend(p1)

figure_sal_str <- ggarrange(p1, p3,
                    labels = c(" "),
                    ncol = 2, nrow = 1,
                    common.legend = TRUE, legend = "right", legend.grob = figure_legend)

figure_sal_str


ggsave(
  filename    = "Alpha_div_Saliva_SampStrat.tiff",
  plot        = figure_sal_str,
  device      = "tiff",
  dpi         = 300,
  width       = 10,      # in inches
  height      = 8,      # adjust as needed
  units       = "in",
  compression = "lzw")

```

```{r}

# 1. Agglomerate at Genus level
ps.genus <- tax_glom(ps.rel.oral, taxrank = "Genus", NArm = FALSE)

# 2. Overwrite the taxa_names() with the Genus labels
genus_labels <- tax_table(ps.genus)[, "Genus"]
taxa_names(ps.genus) <- genus_labels

# 3. Sum abundances and pick top-6
abundances <- taxa_sums(ps.genus)
top6 <- sort(abundances, decreasing = TRUE)[1:6]

# 4. Wrap up in a table
top6_df <- data.frame(
  Genus     = names(top6),
  Abundance = as.numeric(top6),
  row.names = NULL
)
top6_df
```

```{r}

# 1. Agglomerate at Genus level
ps.genus <- tax_glom(ps.rel.nose, taxrank = "Genus", NArm = FALSE)

# 2. Overwrite the taxa_names() with the Genus labels
genus_labels <- tax_table(ps.genus)[, "Genus"]
taxa_names(ps.genus) <- genus_labels

# 3. Sum abundances and pick top-6
abundances <- taxa_sums(ps.genus)
top6 <- sort(abundances, decreasing = TRUE)[1:6]

# 4. Wrap up in a table
top6_df <- data.frame(
  Genus     = names(top6),
  Abundance = as.numeric(top6),
  row.names = NULL
)
top6_df
```

# Boxplots: Top 6 Genera per URT Region
```{r, fig.width=14, fig.height=11}
# Convert the phyloseq object to a data frame
df <- psmelt(ps.rel.oral)

# Define your top 5 genera
top_genera <- c("Streptococcus", "Prevotella_7", "Veillonella", "Haemophilus", "Rothia","Actinomyces")

# Filter the data frame for these genera
df_top <- subset(df, Genus %in% top_genera)
df_top$Genus <- factor(df_top$Genus, levels = top_genera)


# Compute common IDs across sample types
common_ids <- Reduce(intersect, split(df_top$ID, df_top$Sample_Type))

# Filter the data for only those IDs
df_common <- df_top %>% filter(ID %in% common_ids)

# Define the comparison: adjust the group names as needed.
# If you have exactly two sample types per facet (e.g., "NPS" and "MLF"):
comp_list <- list(c("OPS", "Saliva"))

# Create faceted boxplot and add significance bracket with stars in the middle
top5_plot_oral <- ggplot(df_common, aes(x = Sample_Type, y = Abundance, fill = Sample_Type)) +
  geom_boxplot() +
  facet_wrap(~ Genus) +
  scale_fill_manual(name = "Sample Type", values = pcoa_colors) +
  theme_bw() +
  labs(x = "", y = "Relative Abundance (%)") +
   theme(
    legend.position = "right",
    strip.text = element_text(size = 16, face = "italic")  # bigger & italic facet labels
  ) +
  stat_compare_means(comparisons = comp_list,
                     method = "wilcox.test", p.adjust.method = "fdr",
                     paired = TRUE, 
                     label = "p.signif", 
                     tip.length = 0.01,
                     size=5, y.position = c(42, 42, 42, 42, 42, 42)   )

top5_plot_oral


ggsave(
  filename    = "Top_6_oral_cavity.tiff",
  plot        = top5_plot_oral,
  device      = "tiff",
  dpi         = 300,
  width       = 11,      # in inches
  height      = 13,      # adjust as needed
  units       = "in",
  compression = "lzw"
)
```


# Boxplots: Top 6 Genera per URT Region
Comparisons of the samples were made on the core taxa and not on top abundant. 
```{r, fig.width=14, fig.height=12}
# Convert the phyloseq object to a data frame
df <- psmelt(ps.rel.nose)

# Define your top 5 genera
top_genera <- c("Corynebacterium", "Staphylococcus","Dolosigranulum" ,"Lawsonella", "Streptococcus","Moraxella")

# Filter the data frame for these genera
df_top <- subset(df, Genus %in% top_genera)
df_top$Genus <- factor(df_top$Genus, levels = top_genera)

# Compute common IDs across sample types
df_top_2 <- df_top %>% 
  filter(Nostril=="Right"|is.na(Nostril))


common_ids <- Reduce(intersect, split(df_top_2$ID, df_top_2$Sample_Type))

# Filter the data for only those IDs
df_common <- df_top_2 %>% filter(ID %in% common_ids)
df_common <- df_common %>%
  mutate(Sample_Type = factor(Sample_Type, levels = c("NPS","MTS","MLF")))
# Define the comparison: adjust the group names as needed.
# If you have exactly two sample types per facet (e.g., "NPS" and "MLF"):
comp_list <- list(c("NPS", "MTS"),
                  c("NPS","MLF"),
                  c("MTS","MLF"))
# Create boxplots with sample type on x-axis and relative abundance on y-axis, faceted by Genus
top5_plot_nasal <- ggplot(df_common, aes(x = Sample_Type, y = Abundance, fill=Sample_Type)) +
  geom_boxplot() +
  facet_wrap(~ Genus, scales = "free_y") +
  scale_fill_manual(name = "Sample Type", values = pcoa_colors) +
  theme_bw() +
  labs(x = "", y = "Relative Abundance (%)") +
  theme(
    legend.position = "right",
    strip.text = element_text(size = 16, face = "italic")  # bigger & italic facet labels
  ) +
  stat_compare_means(
    comparisons = comp_list,
    method = "wilcox.test", p.adjust.method = "fdr",
    paired = TRUE,
    label = "p.signif",
    tip.length = 0.01,
    size = 5
  )

top5_plot_nasal




ggsave(
  filename    = "Top_6_nasal_cavity.tiff",
  plot        = top5_plot_nasal,
  device      = "tiff",
  dpi         = 300,
  width       = 11,      # in inches
  height      = 13,      # adjust as needed
  units       = "in",
  compression = "lzw"
)
```



```{r, fig.width=15, fig.height=18}
library(grid)
library(gridExtra)
# combined_plot <- grid.arrange(top5_plot_oral, top5_plot_nasal, nrow = 2)
top5_plot_nasal <- top5_plot_nasal+
    theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 28),
    axis.text.x      = element_text( size = 25),
    axis.text.y      = element_text(size = 25),
    strip.text       = element_text(size = 23, face = "italic"),  # facet labels
    legend.title     = element_text(size = 26),
    legend.text      = element_text(size = 25)
  )

top5_plot_oral <- top5_plot_oral+
   theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 28),
    axis.text.x      = element_text( size = 25),
    axis.text.y      = element_text(size = 25),
    strip.text       = element_text(size = 23, face = "italic"),  # facet labels
    legend.title     = element_text(size = 26),
    legend.text      = element_text(size = 25))

nasal_label <- textGrob("Nasal Cavity",
                        rot = 90,
                        x   = unit(0.8, "npc"),   # move 80% of the way across its cell
                        gp  = gpar(fontsize =28, fontface = "bold"))
oral_label  <- textGrob("Oral Cavity",
                        rot = 90,
                        x   = unit(0.8, "npc"),
                        gp  = gpar(fontsize = 28, fontface = "bold"))

# 2) Tweak the widths so the label column is much narrower:
row1 <- arrangeGrob(nasal_label, top5_plot_nasal, ncol = 2,
                    widths = c(0.2, 0.8))   # 20% for the label, 80% for the plot
row2 <- arrangeGrob(oral_label,  top5_plot_oral,  ncol = 2,
                    widths = c(0.2, 0.8))

# Combine rows into one figure
combined_plot<- grid.arrange(row1, row2, nrow = 2)


ggsave(
  filename    = "Top_6_with_regions2.tiff",
  plot        = combined_plot,
  device      = "tiff",
  dpi         = 300,
  width       = 18,      # in inches
  height      = 23,      # adjust as needed
  units       = "in",
  compression = "lzw"
)
```


# Top 10 most abudant
```{r}
get_top10_by_genus <- function(ps, type, sampvar = "Sample_Type"){
  # pull sample data into a data.frame
  sd <- as(sample_data(ps), "data.frame")
  # subset sample names where that column == type
  samples_to_keep <- rownames(sd)[ sd[[sampvar]] == type ]
  # prune the phyloseq object
  ps_sub <- prune_samples(samples_to_keep, ps)
  
  # get ASV sums
  asv_sums <- taxa_sums(ps_sub)
  # get genus labels
  genera  <- as.character(tax_table(ps_sub)[, "Genus"])
  
  # drop unclassified
  keep    <- !is.na(genera) & genera != ""
  asv_sums <- asv_sums[keep]
  genera   <- genera[keep]
  
  # sum by genus
  genus_totals <- tapply(asv_sums, genera, sum)
  
  # pick top 10
  top10 <- sort(genus_totals, decreasing = TRUE)[1:10]
  data.frame(Genus = names(top10),
             Abundance = as.numeric(top10),
             row.names = NULL)
}

# 
top10_sal <- get_top10_by_genus(ps.rel.oral, "Saliva", sampvar = "Sample_Type")
top10_op  <- get_top10_by_genus(ps.rel.oral, "OPS",    sampvar = "Sample_Type")

top10_sal
top10_op
```


## Boxplots: Top 6 Genera per URT Region (colored by genera)
```{r, fig.width=13, fig.height=10}
# Convert the phyloseq object to a data frame

# Define top genera

# Convert the phyloseq object to a data frame
df <- psmelt(ps.rel.oral)

top_genera_op <- c("Streptococcus", "Prevotella_7",  "Veillonella", "Actinomyces","Leptotrichia","Neisseria")
top_genera_sal <- c("Streptococcus",  "Haemophilus", "Rothia","Neisseria",  "Veillonella","Actinomyces")


# Filter the data frame for these genera
df_op <- subset(df, sample_type == "OP")
df_op <- subset(df_op, Genus %in% top_genera_op)
df_op$Genus <- factor(df_op$Genus, levels = top_genera_op)

df_sal <- subset(df, sample_type == "Saliva")
df_sal <- subset(df_sal, Genus %in% top_genera_sal)
df_sal$Genus <- factor(df_sal$Genus, levels = top_genera_sal)

# Create boxplots with sample type on x-axis and relative abundance on y-axis, faceted by Genus
top5_plot_oral_g <- ggplot(df_top, aes(x = Genus, y = Abundance, fill=Genus)) +
  geom_boxplot() +
  facet_wrap(~ Sample_Type) +
  scale_fill_manual(name = "Genus", values = top_genera_cols)+
  theme_bw() +
  labs(x = "", y = "Relative Abundance (%)")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  theme(legend.position = "right")

top5_plot_oral_g

## OP


# Create boxplots with sample type on x-axis and relative abundance on y-axis, faceted by Genus
top6_op <- ggplot(df_op, aes(x = Genus, y = Abundance, fill=Genus)) +
  geom_boxplot() +
  facet_wrap(~ Sample_Type) +
  scale_fill_manual(name = "Genus", values = top_genera_cols)+
  theme_bw() +
  labs(x = "", y = "Relative Abundance (%)")+
  theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 25),
     axis.text.x = element_blank(),
    axis.ticks.x     = element_blank(), 
    axis.text.y      = element_text(size = 22),
    strip.text       = element_text(size = 25),  # facet labels
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
  )+
  theme(legend.position = "right") +ylim(0,65)

top6_op


## Sal
# Filter the data frame for these genera
df_top_Sal <- subset(df_sal, Genus %in% top_genera_sal)
df_top_Sal$Genus <- factor(df_top_Sal$Genus, levels = top_genera_sal)

# Create boxplots with sample type on x-axis and relative abundance on y-axis, faceted by Genus
top6_sal<- ggplot(df_top_Sal, aes(x = Genus, y = Abundance, fill=Genus)) +
  geom_boxplot() +
  facet_wrap(~ Sample_Type) +
  scale_fill_manual(name = "Genus", values = top_genera_cols)+
  theme_bw() +
  labs(x = "", y = "Relative Abundance (%)")+
  theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 25),
    axis.text.x = element_blank(),
    axis.ticks.x     = element_blank(), 
    axis.text.y      = element_text(size = 22),
    strip.text       = element_text(size = 25),  # facet labels
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
  )+
  theme(legend.position = "right")

top6_sal

# Combine rows into one figure


combined_plot_opsal<- grid.arrange(top6_op, top6_sal, nrow = 1)

ggsave(
  filename    = "Top_6_OPSnSaliva.tiff",
  plot        = combined_plot_opsal,
  device      = "tiff",
  dpi         = 300,
  width       = 13,      # in inches
  height      = 17,      # adjust as needed
  units       = "in",
  compression = "lzw"
)
```


```{r}
top_genera_all <- c("Streptococcus", "Prevotella_7", "Veillonella", "Actinomyces", 
                    "Leptotrichia", "Neisseria", "Haemophilus", "Rothia")
df_op$Genus  <- factor(df_op$Genus, levels = top_genera_all)
df_sal$Genus <- factor(df_sal$Genus, levels = top_genera_all)

# Add a column to identify sample origin
df_op$Group <- "OP"
df_sal$Group <- "Saliva"

# Combine both into one data frame
df_combined <- rbind(df_op, df_sal)

# Make sure Genus factor has all desired levels
df_combined$Genus <- factor(df_combined$Genus, levels = top_genera_all)
top_plot_combined <- ggplot(df_combined, aes(x = Genus, y = Abundance, fill = Genus)) +
  geom_boxplot() +
  facet_wrap(~ Group) +  # One panel for OP, one for Saliva
  scale_fill_manual(name = "Genus", values = top_genera_cols) +
  theme_bw(base_size = 25) +
  labs(x = "", y = "Relative Abundance (%)") +
  theme(
    axis.title       = element_text(size = 25),
    axis.text.x      = element_blank(),
    axis.ticks.x     = element_blank(),
    axis.text.y      = element_text(size = 22),
    strip.text       = element_text(size = 25),
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
    legend.position  = "right"
  )

top_plot_combined

ggsave("Top_6_OPSnSaliva_combined.tiff", top_plot_combined,
       device = "tiff", dpi = 300,
       width = 13, height = 10, units = "in", compression = "lzw")
```



## Boxplots: Top 6 Genera per URT Region (colored by genera)
```{r, fig.width=14, fig.height=10}
# Convert the phyloseq object to a data frame
df <- psmelt(ps.rel.nose)

# Define your top 5 genera
top_genera <- c("Corynebacterium", "Staphylococcus","Dolosigranulum" ,"Lawsonella", "Streptococcus","Moraxella")

# Filter the data frame for these genera
df_top <- subset(df, Genus %in% top_genera)
df_top$Genus <- factor(df_top$Genus, levels = top_genera)

# Create boxplots with sample type on x-axis and relative abundance on y-axis, faceted by Genus
top5_plot_nasal_g <- ggplot(df_top, aes(x = Genus, y = Abundance, fill=Genus)) +
  geom_boxplot() +
  facet_wrap(~ Sample_Type) +
  scale_fill_manual(name = "Genus", values = top_genera_cols)+
  theme_bw() +
  labs(x = "", y = "Relative Abundance (%)")+
  theme_bw(base_size=25)+theme(
    axis.title       = element_text(size = 25),
    axis.text.x = element_blank(),
    axis.ticks.x     = element_blank(), 
    axis.text.y      = element_text(size = 22),
    strip.text       = element_text(size = 25),  # facet labels
    legend.title     = element_text(size = 25),
    legend.text      = element_text(size = 24),
  )+
  theme(legend.position = "right")

top5_plot_nasal_g


ggsave(
  filename    = "Top_6_Nose_Samples.tiff",
  plot        = top5_plot_nasal_g,
  device      = "tiff",
  dpi         = 300,
  width       = 13,      # in inches
  height      = 17,      # adjust as needed
  units       = "in",
  compression = "lzw"
)
```



```{r, fig.width=11, fig.height=14}

library(gridExtra)
library(grid)
# Create left side labels as grobs (rotated 90 degrees)
nasal_label <- textGrob("Nasal Cavity", rot = 90, gp = gpar(fontsize = 28, fontface = "bold"))
oral_label  <- textGrob("Oral Cavity", rot = 90, gp = gpar(fontsize = 28, fontface = "bold"))

# Arrange each row with two columns: label and plot
row1 <- arrangeGrob(nasal_label, top5_plot_nasal_g, ncol = 2, widths = c(0.1, 0.9))
row2 <- arrangeGrob(oral_label, top_plot_combined, ncol = 2, widths = c(0.1, 0.9))

# Combine rows into one figure
combined_plot_g <- grid.arrange(row1, row2, nrow = 2)
ggsave(
  filename    = "Top_6_Genera_Niches.tiff",
  plot        = combined_plot_g,
  device      = "tiff",
  dpi         = 300,
  width       = 13,      # in inches
  height      = 17,      # adjust as needed
  units       = "in",
  compression = "lzw"
)
```

