---
title: "MOFA analysis"
output: 
  html_document:
    toc: true            # Enable Table of Contents
    toc_float: true
    code_folding: "hide"      
date: "09-09-2025"
---

```{r Load Packages, include=FALSE}
# Function to install/load packages:

load_packages <- function(packages) {
  sapply(packages, function(pkg) {
    if (!require(pkg, character.only = TRUE)) {
      install.packages(pkg, dependencies = TRUE)
      library(pkg, character.only = TRUE)
    }
  })
}

packages <- c("purrr", "dplyr", "tidyr", "knitr", "magrittr", "reshape2", 
              "tidyverse", "ggplot2", "RColorBrewer", "stringr", 
              "rstatix", "ggpubr", "ggsignif", "FSA",  
              "readxl", "viridis", "patchwork",   
              "ggfortify", "factoextra", "pheatmap",
              "cowplot","coin","psych","cowplot",
              "phyloseq","microbiome","plyr","hrbrthemes",
              "RColorBrewer","readr", "decontam", "microbiomeutilities",
              "DECIPHER", "ape","phangorn", "tibble","utils","MOFA2","microViz",
              "vegan","mixOmics","data.table")

load_packages(packages)
library(MOFA2)

```


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```



#  Load datasets
```{r load ps, message=TRUE}

# Load phyloseq object (microbiome):
ps <- readRDS("07-ps.for.multiomics.rds")

ps.df <- data.frame(sample_data(ps)) #make dataframe 

table(ps.df$Sample_Type) #count number of sample types
table(ps.df$Sampling_Method) #count number of sampling methods

ps #print phyloseq object (7171 taxa in 308 samples)
#-----------------------------------------------------

# Import only LegendPlex data:
lgnd <- read_excel("02-LegendPlex_data.xlsx")
library(dplyr)
lgnd %>%
  distinct(SampleID, Sample_Type, Analytes) %>%        # one row per sample
  dplyr::count(Analytes, Sample_Type, name = "n_samples")
#-----------------------------------------------------

# Import total Abs data:
total_abs <- read_excel("01-Luminex_Total_Abs_Data.xlsx")
total_abs %>%
  distinct(Description, Sample_Type, Analyte) %>%        # one row per sample
  dplyr::count(Analyte, Sample_Type, name = "n_samples")


#-----------------------------------------------------

# Import RespiFinder (multiplex PCR) data:

respifinder <- read_excel("04-antigen_specific_MIA_data.xlsx")
respifinder <- respifinder %>%
  mutate(
    Pathogen_group = case_when(
      Result %in% c("SARS-CoV2", "Rhino/Entero", "Boca", "Influenza A") ~ "Virus",
      Result %in% c("C.Pneu", "B.Pert") ~ "Bacterium",
      is.na(Result) ~ "Negative"
    )
  )

resp_flags <- respifinder %>%
  distinct(SampleID, Pathogen_group) %>%   # one row per sample × group
  mutate(
    Virus_pos = Pathogen_group == "Virus",
    Bact_pos  = Pathogen_group == "Bacterium",
    Any_pos   = Pathogen_group != "Negative"
  ) 
```



# Data cleanup 
We will only use Professionally collected samples for the integrated analysis. 

```{r data-cleanup}

ps.prof <- phyloseq::subset_samples(ps, Sampling_Strategy == "Professional")
ps.prof.df <- data.frame(sample_data(ps.prof))

lgnd.prof <- lgnd %>% filter( Sampling_Strategy == "Professional")
lgnd.prof <- lgnd.prof %>% mutate(log10_conc = log10(concentration.est))
# lgnd.prof <- lgnd.prof %>% mutate(log_conc = log(concentration.est))
lgnd.prof <- lgnd.prof %>%
  mutate(Sample_Type = recode(Sample_Type, "Mid-turbinate swab" = "MTS"))
lgnd.prof <- lgnd.prof %>%
  mutate(Sample_Type = recode(Sample_Type, "Nasopharyngeal swab" = "NPS"))
lgnd.prof <- lgnd.prof %>%
  mutate(Sample_Type = recode(Sample_Type, "Oropharyngeal swab" = "OPS"))

total_abs.prof <- total_abs %>% filter( Sampling_Strategy == "Professional")

# Function to replace "OOR <" values with half of the minimum or half of LLOQ if all values are NA
replace_missing_with_min <- function(x, lloq) {
  min_val <- min(x, na.rm = TRUE)
  if (is.finite(min_val)) {
    return((1/2) * min_val)
  } else {
    return((1/2) * lloq)  # Use half of LLOQ if all values are NA
  }
}

# Function to replace "OOR >" values with the maximum x2 concentration in range
replace_missing_with_max <- function(x) {
  max_val <- max(x, na.rm = TRUE)
  if (is.finite(max_val)) {
    return(max_val*2)
  } else {
    return(NA)  # If all values are NA, return NA or another desired fallback
  }
}

# Apply the functions per isotype and sample type
total_abs.prof <- total_abs.prof %>%
  group_by(Analyte, Sample_Type) %>%
  mutate(
    concentration = case_when(
      is.na(Conc_in_Range) & Notes == "OOR <" ~ replace_missing_with_min(Conc_in_Range, LLOQ),
      is.na(Conc_in_Range) & Notes == "OOR >" ~ replace_missing_with_max(Conc_in_Range),
      TRUE ~ Conc_in_Range
    )
  ) %>%
  ungroup()

#  Calculate log10 of the new concentration column
total_abs.prof <- total_abs.prof %>%
  mutate(log10_conc = log10(concentration))
```


# Build metadata
```{r}
# 1. Build metadata starting from LegendPlex professional
resp_flags_1 <- resp_flags %>% subset(select=c(SampleID, Virus_pos, Bact_pos, Any_pos))
meta <- lgnd.prof %>%
  distinct(SampleID, Sample_Type, ID, Biological_Sex, Age_Range, Nostril) %>%
  left_join(resp_flags_1, by = "SampleID")

# check
dim(meta)
table(meta$Sample_Type)

  # MLF    MTS    NPS    OPS Saliva 
  #   88     88     44     44     40 

```

```{r}
total_abs.prof <- total_abs.prof %>%
  dplyr::rename(SampleID = Description)

```

# IDs check
```{r}

ids_microbe <- sample_names(ps.prof)   # from phyloseq
ids_cytokine <- unique(lgnd.prof$SampleID)
ids_igs <- unique(total_abs.prof$SampleID)

# overlap with meta
table(meta$SampleID %in% ids_microbe)
table(meta$SampleID %in% ids_cytokine)
table(meta$SampleID %in% ids_igs)

```

```{r}
setdiff(meta$SampleID, ids_microbe)   # in meta but not microbiome
setdiff(meta$SampleID, ids_cytokine)  # in meta but not cytokines
setdiff(meta$SampleID, ids_igs)       # in meta but not Igs

```

# Find the shared IDs
```{r}
integration_ids <- Reduce(intersect, list(
  meta$SampleID,
  ids_microbe,
  ids_cytokine,
  ids_igs
))

length(integration_ids)  # should be 264
```


# Subset to these IDs (meta)
```{r}
meta_sub <- meta %>% 
  filter(SampleID %in% integration_ids)

nrow(meta_sub)   #  264

table(meta_sub$Sample_Type)

# MLF    MTS    NPS    OPS Saliva 
#     69     80     36     44     35 
meta_sub <- meta_sub %>%
  mutate(SampleID_site = paste(ID, Sample_Type, sep="_"))


```
## microbiome
```{r}
ps.sub <- phyloseq::subset_samples(ps.prof, sample_identifier %in% integration_ids)

# sanity check
nsamples(ps.sub)   # 264
```

# legendplex
```{r}
lgnd.sub <- lgnd.prof %>%
  filter(SampleID %in% integration_ids) %>%
  subset(select=c(SampleID, Analytes, log10_conc))

# wide format (rows = samples, cols = analytes)
cyto.mat <- tidyr::pivot_wider(lgnd.sub,
  names_from = Analytes,
  values_from = log10_conc
)

# reorder rows to match meta_sub
cyto.mat <- cyto.mat[match(meta_sub$SampleID, cyto.mat$SampleID), ]
```

# total Igs
```{r}
igs.sub <- total_abs.prof %>%
  filter(SampleID %in% integration_ids) %>%
  subset(select=c(SampleID, Analyte, log10_conc))

igs.mat <- tidyr::pivot_wider(igs.sub,
  names_from = Analyte,
  values_from = log10_conc
) 

# reorder rows to match meta_sub
igs.mat <- igs.mat[match(meta_sub$SampleID, igs.mat$SampleID), ]
```

# collapse microbiome & CLRS
```{r}
library(compositions)
library(dplyr)

# 1. Aggregate microbiome to Genus level
ps.prof.genus <- tax_glom(ps.sub, "Genus", NArm = TRUE)

# 2. Relative abundance (%)
ps.prof.rel <- transform_sample_counts(ps.prof.genus, function(OTU) OTU/sum(OTU) * 100)

# 3. Convert to matrix
microbe_data <- as.data.frame(otu_table(ps.prof.rel))
if(taxa_are_rows(ps.prof.rel)) {
  microbe_data <- t(microbe_data)
}

# 4. Replace column names with Genus names
genus_names <- as.character(tax_table(ps.prof.rel)[, "Genus"])
colnames(microbe_data) <- genus_names

# 5. Filter genera by prevalence (>10%)
prevalence_filter <- function(x, threshold = 0.1) mean(x > 0) > threshold
filtered_microbe <- microbe_data[, apply(microbe_data, 2, prevalence_filter)]

# 6. CLR transform
microbe_data_clr <- as.data.frame(clr(filtered_microbe))

# 7. Add SampleID
microbe_data_clr$SampleID <- rownames(microbe_data_clr)

# 8. Collapse left/right nostrils → subject × site
microbe_data_clr <- microbe_data_clr %>%
  left_join(meta_sub %>% subset(select=c(SampleID, SampleID_site)),
            by = "SampleID") %>%
  group_by(SampleID_site) %>%
  summarise_at(vars(-SampleID), mean, na.rm=TRUE) %>%
  ungroup()

# 9. Final matrix for MOFA
X_microbe <- scale(as.matrix(microbe_data_clr[,-1]))
rownames(X_microbe) <- microbe_data_clr$SampleID_site

```


# collapse legendplex
```{r}
# start from your long lgnd.sub (SampleID, Analytes, log10_conc)
cyto.mat.collapsed <- lgnd.sub %>%
  dplyr::left_join(meta_sub %>% dplyr::select(SampleID, SampleID_site), by = "SampleID") %>%
  dplyr::group_by(SampleID_site, Analytes) %>%
  dplyr::summarise(log10_conc = mean(log10_conc, na.rm = TRUE)) %>%
  dplyr::ungroup() %>%
  tidyr::pivot_wider(names_from = Analytes, values_from = log10_conc)

# matrix + scaling
X_cytokine <- as.data.frame(cyto.mat.collapsed)
rownames(X_cytokine) <- X_cytokine$SampleID_site
X_cytokine <- as.matrix(X_cytokine[ , -1, drop = FALSE])
X_cytokine <- scale(X_cytokine)
```

# collapse total antibodies
```{r}
igs.mat.collapsed <- igs.sub %>%
  dplyr::left_join(meta_sub %>% dplyr::select(SampleID, SampleID_site), by = "SampleID") %>%
  dplyr::group_by(SampleID_site, Analyte) %>%
  dplyr::summarise(log10_conc = mean(log10_conc, na.rm = TRUE)) %>%
  dplyr::ungroup() %>%
  tidyr::pivot_wider(names_from = Analyte, values_from = log10_conc)



X_igs <- as.data.frame(igs.mat.collapsed)
rownames(X_igs) <- X_igs$SampleID_site
X_igs <- as.matrix(X_igs[ , -1, drop = FALSE])
X_igs <- scale(X_igs)
```

```{r}
common_ids <- Reduce(intersect, list(rownames(X_microbe), rownames(X_cytokine), rownames(X_igs)))

# Build collapsed metadata
meta_collapsed <- meta_sub %>%
  dplyr::distinct(ID, Sample_Type, Biological_Sex, Age_Range, 
                  Virus_pos, Bact_pos, Any_pos, SampleID_site) %>%
  dplyr::filter(SampleID_site %in% common_ids) %>%
  dplyr::arrange(match(SampleID_site, common_ids))

meta_collapsed <- meta_collapsed[match(rownames(X_microbe), meta_collapsed$SampleID_site), ]

```


# check alignment
```{r}

X_microbe  <- X_microbe [common_ids, , drop = FALSE]
X_cytokine <- X_cytokine[common_ids, , drop = FALSE]
X_igs      <- X_igs     [common_ids, , drop = FALSE]

meta_collapsed <- meta_sub %>%
  dplyr::distinct(ID, Sample_Type, Biological_Sex, Age_Range, Virus_pos, Bact_pos, Any_pos, SampleID_site) %>%
  dplyr::filter(SampleID_site %in% common_ids) %>%
  dplyr::arrange(match(SampleID_site, common_ids))

stopifnot(identical(rownames(X_microbe), rownames(X_cytokine)))
stopifnot(identical(rownames(X_microbe), rownames(X_igs)))
meta_collapsed <- meta_collapsed[match(rownames(X_microbe), meta_collapsed$SampleID_site), ]

# now check again
stopifnot(identical(meta_collapsed$SampleID_site, rownames(X_microbe)))
# setdiff(rownames(X_microbe), meta_collapsed$SampleID_site)
# setdiff(meta_collapsed$SampleID_site, rownames(X_microbe))
```

```{r}
table(meta_collapsed$Sample_Type)
table_input <- table(meta_collapsed$Sample_Type)
table_input <- as.data.frame(table_input)

clean_names <- function(x) {
  x <- gsub("α", "alpha", x)
  x <- gsub("γ", "gamma", x)
  x <- gsub("λ", "lambda", x)
  x <- gsub("[^A-Za-z0-9_]", "_", x)  # replace any other weird chars with underscore
  x
}

colnames(X_cytokine) <- clean_names(colnames(X_cytokine))

```


# Build MOFA
```{r}
X_igs <- X_igs[,-4]
views <- list(
  Microbiome = t(X_microbe),   # features x samples
  Cytokines  = t(X_cytokine),
  Antibodies = t(X_igs)
)

mofa <- create_mofa(views)
mofa

do<- plot_data_overview(mofa)

data_opts <- get_default_data_options(mofa)
data_opts

model_opts <- get_default_model_options(mofa)
model_opts$spikeslab_weights <- TRUE

#model_opts$num_factors <- 15

model_opts
# Training options
train_opts <- get_default_training_options(mofa)
train_opts
train_opts$verbose <- TRUE     # prints progress

ggsave(
  filename = "mofa_overview.tiff",
  plot     = do,
  device   = "tiff",
  dpi      = 300,
  width    = 10,    # adjust as needed
  height   = 7,
  units    = "in",
  compression = "lzw"
)
```


```{r}
MOFAobject <- prepare_mofa(mofa,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)

slotNames(MOFAobject)
names(MOFAobject@data)

meta_mofa <-meta_collapsed %>% mutate(sample=SampleID_site)
samples_metadata(MOFAobject) <- meta_mofa

mofa_fit <- run_mofa(MOFAobject, use_basilisk = TRUE)

# correlation between factors?
#Not correlated
plot_factor_cor(mofa_fit)

```



# Variance decomposition analysis
The most important insight that MOFA generates is the **variance decomposition analysis**. This plot shows the percentage of variance explained by each factor across each data modality (and group, if provided). It summarises the sources of variation from a complex heterogeneous data set in a single figure.
```{r}
vp2<- plot_variance_explained(mofa_fit)+ theme(
  axis.text.x = element_text(size = 16),
  axis.text.y = element_text(size = 16),
  legend.text = element_text(size = 16),
    legend.title = element_text(size = 16))

vp2
ggsave(
  filename = "mofa_vp2.tiff",
  plot     = vp2,
  device   = "tiff",
  dpi      = 300,
  width    = 10,    # adjust as needed
  height   = 7,
  units    = "in",
  compression = "lzw"
)
```

```{r}
# variance explained by factor × view
var_exp <- get_variance_explained(mofa_fit)

# total per factor across views

# plot
plot_variance_explained(mofa_fit)

ggsave(
  filename = "mofa_vp.tiff",
  plot     = vp,
  device   = "tiff",
  dpi      = 300,
  width    = 10,    # adjust as needed
  height   = 7,
  units    = "in",
  compression = "lzw"
)
```





# Total variance explained per view
A reasonable question is whether the model is providing a good fit to the data. For this we can plot the total variance explained (using all factors). The resulting values will depend on the nature of the data set, the number of samples, the number of factors, etc. Some general guidelines:

Noisy data sets with strong non-linearities will result in small amounts of variance explained (<10%).
The higher the number of samples the smaller the total variance explained.
The higher the number of factors, the higher the total variance explained.
MOFA is a linear and sparse model. This is helpful to prevent overfitting, but it will never explain 100% of the variance, even if using a lot of Factors.

In this data set, using only K=15 factors the model explains up to ~56.8% of the variation in the Total Antibodies concentrations and ~47% of the variation in the immune marker data. For microbiome that's ~24%.
```{r}
vp <- plot_variance_explained(mofa_fit, plot_total = T)[[2]]
vp
ggsave(
  filename = "mofa_vp.tiff",
  plot     = vp,
  device   = "tiff",
  dpi      = 300,
  width    = 10,    # adjust as needed
  height   = 7,
  units    = "in",
  compression = "lzw"
)
```

# Association Analysis
Factor 2 and to a lesser extent Factor 4 have an association with Sample Type.
```{r}
ex_data <- correlate_factors_with_covariates(
  mofa_fit,
  covariates = c("Sample_Type","Biological_Sex","Age_Range","Virus_pos","Bact_pos","Any_pos"),
  plot = "log_pval",
  return_data = TRUE
)

head(ex_data)

# Step 2: pivot to long format
ex_long <- as.data.frame(ex_data) %>%
  rownames_to_column("Factor") %>%
  pivot_longer(-Factor, names_to = "Covariate", values_to = "log_pval")

ex_plot <- ggplot(ex_long, aes(x = Covariate, y = Factor, fill = log_pval)) +
  geom_tile(color = "white") +
  scale_fill_distiller(palette = "Greys", direction = 1) +
  labs(x = NULL, y = NULL, fill = "-log10(p)") +
  theme_minimal(base_size = 16) +
  theme(
    axis.text.x  = element_text(size = 16, angle = 45, hjust = 1),
    axis.text.y  = element_text(size = 16),
    legend.text  = element_text(size = 16),
    legend.title = element_text(size = 16, face = "bold")
  )

ex_plot
# # Save the Factor 2 plot
ggsave(
  filename = "mofa_associations_analysis.tiff",
  plot     = ex_plot,
  device   = "tiff",
  dpi      = 300,
  width    = 10,    # adjust as needed
  height   = 7,
  units    = "in",
  compression = "lzw"
)


```



# Plot Factor Values
**How do we interpret the factor values?**
Each factor captures a different source of variability in the data. Mathematically, each Factor is defined by a linear combination of the input features. As the data is centered prior to running MOFA, each Factor ordinates cells along a one-dimensional axis that is centered at zero. Samples with different signs manifest opposite phenotypes along the inferred axis of variation, with higher absolute value indicating a stronger effect. Note that the interpretation of MOFA factors is analogous to the interpretation of the principal components in PCA.

```{r}
plot_factors(mofa_fit, factors=1, color_by="Sample_Type")
plot_factors(mofa_fit, factors=2, color_by="Sample_Type")
plot_factors(mofa_fit, factors=3, color_by="Sample_Type")


```


# Plot Feature Weights
**How do we interpret the weights?**
The weights provide a score for each feature on each factor. Features with no association with the corresponding factor are expected to have values close to zero, whereas features with strong association with the factor are expected to have large absolute values. The sign of the weights indicates the direction of the effect: a positive weights indicates that the feature has higher levels in the cells with positive factor values, and vice-versa.

By looking at the variance explained plot, we saw that Factor 2 captures variation in all data modalities. We do see clear separation in microbiome especially but moderate in the other two as well.
```{r}
plot_weights(mofa_fit,
 view = "Microbiome",
 factor = 2,
 nfeatures = 10,     # Top number of features to highlight
 scale = T           # Scale weights from -1 to 1
)


plot_weights(mofa_fit,
 view = "Cytokines",
 factor = 2,
 nfeatures = 10,     # Top number of features to highlight
 scale = T           # Scale weights from -1 to 1
)


plot_weights(mofa_fit,
 view = "Antibodies",
 factor = 2,
 nfeatures = 10,     # Top number of features to highlight
 scale = T           # Scale weights from -1 to 1
)
```

```{r}
plot_top_weights(mofa_fit,
 view = "Microbiome",
 factor = 2,
 nfeatures = 10,     # Top number of features to highlight
 scale = T           # Scale weights from -1 to 1
)

plot_top_weights(mofa_fit,
 view = "Cytokines",
 factor = 2,
 nfeatures = 10,     # Top number of features to highlight
 scale = T           # Scale weights from -1 to 1
)


plot_top_weights(mofa_fit,
 view = "Antibodies",
 factor = 2,
 nfeatures = 10,     # Top number of features to highlight
 scale = T           # Scale weights from -1 to 1
)
```
IFN_gamma has a positve weight. This means that samples with positive Factor 2 values have IFN_gamma whereas samples with negative Factor 2 values do not have the IFN_gamma. 

```{r}
# my palette
custom_colors <- c(
  "NPS"    = "#A2D2DF", 
  "MTS"    = "bisque3",       
  "MLF"    = "#FFBD73",
  "OPS"    = "#BC7C7C",
  "Saliva" = "#8ABFA3"
)

# factor plot
p <- plot_factors(
  mofa_fit, factors = c(2, 1), color_by = "Sample_Type",
  dot_size = 2.5, show_missing = TRUE
)

 
# add our custom colors
p <- p + scale_color_manual(values = custom_colors, name = "Sample Type",
                     breaks = names(custom_colors), drop = FALSE) +
  scale_fill_manual(values = custom_colors, name = "Sample Type",
                    breaks = names(custom_colors), drop = FALSE) + theme(
  axis.text.x = element_text(size = 16),
  axis.text.y = element_text(size = 16),
  legend.text = element_text(size = 16),axis.title   = element_text(size = 18, face = "bold"),
    legend.title = element_text(size = 16))

p

ggsave("mofa_Factor2_vs_Factor1.tiff",
       plot = p,
       device = "tiff",
       width = 10, height = 6, units = "in",
       dpi = 600, compression = "lzw")
```

```{r}
p <- plot_factors(
  mofa_fit, factors = c(2, 1), color_by = "Sample_Type",
  dot_size = 2.5, show_missing = TRUE
)

p +
  scale_color_manual(values = custom_colors, name = "Sample Type",
                     breaks = names(custom_colors), drop = FALSE) +
  scale_fill_manual(values = custom_colors, name = "Sample Type",
                    breaks = names(custom_colors), drop = FALSE)
```




```{r}
#Extract weights for Factor2 in each view
w_micro <- get_weights(mofa_fit, views = "Microbiome", factors = 2)[[1]]
w_cyto  <- get_weights(mofa_fit, views = "Cytokines", factors = 2)[[1]]
w_igs   <- get_weights(mofa_fit, views = "Antibodies", factors = 2)[[1]]

# Convert to tibble
w_micro <- tibble(feature = rownames(w_micro), value = w_micro[,1], factor = "Factor2")
w_cyto  <- tibble(feature = rownames(w_cyto),  value = w_cyto[,1], factor = "Factor2")
w_igs   <- tibble(feature = rownames(w_igs),   value = w_igs[,1], factor = "Factor2")

# Top N by absolute weight
topN <- 10
w_micro <- w_micro %>% slice_max(order_by = abs(value), n = topN)
w_cyto  <- w_cyto  %>% slice_max(order_by = abs(value), n = topN)
w_igs   <- w_igs   %>% slice_max(order_by = abs(value), n = topN)

# Combine
W2 <- bind_rows(
  mutate(w_micro, View = "Microbiome"),
  mutate(w_cyto,  View = "Cytokines"),
  mutate(w_igs,   View = "Antibodies")
)

# Plot
domain_colors <- c(
  "Microbiome" = "#3d3a37",   # dark grey
  "Cytokines"  = "#7f0000",   # dark red
  "Antibodies" = "#1f4e79"    # navy
)

ggplot(W2, aes(x = reorder(feature, value), y = value, fill = View)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = domain_colors, name = "Domain") +
  labs(x = "", y = "Loading (Factor 2)", title = "Cross-modal features driving Factor 2") +
  theme_minimal(base_size = 14)


library(ggtext)

# Keep feature for ordering, add feature_label for display
W2 <- W2 %>%
  mutate(feature_label = ifelse(View == "Microbiome",
                                paste0("*", feature, "*"),
                                feature))

ggplot(W2, aes(x = reorder(feature, value), y = value, fill = View)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = domain_colors, name = "Domain") +
  labs(x = "", y = "Loading (Factor 2)", title = "Cross-modal features driving Factor 2") +
  theme_bw(base_size = 16) +
  theme(axis.text.y = element_markdown(size = 16), axis.text.x = element_text(size = 16),
  legend.text = element_text(size = 16),axis.title   = element_text(size = 16),
    legend.title = element_text(size = 16),
  plot.title  = element_text(size = 18, face="bold") )+
  scale_x_discrete(labels = setNames(W2$feature_label, W2$feature))


# Save the Factor 2 plot
ggsave(
  filename = "Factor2_top_featuresv2.tiff",
  plot     = last_plot(),
  device   = "tiff",
  dpi      = 300,
  width    = 10,    # adjust as needed
  height   = 7,
  units    = "in",
  compression = "lzw"
)
```
Positive values  -> oral sites (saliva or oropharynx) \n
Negative values -> nasal sites (nasopharynx, MLF or MTS)

