---
title: "RDA"
output: html_document
date: "2025-09-18"
---


```{r Load Packages, include=FALSE}
# Function to install/load packages:

load_packages <- function(packages) {
  sapply(packages, function(pkg) {
    if (!require(pkg, character.only = TRUE)) {
      install.packages(pkg, dependencies = TRUE)
      library(pkg, character.only = TRUE)
    }
  })
}

packages <- c("purrr", "dplyr", "tidyr", "knitr", "magrittr", "reshape2", 
              "tidyverse", "ggplot2", "RColorBrewer", "stringr","scales", 
              "rstatix", "ggpubr", "ggsignif", "FSA",  
              "readxl", "viridis", "patchwork",   
              "ggfortify", "factoextra", "pheatmap",
              "cowplot","coin","psych","cowplot",
              "phyloseq","microbiome","plyr","hrbrthemes",
              "RColorBrewer","readr", "decontam", "microbiomeutilities",
              "DECIPHER", "ape","phangorn", "tibble","utils","MOFA2","microViz",
              "vegan","mixOmics","data.table")

load_packages(packages)
# library(microViz)
# library(vegan)
# library(mixOmics)
```

#  Load datasets
```{r load ps, message=TRUE}

# Load phyloseq object (microbiome):
ps <- readRDS("07-ps.for.multiomics.rds")

ps.df <- data.frame(sample_data(ps)) #make dataframe 

table(ps.df$Sample_Type) #count number of sample types
table(ps.df$Sampling_Method) #count number of sampling methods

ps #print phyloseq object (7171 taxa in 308 samples)
#-----------------------------------------------------

# Import only LegendPlex data:
lgnd <- read_excel("02-LegendPlex_data.xlsx")
library(dplyr)
lgnd %>%
  distinct(SampleID, Sample_Type, Analytes) %>%        # one row per sample
  dplyr::count(Analytes, Sample_Type, name = "n_samples")
#-----------------------------------------------------

# Import total Abs data:
total_abs <- read_excel("01-Luminex_Total_Abs_Data.xlsx")
total_abs %>%
  distinct(Description, Sample_Type, Analyte) %>%        # one row per sample
  dplyr::count(Analyte, Sample_Type, name = "n_samples")


#-----------------------------------------------------

# Import RespiFinder (multiplex PCR) data:

respifinder <- read_excel("04-antigen_specific_MIA_data.xlsx")
respifinder <- respifinder %>%
  mutate(
    Pathogen_group = case_when(
      Result %in% c("SARS-CoV2", "Rhino/Entero", "Boca", "Influenza A") ~ "Virus",
      Result %in% c("C.Pneu", "B.Pert") ~ "Bacterium",
      is.na(Result) ~ "Negative"
    )
  )

resp_flags <- respifinder %>%
  distinct(SampleID, Pathogen_group) %>%   # one row per sample × group
  mutate(
    Virus_pos = Pathogen_group == "Virus",
    Bact_pos  = Pathogen_group == "Bacterium",
    Any_pos   = Pathogen_group != "Negative"
  ) 
```



# Data cleanup 
We will only use Professionally collected samples for the integrated analysis. 

```{r data-cleanup}

ps.prof <- phyloseq::subset_samples(ps, Sampling_Strategy == "Professional")
ps.prof.df <- data.frame(sample_data(ps.prof))

lgnd.prof <- lgnd %>% filter( Sampling_Strategy == "Professional")
lgnd.prof <- lgnd.prof %>% mutate(log10_conc = log10(concentration.est))
# lgnd.prof <- lgnd.prof %>% mutate(log_conc = log(concentration.est))
lgnd.prof <- lgnd.prof %>%
  mutate(Sample_Type = recode(Sample_Type, "Mid-turbinate swab" = "MTS"))
lgnd.prof <- lgnd.prof %>%
  mutate(Sample_Type = recode(Sample_Type, "Nasopharyngeal swab" = "NPS"))
lgnd.prof <- lgnd.prof %>%
  mutate(Sample_Type = recode(Sample_Type, "Oropharyngeal swab" = "OPS"))

total_abs.prof <- total_abs %>% filter( Sampling_Strategy == "Professional")

# Function to replace "OOR <" values with half of the minimum or half of LLOQ if all values are NA
replace_missing_with_min <- function(x, lloq) {
  min_val <- min(x, na.rm = TRUE)
  if (is.finite(min_val)) {
    return((1/2) * min_val)
  } else {
    return((1/2) * lloq)  # Use half of LLOQ if all values are NA
  }
}

# Function to replace "OOR >" values with the maximum x2 concentration in range
replace_missing_with_max <- function(x) {
  max_val <- max(x, na.rm = TRUE)
  if (is.finite(max_val)) {
    return(max_val*2)
  } else {
    return(NA)  # If all values are NA, return NA or another desired fallback
  }
}

# Apply the functions per isotype and sample type
total_abs.prof <- total_abs.prof %>%
  group_by(Analyte, Sample_Type) %>%
  mutate(
    concentration = case_when(
      is.na(Conc_in_Range) & Notes == "OOR <" ~ replace_missing_with_min(Conc_in_Range, LLOQ),
      is.na(Conc_in_Range) & Notes == "OOR >" ~ replace_missing_with_max(Conc_in_Range),
      TRUE ~ Conc_in_Range
    )
  ) %>%
  ungroup()

#  Calculate log10 of the new concentration column
total_abs.prof <- total_abs.prof %>%
  mutate(log10_conc = log10(concentration))
```


# Build metadata
```{r}
# 1. Build metadata starting from LegendPlex professional
resp_flags_1 <- resp_flags %>% subset(select=c(SampleID, Virus_pos, Bact_pos, Any_pos))
meta <- lgnd.prof %>%
  distinct(SampleID, Sample_Type, ID, Biological_Sex, Age_Range, Nostril) %>%
  left_join(resp_flags_1, by = "SampleID")

# check
dim(meta)
table(meta$Sample_Type)

  # MLF    MTS    NPS    OPS Saliva 
  #   88     88     44     44     40 

```

```{r}
total_abs.prof <- total_abs.prof %>%
  dplyr::rename(SampleID = Description)

```

# IDs check
```{r}

ids_microbe <- sample_names(ps.prof)   # from phyloseq
ids_cytokine <- unique(lgnd.prof$SampleID)
ids_igs <- unique(total_abs.prof$SampleID)

# overlap with meta
table(meta$SampleID %in% ids_microbe)
table(meta$SampleID %in% ids_cytokine)
table(meta$SampleID %in% ids_igs)

```

```{r}
setdiff(meta$SampleID, ids_microbe)   # in meta but not microbiome
setdiff(meta$SampleID, ids_cytokine)  # in meta but not cytokines
setdiff(meta$SampleID, ids_igs)       # in meta but not Igs

```

# Find the shared IDs
```{r}
integration_ids <- Reduce(intersect, list(
  meta$SampleID,
  ids_microbe,
  ids_cytokine,
  ids_igs
))

length(integration_ids)  # should be 264
```


# Subset to these IDs (meta)
```{r}
meta_sub <- meta %>% 
  filter(SampleID %in% integration_ids)

nrow(meta_sub)   #  264

table(meta_sub$Sample_Type)

# MLF    MTS    NPS    OPS Saliva 
#     69     80     36     44     35 
meta_sub <- meta_sub %>%
  mutate(SampleID_site = paste(ID, Sample_Type, sep="_"))


```
## microbiome
```{r}
ps.sub <- phyloseq::subset_samples(ps.prof, sample_identifier %in% integration_ids)

# sanity check
nsamples(ps.sub)   # 264
```

# legendplex
```{r}
lgnd.sub <- lgnd.prof %>%
  filter(SampleID %in% integration_ids) %>%
  subset(select=c(SampleID, Analytes, log10_conc))

# wide format (rows = samples, cols = analytes)
cyto.mat <- tidyr::pivot_wider(lgnd.sub,
  names_from = Analytes,
  values_from = log10_conc
)

# reorder rows to match meta_sub
cyto.mat <- cyto.mat[match(meta_sub$SampleID, cyto.mat$SampleID), ]
```

# total Igs
```{r}
igs.sub <- total_abs.prof %>%
  filter(SampleID %in% integration_ids) %>%
  subset(select=c(SampleID, Analyte, log10_conc))

igs.mat <- tidyr::pivot_wider(igs.sub,
  names_from = Analyte,
  values_from = log10_conc
) 

# reorder rows to match meta_sub
igs.mat <- igs.mat[match(meta_sub$SampleID, igs.mat$SampleID), ]
```
# collapse legendplex
```{r}
# start from your long lgnd.sub (SampleID, Analytes, log10_conc)
cyto.mat.collapsed <- lgnd.sub %>%
  dplyr::left_join(meta_sub %>% dplyr::select(SampleID, SampleID_site), by = "SampleID") %>%
  dplyr::group_by(SampleID_site, Analytes) %>%
  dplyr::summarise(log10_conc = mean(log10_conc, na.rm = TRUE)) %>%
  dplyr::ungroup() %>%
  tidyr::pivot_wider(names_from = Analytes, values_from = log10_conc)

# matrix + scaling
X_cytokine <- as.data.frame(cyto.mat.collapsed)
rownames(X_cytokine) <- X_cytokine$SampleID_site
X_cytokine <- as.matrix(X_cytokine[ , -1, drop = FALSE])
X_cytokine <- scale(X_cytokine)
```

# collapse total antibodies
```{r}
igs.mat.collapsed <- igs.sub %>%
  dplyr::left_join(meta_sub %>% dplyr::select(SampleID, SampleID_site), by = "SampleID") %>%
  dplyr::group_by(SampleID_site, Analyte) %>%
  dplyr::summarise(log10_conc = mean(log10_conc, na.rm = TRUE)) %>%
  dplyr::ungroup() %>%
  tidyr::pivot_wider(names_from = Analyte, values_from = log10_conc)



X_igs <- as.data.frame(igs.mat.collapsed)
rownames(X_igs) <- X_igs$SampleID_site
X_igs <- as.matrix(X_igs[ , -1, drop = FALSE])
X_igs <- scale(X_igs)
```
# collapse microbiome & CLRS
```{r}
library(compositions)
library(dplyr)

# 1. Aggregate microbiome to Genus level
ps.prof.genus <- tax_glom(ps.sub, "Genus", NArm = TRUE)

# 2. Relative abundance (%)
ps.prof.rel <- transform_sample_counts(ps.prof.genus, function(OTU) OTU/sum(OTU) * 100)

# 3. Convert to matrix
microbe_data <- as.data.frame(otu_table(ps.prof.rel))
if(taxa_are_rows(ps.prof.rel)) {
  microbe_data <- t(microbe_data)
}

# 4. Replace column names with Genus names
genus_names <- as.character(tax_table(ps.prof.rel)[, "Genus"])
colnames(microbe_data) <- genus_names

# 5. Filter genera by prevalence (>10%)
prevalence_filter <- function(x, threshold = 0.1) mean(x > 0) > threshold
filtered_microbe <- microbe_data[, apply(microbe_data, 2, prevalence_filter)]

# 6. CLR transform
microbe_data_clr <- as.data.frame(clr(filtered_microbe))

# 7. Add SampleID
microbe_data_clr$SampleID <- rownames(microbe_data_clr)

# 8. Collapse left/right nostrils → subject × site
microbe_data_clr <- microbe_data_clr %>%
  left_join(meta_sub %>% subset(select=c(SampleID, SampleID_site)),
            by = "SampleID") %>%
  group_by(SampleID_site) %>%
  summarise_at(vars(-SampleID), mean, na.rm=TRUE) %>%
  ungroup()

# 9. Final matrix 
X_microbe <- scale(as.matrix(microbe_data_clr[,-1]))
rownames(X_microbe) <- microbe_data_clr$SampleID_site

```

```{r}
common_ids <- Reduce(intersect, list(rownames(X_microbe), rownames(X_cytokine), rownames(X_igs)))

# Build collapsed metadata
meta_collapsed <- meta_sub %>%
  dplyr::distinct(ID, Sample_Type, Biological_Sex, Age_Range, 
                  Virus_pos, Bact_pos, Any_pos, SampleID_site) %>%
  dplyr::filter(SampleID_site %in% common_ids) %>%
  dplyr::arrange(match(SampleID_site, common_ids))

meta_collapsed <- meta_collapsed[match(rownames(X_microbe), meta_collapsed$SampleID_site), ]

```


# check alignment
```{r}

X_microbe  <- X_microbe [common_ids, , drop = FALSE]
X_cytokine <- X_cytokine[common_ids, , drop = FALSE]
X_igs      <- X_igs     [common_ids, , drop = FALSE]

meta_collapsed <- meta_sub %>%
  dplyr::distinct(ID, Sample_Type, Biological_Sex, Age_Range, Virus_pos, Bact_pos, Any_pos, SampleID_site) %>%
  dplyr::filter(SampleID_site %in% common_ids) %>%
  dplyr::arrange(match(SampleID_site, common_ids))

stopifnot(identical(rownames(X_microbe), rownames(X_cytokine)))
stopifnot(identical(rownames(X_microbe), rownames(X_igs)))
meta_collapsed <- meta_collapsed[match(rownames(X_microbe), meta_collapsed$SampleID_site), ]

# now check again
stopifnot(identical(meta_collapsed$SampleID_site, rownames(X_microbe)))
# setdiff(rownames(X_microbe), meta_collapsed$SampleID_site)
# setdiff(meta_collapsed$SampleID_site, rownames(X_microbe))
```

```{r}
table(meta_collapsed$Sample_Type)
table_input <- table(meta_collapsed$Sample_Type)
table_input <- as.data.frame(table_input)
#writexl::write_xlsx(table_input,"input_samples_mofa.xlsx")

clean_names <- function(x) {
  x <- gsub("α", "alpha", x)
  x <- gsub("γ", "gamma", x)
  x <- gsub("λ", "lambda", x)
  x <- gsub("[^A-Za-z0-9_]", "_", x)  # replace any other weird chars with underscore
  x
}

colnames(X_cytokine) <- clean_names(colnames(X_cytokine))

```


# RDA


## collapse microbiome 
```{r}
library(compositions)
library(compositions)
library(dplyr)

# 1. Aggregate microbiome to Genus level
ps.prof.genus <- tax_glom(ps.sub, "Genus", NArm = TRUE)

# 2. Relative abundance (%)
ps.prof.rel <- transform_sample_counts(ps.prof.genus, function(OTU) OTU/sum(OTU) * 100)

# 3. Convert to matrix
microbe_data <- as.data.frame(otu_table(ps.prof.rel))
if(taxa_are_rows(ps.prof.rel)) {
  microbe_data <- t(microbe_data)
}

# 4. Replace column names with Genus names
genus_names <- as.character(tax_table(ps.prof.rel)[, "Genus"])
colnames(microbe_data) <- genus_names

# 5. Filter genera by prevalence (>10%)
prevalence_filter <- function(x, threshold = 0.1) mean(x > 0) > threshold
filtered_microbe <- microbe_data[, apply(microbe_data, 2, prevalence_filter)]

filtered_microbe <- as.data.frame(filtered_microbe)
# 6. Hellinger transform (recommended for RDA)
# microbe_data_hell <- decostand(filtered_microbe, method = "hellinger")
# microbe_data_hell <- as.data.frame(microbe_data_hell)

# 7. Add SampleID
filtered_microbe$SampleID <- rownames(filtered_microbe)

# 8. Collapse left/right nostrils → subject × site
filtered_microbe <- filtered_microbe %>%
  left_join(meta_sub %>% subset(select=c(SampleID, SampleID_site)),
            by = "SampleID") %>%
  group_by(SampleID_site) %>%
  summarise_at(vars(-SampleID), mean, na.rm=TRUE) %>%
  ungroup()

# 9. Final matrix 
X_microbe <- as.matrix(filtered_microbe[,-1])
rownames(X_microbe) <- filtered_microbe$SampleID_site

```

```{r}
# matrix + scaling
X_cytokine <- as.data.frame(cyto.mat.collapsed)
rownames(X_cytokine) <- X_cytokine$SampleID_site
X_cytokine <- as.matrix(X_cytokine[ , -1, drop = FALSE])
X_cytokine <- scale(X_cytokine)

X_igs <- as.data.frame(igs.mat.collapsed)
rownames(X_igs) <- X_igs$SampleID_site
X_igs <- as.matrix(X_igs[ , -1, drop = FALSE])
X_igs <- scale(X_igs)
```

## Hellinger transformation
```{r}
X_igs_df<- as.data.frame(X_igs) %>%
  tibble::rownames_to_column("SampleID_site")
X_igs_df <- X_igs_df[,-5]

igs <- colnames(X_igs)
cytos <- colnames(X_cytokine)
bacs <- colnames(X_microbe)

# transformation
X_microbe <- decostand(X_microbe, method = "hellinger")

X_microbe_df<- as.data.frame(X_microbe) %>%
  tibble::rownames_to_column("SampleID_site")

X_cytokine_df <- as.data.frame(X_cytokine)%>%  tibble::rownames_to_column("SampleID_site")
# Combine assays
dat_rda <- X_cytokine_df %>%
  left_join(X_microbe_df, by="SampleID_site") %>%
  left_join(X_igs_df, by="SampleID_site") %>%
  left_join(meta_collapsed, by="SampleID_site") 
dim(dat_rda)
dat_rda <- dat_rda %>% filter(!is.na(IgM))



```



```{r}
custom_colors <- c("NPS" = "#A2D2DF", 
                   "MTS" = "bisque3",       
                   "MLF" = "#FFBD73",
                   "OPS" = "#BC7C7C",
                   "Saliva" = "#8ABFA3")
```


# RDA (FINAL)
```{r rda-v2, fig.width=11, fig.height=8}
library(vegan)
library(ggplot2)
library(ggrepel)
library(permute)
library(dplyr)

# microbes: Hellinger (on relative abundances)
X_microbe <- dat_rda[, bacs]
# immune + Igs: z-score (already log scaled)
X_immune  <- dat_rda[, cytos, drop=FALSE]
X_igs     <- dat_rda[, igs,   drop=FALSE]

# (Optional) block-weight so each domain contributes similarly
bw <- function(M) M / sqrt(sum(apply(M, 2, var)))
X_features <- cbind(bw(X_microbe), bw(X_immune), bw(X_igs))

# # ---- PERMANOVA: do niches differ in joint space? ----
# dist_features <- vegdist(X_features, method = "euclidean")
set.seed(123)
# perm <- how(blocks = dat_rda$ID)
# permanova <- adonis2(dist_features ~ Sample_Type + Biological_Sex + Age_Range + Any_pos,
#                      data = dat_rda, permutations = perm, by = "margin")
# 
# permanova_df <- as.data.frame(permanova)



#writexl::write_xlsx(permanova_df, "RDA_PERMANOVA_result.xlsx")
# ---- RDA: all domains ~ Sample_Type | covariates ----
dat_rda <- dat_rda[!is.na(dat_rda$Any_pos), ]
mod_rda <- rda(X_features ~ Sample_Type + Condition(ID + Biological_Sex + Age_Range + Any_pos),
               data = dat_rda)


# Global model test
anova(mod_rda, permutations = 999)

# Each predictor (Sample_Type vs covariates you conditioned on)
mod_rda_all <- rda(X_features ~ Sample_Type + ID + Biological_Sex + Age_Range + Any_pos,
                   data = dat_rda)
anova(mod_rda_all, by = "term", permutations = 999)

# Each canonical axis
anova(mod_rda, by = "axis", permutations = 999)

library(writexl)

# ---- Permutation tests ----
res_global <- anova(mod_rda, permutations = 999)
res_terms  <- anova(mod_rda_all, by = "term", permutations = 999)
res_axes   <- anova(mod_rda, by = "axis", permutations = 999)

df_global <- as.data.frame(res_global)
df_terms  <- as.data.frame(res_terms)
df_axes   <- as.data.frame(res_axes)

# ---- Variance explained ----
Rsq_adj <- RsquareAdj(mod_rda)
imp <- summary(mod_rda)$concont$importance

df_variance <- data.frame(
  Metric = c("Total constrained (%)", "R2", "Adj R2", "RDA1 (%)", "RDA2 (%)"),
  Value  = c(
    round(100 * mod_rda$CCA$tot.chi / mod_rda$tot.chi, 2),
    round(Rsq_adj$r.squared, 3),
    round(Rsq_adj$adj.r.squared, 3),
    round(100 * imp[2,1], 1),
    round(100 * imp[2,2], 1)
  )
)

# ---- Write all results into one Excel workbook ----
# write_xlsx(
#   list(
#     Global   = df_global,
#     Terms    = df_terms,
#     Axes     = df_axes,
#     Variance = df_variance
#   ),
#   "RDA_results.xlsx"
# )

# Stats
Rsq_adj <- Rsq_adj$adj.r.squared
imp <- summary(mod_rda)$concont$importance
axis1 <- round(100*imp[2,1], 1); axis2 <- round(100*imp[2,2], 1)
total_constrained <- round(100*mod_rda$CCA$tot.chi / mod_rda$tot.chi, 1)


# vif.cca(mod_rda)
# ---- Scores ----
sites <- as.data.frame(scores(mod_rda, display = "sites", scaling = 2)) %>%
  mutate(Sample_Type = dat_rda$Sample_Type)

# Species scores = loadings for EVERY feature in Y (microbes + immune + Igs)
sp <- as.data.frame(scores(mod_rda, display = "species", scaling = 2)) %>%
  mutate(var = rownames(.),
         domain = case_when(
           var %in% colnames(X_microbe) ~ "Microbe",
           var %in% colnames(X_immune)  ~ "Cytokine",
           var %in% colnames(X_igs)     ~ "Ig",
           TRUE ~ "Other"
         ))

# pick top drivers per domain by vector length
sp$len <- sqrt(sp$RDA1^2 + sp$RDA2^2)
top_microbes <- sp %>% filter(domain=="Microbe")  %>% arrange(dplyr::desc(len)) %>% slice_head(n=10)
top_cyt      <- sp %>% filter(domain=="Cytokine")%>% arrange(dplyr::desc(len)) %>% slice_head(n=6)
top_igs      <- sp %>% filter(domain=="Ig")       %>% arrange(dplyr::desc(len)) %>% slice_head(n=3)
# remove IgA, IgG, IgM from the plotted loadings
top_igs_plot <- top_igs 
top_microbes_plot <- top_microbes 
top_all <- bind_rows(top_microbes_plot, top_cyt, top_igs_plot)

# ---- Hulls ----
hulls <- do.call(rbind, lapply(split(sites, sites$Sample_Type), function(df) df[chull(df$RDA1, df$RDA2), ]))



# rescale loadings to stroke thickness
top_all$stroke <- scales::rescale(top_all$len, to = c(0.5, 2.5))
library(ggnewscale)
```

```{r, fig.width=11, fig.height=8}
rda_plot <- ggplot() +
  # sample points (colored by Sample Type)
  geom_point(data = sites, aes(RDA1, RDA2, color = Sample_Type), size = 3, alpha = 0.5) +
  
  # convex hulls = single constant fill
  geom_polygon(data = hulls, aes(RDA1, RDA2, group = Sample_Type),
               fill = "grey85", alpha = 0.2, color = "grey75",   # subtle outline
  linewidth = 0.2, linetype="dashed" ) +
  scale_color_manual(name = "Sample Type", values = custom_colors,
                     guide = guide_legend(order = 1)) +

  # reset scales before domain layers
  ggnewscale::new_scale_color() +
  ggnewscale::new_scale_fill() +

  # drivers (Domain × len × fill)
  geom_point(data = top_all,
             aes(RDA1, RDA2, shape = domain, size = len, fill = domain),
             color = "black") +

  # labels
  geom_text_repel(data = subset(top_all, domain == "Microbe"),
                  aes(RDA1, RDA2, label = var),
                  size = 5, fontface = "italic", max.overlaps = 10,
                  box.padding = 0.6, point.padding = 0.5,
                  segment.color = "grey60", show.legend = FALSE) +
  geom_text_repel(data = subset(top_all, domain != "Microbe"),
                  aes(RDA1, RDA2, label = var),
                  size = 5, box.padding = 0.65, point.padding = 0.6,
                  segment.color = "grey60", show.legend = FALSE) +

  # Domain (merged legend: shape + fill both named "Domain", same order)
  scale_shape_manual(
    name  = "Domain",
    values = c(Microbe = 24, Cytokine = 22, Ig = 23),   # filled triangle, diamond, square
    guide  = guide_legend(order = 2, override.aes = list(size = 6))
  ) +
  scale_fill_manual(
    name  = "Domain",
    values = c(Microbe = "grey20", Ig = "grey70", Cytokine = "white"),
    guide  = guide_legend(order = 2)  # no size override here; shape guide handles size
  ) +

  # Contribution (len) as discrete-looking sizes; legend shows squares
  scale_size_continuous(
    name   = "Contribution",
    range  = c(3, 9),
    breaks = c(0.4, 0.7, 1.0, 1.3),
    labels = c("0.4", "0.7", "1.0", "1.3"),
    guide  = guide_legend(order = 3,
               override.aes = list(shape = 22, fill = "grey70", color = "black"))
  ) +

  labs(
    x = paste0("RDA1 (", axis1, "%)"),
    y = paste0("RDA2 (", axis2, "%)"),
    subtitle = paste0("Adj. R² = ", round(Rsq_adj, 2),
                      ", ANOVA F = 56.1, p < 0.01 ")
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title = element_text(size = 16),
    axis.text  = element_text(size = 14),
    plot.subtitle = element_text(size = 14),
    legend.title = element_text(size = 14),
    legend.text  = element_text(size = 13),
    legend.key.size = unit(1.1, "cm")
  )

rda_plot
```
